---
title: "Shrinking body size in Atlantic Forest birds as a response to climate change"
author: 
- Eduardo S.A. Santos
- João C.T. Menezes
- Gustavo Burin
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
bibliography: atlantic_bird_bibliography.bib  
---

# Introductory paragraph
Guidelines^[The paper should then start with an introductory paragraph (not abstract) of no more than 200 words, summarizing the background, rationale, main results (introduced by ‘Here we show’ or some equivalent phrase) and implications of the study. This paragraph should be fully referenced. Check other guidelines to **Nature Climate Change** here: https://www.nature.com/nclimate/about/content]

Two of the main responses of organisms to climate change are phenological changes and geographical range shifts [@parmesan_globally_2003]. A third response of organisms to global warming is increasingly being observed, the shrinkage of bodies [@teplitsky_climate_2014; @sheridan_shrinking_2011; @gardner_declining_2011]. Body shrinkage due to climate change has been hypothesized to be a consequence of ecological or evolutionary processes [@ozgul_dynamics_2009]. The ecological hypothesis posits that climate change influences trophic interactions, thus leading to a reduction in body size. Changes in temperature can result in temperature-dependent or -independent mechanisms that influece body size [reviewed in @gardner_declining_2011]. Among temperature-independent mechanisms, decreases in food availability caused by habitat loss or fragmentation may lead to the reduction of body size [@schmidt_concomitant_2005]. Thus, limiting juveniles access to resources could lead to smaller adult body size [@teplitsky_bergmanns_2008; @rode_reduced_2010]. 

Reductions in body size have been attributed to genetic microevolutionary responses to warming; smaller individuals dissipate heat better than larger individuals due to the larger ratio between surface:volume of their bodies [e.g., Bergmann's rule, @teplitsky_climate_2014].^[I think that this paragraph is not necessary. Talk only about the ecological consequences.]

Based on a large scale analysis of Neotropical passerines that inhabit the Brazilian Atlantic Forest, we show here that annual mean temperature has been increasing, while precipitation^[We need to choose what precipitation variable to use. It would be good to choose one that relates to the availability of food (insects or fruits).] has been decreasing over a period of **XX** years in the areas inhabited by these passerines.

> Paragraph to talk about the main finding of this study in relation to body size and climate change. For *Nature Climate Change* remember to combine results and some methods so that readers understand what you did.

# Methods
## Load libraries
Here are all the libraries that we used for data wrangling, plotting and model building
```{r libraries, echo=TRUE,cache=TRUE, message=FALSE}
library("tidyverse")
library("ggthemes") 
library("brms")
library("lme4")
library("ape")
library("geiger")
library("MCMCglmm")
library("mice")
library("tidybayes")
```

## Load data
We used the Atlantic Bird Traits [@rodrigues_atlantic_2019] dataset as a source of information upon which to address our hypothesis.
```{r data,echo=TRUE, cache=TRUE, message=FALSE, eval=FALSE}
birds <- read_csv("ATLANTIC_BIRD_TRAITS_completed_2018_11_d05.csv", guess_max = 70000)
```

## Data wrangling and new variables
Subsetting the original dataframe into sets to explore.

We restricted our data collection to the order Passeriformes, as passerine birds represent over 82% of the entries in the Atlantic Bird Traits dataset (60,066 entries of a total of 72,483). We restricted our data to individuals that were recorded as adults during the original data collection (n = 43,149). We used records of individuals that were collected/recorded between 1990 and 2018, which yielded a total of 40,183 passerine entries in our dataset. We will only keep records of species that have at least 50 entries in the dataset. This will reduce the dataset from 435 passerine species to 143 species. The last wrangling step is to remove all individuals that were not sexed in the original dataset, leaving a total of 68 passerine birds for analyses. Using only records that were registed within the 20 km buffer polygon for the Atlantic Forest domain 
```{r newdataframes,cache=TRUE, eval=FALSE}
passer90 <- birds %>%
          filter(Year >= 1990, Age == "Adult", Sex != "Unknown", Order == "Passeriformes", AtlanticForests_20km_Buffer == "inside the 20 km polygon") %>%
          mutate(
            conc.wing.length = coalesce(Wing_length_right.mm., Wing_length_left.mm., Wing_length.mm.), # combining all wing lengths to minimize number of NAs in dataset
            ln_conc_wing_length = log(conc.wing.length),
            ln_wing_length = log(Wing_length_right.mm.),
            scaled_lat = scale(Latitude_decimal_degrees*-1),
            scaled_yr = scale(Year),
            Binomial = gsub(" ", "_", Binomial),
            Binomial = gsub("Ceratopipra_rubrocapilla", "Pipra_rubrocapilla", Binomial),
            Binomial = gsub("Myiothlypis_leucoblephara", "Basileuterus_leucoblepharus", Binomial),
            Binomial = gsub("Myrmoderus_squamosus", "Myrmeciza_squamosa", Binomial),
            Binomial = gsub("Pyriglena_pernambucensis", "Pyriglena_leuconota", Binomial),
            Binomial = gsub("Rhopias_gularis", "Myrmotherula_gularis", Binomial),
            Binomial = gsub("Tangara_sayaca", "Thraupis_sayaca", Binomial),
            spp = Binomial
          ) %>%
          group_by(Binomial) %>%
          filter(n() >= 50) %>%
            mutate(
                  min = min(Year),
                  max = max(Year),
                  range = max-min
                        ) %>%
            filter(range >= 10)

# keeping only variables that I will use in the models:
passer90 <- passer90 %>%
              dplyr::select(spp, Binomial, Sex, Year, scaled_yr, Latitude_decimal_degrees, scaled_lat, Longitude_decimal_degrees, Wing_length_right.mm.,ln_wing_length,conc.wing.length,ln_conc_wing_length)

passer90 <- as.data.frame(passer90)

passer90$scaled_lat <- as.numeric(passer90$scaled_lat)
passer90$scaled_yr <- as.numeric(passer90$scaled_yr)

save(passer90, file = "passer90.rda")

rm(birds)
```


Checking the range of years per species in the *passer90* dataframe. Creating a new data frame that contains one entry per species in order to prune the phylogeny later.
```{r year_range, cache=TRUE}
load("passer90.rda")

by_binomial <- group_by(passer90, Binomial)
year_range <- summarise(by_binomial,
                  min = min(Year),
                  max = max(Year),
                  range = max-min
                        )

year_range <- as.data.frame(year_range)
row.names(year_range) <- year_range$Binomial # having rows named with the species name is required for any PCM
```

Plots of the range of years per species:
```{r plots_year_range, cache=TRUE}
g_year_range <- ggplot(data = year_range, aes(x = range)) + 
                  geom_bar() +
                  theme_fivethirtyeight()

g_year_range

summary(year_range$range)

g_year_min <- ggplot(data = year_range, aes(x = min)) + 
                  geom_bar() +
                  theme_fivethirtyeight()
g_year_min

g_year_max <- ggplot(data = year_range, aes(x = max)) + 
                  geom_bar() +
                  theme_fivethirtyeight()
g_year_max
```


For each species in the dataset, the range of years of data available ranges from 2 to 28. It seems reasonable to use only those species for which we have at least 10 years of data. I haven't looked into what constitutes a good longitudinal dataset, but less than 10 years just seems too little in order to establish a link between temperature and body size.

After filtering out the species with less than 10 years of data, we ended up with a total of 68 passerine species.

## Exploratory plots
Plotting some relationships:
```{r plots, cache=TRUE, eval = FALSE}
g_wingl <- ggplot(data = passer90, aes(x = scaled_yr, y = conc.wing.length, group = Binomial)) +
            geom_point(aes(alpha = 0.09)) +
            stat_smooth(method = "lm", se = FALSE, colour = "grey50", size = 0.5, alpha = 0.1) + 
            theme_fivethirtyeight()
g_wingl

g2 <- ggplot(data = passer90, aes(x = Year, y = Annual_mean_temperature)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g2

g3 <- ggplot(data = passer90, aes(x = Year, y = Annual_rainfall)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g3
```

## Plotting map of species locations:
```{r maps_birds, cache=TRUE}
library("ggmap")

bbox <- make_bbox(passer90$Longitude_decimal_degrees, passer90$Latitude_decimal_degrees, f = 0.01)
map <- get_map(bbox, source = 'osm', color = 'bw')

passer90$period <- ifelse(passer90$Year <= 2004, '1990-2004', '2005-2018')

points <- ggmap(map) + 
            geom_point(aes(x = Longitude_decimal_degrees, y = Latitude_decimal_degrees, colour = conc.wing.length), data = remove_missing(passer90, vars = "conc.wing.length"), alpha = .5) +
            scale_color_viridis_c(option = "plasma")+
            labs(title = "Localities of passerine records",
                 x = "Longitude (decimal degrees)",
                 y = "Latitude (decimal degrees)") +
            facet_wrap(~period)

points
```


## Exploratory models
A few quick models (using **lme4**) to check the trends. It seems like we have overall evidence that there is a negative relationship between *Year* and *Wing length* for all species.
```{r models, cache=TRUE, eval = FALSE}
m0 <- lmer(Wing_length_right.mm. ~ scaled_yr + Sex + scaled_lat + (1 + scaled_yr | Binomial),
           data = passer90,
           REML = FALSE,
           control = lmerControl(optimizer ="Nelder_Mead")
           )

summary(m0)

m0.1 <- lmer(ln_conc_wing_length ~ scaled_yr + Sex + scaled_lat + (1 + scaled_yr | Binomial),
           data = passer90,
           REML = FALSE,
           control = lmerControl(optimizer ="Nelder_Mead")
           )

summary(m0.1)


m1 <- lmer(Body_mass.g. ~ scaled_yr + Sex + scaled_lat + (1 + scaled_yr | Binomial),
           data = passer90,
           REML = FALSE,
           control = lmerControl(optimizer ="Nelder_Mead")
           )

summary(m1)

```

### Loading the phylogeny
Loading phylogeny to include in the multi-level model.
For all analyses, we used species-level Passeriformes trees taken from a sample of 1,000 topologies based on the Hackett backbone from Jetz et al.'s [-@jetz2012global] [BirdTree.org](http://birdtree.org). We will later prune these 5,966-species trees to fit the species in each analyses for which we have succesfully collected complete data.

The chunk to load the phylogeny will only be evaluated in the local version of the code, to save time and space when in version control.
```{r phylo, cache=TRUE, eval= FALSE}
trees <- read.nexus("Hackett_trees.nex")
```

Chunk to process and save a sample of 50 trees:
```{r tree_sample, cache=TRUE, eval=FALSE}
namecheck_passer <- name.check(trees[[1]], year_range) # checking which species are in the phylogeny but not in our data frame

# namecheck_subset <- namecheck_passer$tree_not_data[grep("Thraupis", namecheck_passer$tree_not_data)] # Code to check if the species is available in the phylogeny under a different (synonim) genus name.

tree_samp <- sample(trees, size = 50) # Sampling 50 trees from the 1000 file. This is enough for the phylogenetic uncertainty.

trees_passer <- lapply(tree_samp, drop.tip, namecheck_passer$tree_not_data) # pruning these species from all the 50 trees

class(trees_passer) <- "multiPhylo"

nrow(year_range) # number of species for which we have complete data
length(trees_passer[[1]]$tip.label) # number of tips in the phylogenic trees

save(trees_passer, file = "trees_passer.rda")


rm(list = c("trees", "year_range", "by_binomial", "tree_samp", "namecheck_passer")) # remove objects that aren't needed anymore
```

Load multiphylo file with 50 sampled trees:
```{r load_multiphylo, cache=TRUE}
# The phylo object contains information on the relationship between species. Using this information, we can construct a covariance matrix of species.
load("trees_passer.rda")
inv.phylo <- inverseA(trees_passer[[1]], nodes = "TIPS", scale = TRUE)
A <- solve(inv.phylo$Ainv)
rownames(A) <- rownames(inv.phylo$Ainv)

```


### Multiple imputation
Before running the multi-level Bayesian model, I will use multiple imputation to impute missing data with the package *mice*. The idea is to create 5 new datasets with fully imputed data.
```{r mice_data, cache = TRUE, eval=FALSE}
require(lattice)
require(pan)
require(multilevel)

# Exploring the missing data pattern:
md.pattern(passer90, rotate.names = TRUE)

histogram(~ as.numeric(as.factor(Binomial)) | is.na(Wing_length_right.mm.), data=passer90)
# We do see that the histogram for the missing wing length (TRUE) is further to the right than the histogram for observed wing length (FALSE). This would indicate a right-tailed MAR missingness.

histogram(~ as.numeric(as.factor(Binomial)) | is.na(Latitude_decimal_degrees), data=passer90)
# We do see that the histogram for the missing Latitude (TRUE) is centered on the histogram for observed Latitude (FALSE). This would indicate a non-skewed MAR missingness.

# Have a look at the intraclass correlation (ICC) for the incomplete variables: Wing length & Latitude:
ICC1(aov(Wing_length_right.mm. ~ Binomial, data = passer90))

ICC1(aov(Latitude_decimal_degrees ~ Binomial, data = passer90))

# Impute the passer90 dataset with mice using imputation method norm
ini <- mice(passer90, maxit = 0)
meth <- ini$meth
meth

meth[c(6, 7, 8, 9)] <- "norm"
meth

pred <- ini$pred
pred

start_time <- Sys.time()
#passer90_imp <- mice(passer90, meth = meth, pred = pred, print = FALSE)
end_time <- Sys.time();end_time - start_time

# Compare the means of the variables in the first imputed dataset and in the incomplete dataset:
summary(complete(passer90_imp))

summary(passer90)

# Compare the ICC’s of the variables in the first imputed dataset with those in the incomplete dataset:
data.frame(vars = names(passer90[c(8, 6)]), 
           observed = c(ICC1(aov(Wing_length_right.mm. ~ Binomial, passer90)), 
                        ICC1(aov(Latitude_decimal_degrees ~ Binomial, passer90)) 
                        ), 
           norm     = c(ICC1(aov(Wing_length_right.mm. ~ Binomial, complete(passer90_imp))), 
                        ICC1(aov(Latitude_decimal_degrees ~ Binomial, complete(passer90_imp))) 
                        ))

# Now impute the passer90 dataset again with mice using imputation method norm for the same variables as before, but now include Binomial as a predictor for all variables:
pred <- ini$pred
pred[, "Binomial"] <- 1
pred

start_time <- Sys.time()
#passer90_imp1 <- mice(passer90, meth = meth, pred = pred, print = FALSE)
end_time <- Sys.time();end_time - start_time

# Compare the ICC’s of the variables in the first imputed dataset with those in the incomplete dataset and the 2nd imputed set:
data.frame(vars = names(passer90[c(8, 6)]), 
           observed = c(ICC1(aov(Wing_length_right.mm. ~ Binomial, passer90)), 
                        ICC1(aov(Latitude_decimal_degrees ~ Binomial, passer90)) 
                        ), 
           norm     = c(ICC1(aov(Wing_length_right.mm. ~ Binomial, complete(passer90_imp))), 
                        ICC1(aov(Latitude_decimal_degrees ~ Binomial, complete(passer90_imp))) 
                        ), 
           normBinomial     = c(ICC1(aov(Wing_length_right.mm. ~ Binomial, complete(passer90_imp1))), 
                        ICC1(aov(Latitude_decimal_degrees ~ Binomial, complete(passer90_imp1))) 
                        )
           )

plot(passer90_imp1, "Wing_length_right.mm.")

# Add 10 more iterations:
#passer90_imp2 <- mice.mids(passer90_imp1, maxit = 10)

plot(passer90_imp2, "Wing_length_right.mm.")

# Add 20 more iterations:
#passer90_imp3 <- mice.mids(passer90_imp2, maxit = 20, print = FALSE)

plot(passer90_imp3, "Wing_length_right.mm.")

# Plot the densities of the observed and imputed data (use passer90_imp3) with the function densityplot().
densityplot(passer90_imp3)

# Impute the passer90 data once more where you use predictive mean matching and include all variables as predictors. Name the object passer90_imp4.
passer90_imp4 <- mice(passer90, print = FALSE)

densityplot(passer90_imp4, ~Wing_length_right.mm.)
plot(passer90_imp4)

passer90_imp4a <- mice.mids(passer90_imp4, maxit = 20, print = FALSE)
plot(passer90_imp4a)
densityplot(passer90_imp4a, ~Wing_length_right.mm.)

save(passer90_imp4a, file = "passer90_imp4a.rda")
```

### Multilevel models with imputed data and phylogeny
Here is a multi-level Bayesian model of the response variable *Wing length* including the phylogeny of the species in question. If I run with the original data set, even though we have over 14k entries, there are only 4.6k complete entries that **brm** will use to estimate parameters. This model is running over the imputed dataset passer90_imp4a:
```{r model_phylo_imp, cache=TRUE, eval = FALSE}
load("passer90_imp4a.rda")

start_time <- Sys.time()
brm <- brm_multiple(Wing_length_right.mm. ~ 1 + Sex + scaled_yr + scaled_lat + (1 + scaled_yr || spp) + (1 | Binomial),
            cov_ranef = list(Binomial = A),
            data = passer90_imp4a,
            family = gaussian,
            prior = c(prior(normal(71, 15), class = Intercept),
                      prior(normal(0, 10), class = b),
                      prior(cauchy(0, 1), class = sd),
                      prior(cauchy(0, 1), class = sigma)),
            chains = 2, 
            cores = 6, 
            control = list(max_treedepth = 12))

end_time <- Sys.time();end_time - start_time

save(brm, file = "brm.rda")
```

I will keep this full model here for a while. I will run the model from the chunk above as a test model with the multiple imputation data.
```{r model_phylo, cache=TRUE, eval = FALSE}
start_time <- Sys.time()
brm0 <- brm(conc.wing.length ~ 1 + Sex + scaled_yr + scaled_lat + (1 + scaled_yr || spp) + (1 | Binomial),
            cov_ranef = list(Binomial = A),
            data = passer90,
            family = gaussian,
            prior = c(prior(normal(71, 15), class = Intercept),
                      prior(normal(0, 10), class = b),
                      prior(cauchy(0, 1), class = sd),
                      prior(cauchy(0, 1), class = sigma)),
            iter = 1500, 
            warmup = 1000, 
            chains = 4, 
            cores = 4, 
            control = list(max_treedepth = 12))

end_time <- Sys.time();end_time - start_time

save(brm0, file = "brm0.rda")
```

Summary and model convergence checks:
```{r model_summary, cache=TRUE}
load("brm0.rda")

summary(brm0)

stanplot(brm0)
#brms::launch_shinystan(brm0)
plot(brm0)
plot(marginal_effects(brm0))
pp_check(brm0)

hyp <- "sd_Binomial__Intercept^2 / (sd_Binomial__Intercept^2 + sd_spp__Intercept^2 + sd_spp__scaled_yr^2 + sigma^2) = 0"
(hyp <- hypothesis(brm0, hyp, class = NULL))
```

### Multi-level model with phylogenetic uncertainty
I asked Ari to run the multilevel model of **Wing length** on the server so that we could run the model over a sample of 50 tree topologies, in order to incorporate phylogenetic uncertainty. The code to run that model is in the file *atlantic_parallel.R*, which is in the manuscript's folder (*/Volumes/Macintosh HD/Users/eduardosantos/Google Drive/Manuscripts/MyPapers/CurrentlyWorking/2019_atlantic_birds/edu_joao/atlantic_parallel.R*). We saved the results of this run in the file **brm0.rda**. Below, I will load this file from the **edu_joao** folder in order to check its summary and diagnostics.

```{r model_summary_multi_phylo, cache=TRUE}
load("~/Google Drive/Manuscripts/MyPapers/CurrentlyWorking/2019_atlantic_birds/edu_joao/brm0.rda")

brm_multi_phylo <- combine_models(mlist = brm0)
summary(brm_multi_phylo)


stanplot(brm_multi_phylo)
#brms::launch_shinystan(brm0)
plot(brm_multi_phylo)
plot(marginal_effects(brm_multi_phylo))
pp_check(brm_multi_phylo)

hyp <- "sd_Binomial__Intercept^2 / (sd_Binomial__Intercept^2 + sd_spp__Intercept^2 + sd_spp__scaled_yr^2 + sigma^2) = 0"
(hyp <- hypothesis(brm_multi_phylo, hyp, class = NULL))

post <- posterior_samples(brm_multi_phylo)

beta_yr_ef <- 
  post %>% 
  transmute(beta_yr_ef = b_scaled_yr)

require("tidybayes")

beta_yr_ef %>%
  ggplot(aes(x = beta_yr_ef)) +
  geom_density(size = 0, fill = "royalblue") +
  stat_pointintervalh(aes(y = 0), 
                      point_interval = mode_hdi, .width = .89) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(x = "Year effect (centred/scaled)") +
  theme_classic()
```



Here is a multi-level Bayesian model of the response variable *Body mass* including the phylogeny of the species in question. I want to try using the *get_prior()* function the next time I run this model^[https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/missing-data-and-other-opportunities.html]. Right now, I won't evaluate this chunk, to save time. I will run the code locally and just past the results from a local **brms** run.
```{r model_phylo_mass, cache=TRUE, eval= FALSE}
start_time <- Sys.time()
brm1 <- brm(Body_mass.g. ~ 1 + Sex + scaled_yr + scaled_lat + (1 + scaled_yr || spp) + (1 | Binomial),
            cov_ranef = list(Binomial = A),
            data = passer90,
            family = gaussian,
            prior = c(prior(normal(71, 15), class = Intercept),
                      prior(normal(0, 10), class = b),
                      prior(cauchy(0, 1), class = sd),
                      prior(cauchy(0, 1), class = sigma)),
            iter = 2000, 
            warmup = 1000, 
            chains = 3, 
            cores = 3, 
            control = list(max_treedepth = 12))

end_time <- Sys.time();end_time - start_time

summary(brm1)
stanplot(brm1)
#brms::launch_shinystan(brm0)

plot(marginal_effects(brm1), points = TRUE)
pp_check(brm1)

hyp <- "sd_Binomial__Intercept^2 / (sd_Binomial__Intercept^2 + sd_spp__Intercept^2 + sigma^2) = 0"
(hyp <- hypothesis(brm1, hyp, class = NULL))

```

# Elton traits to check bird diet
I am going to have a look at Elton Traits [@wilman2014eltontraits] in order to check what is the functional diet category of the birds in our dataset. Elton traits' metadata can be accessed through this  <a href="http://www.esapubs.org/archive/ecol/E095/178/metadata.php" target="_blank">link</a>.

Import Elton traits dataset:
```{r import_elton_traits, cache=TRUE, eval=FALSE}
elton <- read_tsv("BirdFuncDat.txt")
```

Filtering Elton Traits to the species in the wing length dataset:
```{r filtering_elton, cache=TRUE}
elton <- elton %>%
  mutate(Scientific = gsub(" ", "_", Scientific)
         ) %>%
  semi_join(passer90, by = c("Scientific" = "Binomial")
            ) # Keeping only the Elton traits entries that match the species in the passer90 tibble
```

Of the passerines in the **wing length** dataset, here is a breakdown of their dominant diet categories, according to the data from Elton traits:
```{r summary_diets, cache=TRUE}
table(elton[,20])

(table(elton[,20])/68)*100
```
The tables above present the counts and percetange of species, respectively, per diet category.

Only `r length(elton$Scientific[elton[,10] == 0])` species are listed in Elton Traits as having 0% of their diet consisting of invertebrates. The species that do not have any invertebrates in their diet are:
```{r spp_no_inv_diet, cache=TRUE, results='asis'}
cat(paste('-', elton$Scientific[which(elton$`Diet-Inv` == 0)]), sep = '\n') 
```

All other species in the **wing length** dataset have at least some invertebrate prey as part of their diet.

# Arthropod abundance over time analysis
I wanted to look for a dataset that contained information on the abundance of arthropods in Brazil (more specifically, if available, in the Atlantic Forest) over time. The idea being that we could use this information as part of the argument that what we are observing for the birds is in fact being caused by a decline in the abundance of their food source over time. Thus, we have a signature of a phenotypic plasticity process at hand, and not a micro-evolutionary response.

I found the PREDICTS project [@hudson2017database], which has collated a huge amount of biodiversity data per locality, with more than 3.2 million records. I downloaded the main database (**/Volumes/Macintosh HD/Users/eduardosantos/Downloads/database.csv**) and then proceeded to wrangle the data to obtain records for Arachnida and Insecta for whole Brazil:
```{r load_predicts_main, cache=TRUE, eval=FALSE}
dat <- read_csv("/Volumes/Macintosh HD/Users/eduardosantos/Downloads/database.csv", guess_max = 70000)
```

Subsetting PREDICTS dataset to Brazil and Insecta + Arachnida classes only:
```{r subset_predicts, cache=TRUE, eval=FALSE}
# Playing with data from WHOLE Brazil to see what is going on:
dat_test <- dat %>%
        filter(Phylum == "Arthropoda", Diversity_metric_type == "Abundance", UN_subregion == "South America", Class %in% c("Insecta", "Arachnida")) %>%
        mutate(
          scaled_yr = as.numeric(scale(Sample_midpoint)),
          log_abundance = log(Effort_corrected_measurement),
          Measurement = as.integer(Measurement),
          Class = as.factor(Class)
) # I logged the abundance divided by sampling effort in order to normalize the data

rm("dat")
save(dat_test, file = "dat_test.rds")
```

Loading the subset Brazil data that I had already saved as a .rds file
```{r load_predicts_brasil, cache=TRUE}
load("dat_test.rds")

dat_test <- dat_test %>%
      mutate(
        Class = as.factor(Class),
        log_sampling = log(Sampling_effort) # logging sampling effort to use as offset in Poisson model
        )

```

Plotting the relationship between year and abundance of arthropods in Brazil.
```{r plot_arthro_abund, cache=TRUE}
g_test <-  ggplot(data = dat_test, aes(x = Sample_midpoint, y = Measurement)) +
                  geom_point() +
                  stat_smooth(method = "lm", se = FALSE) +
                  facet_wrap(~Class, scales = "free")

g_test
```

Negative binomial model with offset to investigate the relationship between the predictor **year** and the response (count) abundance of arthropods in Brazil.
```{r brm_arthro_abund, cache=TRUE}
start_time <- Sys.time()
brm1 <- 
  brm(data = dat_test, family = negbinomial,
      Measurement ~ 1 + offset(log_sampling) + scaled_yr,
      prior = c(prior(normal(0, 10), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(gamma(0.01, 0.01), class = shape) # the brms default
                ), 
      cores = 2,
      chains = 2,
      iter = 1000) 
end_time <- Sys.time();end_time - start_time

save(brm1, file = "brm_arthro_abund.rda")
```

Loading arthropod abundance model to check summary and convergence.
```{r arthro_model_summary, cache=TRUE}
load("brm_arthro_abund.rda")

summary(brm1)
plot(brm1)

fixef(brm1)[1:2, ] %>%
  exp()

plot(marginal_effects(brm1))
pp_check(brm1)
```

> One idea I had today (Jun/24) was to use the predicted abundance of arthropods per year as a predictor in the wing length model to investigate whether bird body size is tracking temperature per se or food availability.

# References
