---
title: "Shrinking body size in Atlantic Forest birds as a response to climate change"
author: 
- Eduardo S.A. Santos
- João C.T. Menezes
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
bibliography: atlantic_bird_bibliography.bib  
---

# Introductory paragraph
Guidelines^[The paper should then start with an introductory paragraph (not abstract) of no more than 200 words, summarizing the background, rationale, main results (introduced by ‘Here we show’ or some equivalent phrase) and implications of the study. This paragraph should be fully referenced. Check other guidelines to **Nature Climate Change** here: https://www.nature.com/nclimate/about/content]

Climate change can cause rapid phenotypic change as a consequence of ecological and evolutionary processes [@ozgul_dynamics_2009].


# Data manipulation, plots, models
## Load libraries

```{r libraries, echo=FALSE,cache=TRUE}
library(tidyverse)
library(brms)
library(lme4)
```

## Load data
We used the Atlantic Bird Traits [@rodrigues_atlantic_2019] dataset as a source of information upon which to address our hypothesis.
```{r data,echo=FALSE, cache=TRUE}
birds <- read_csv("ATLANTIC_BIRD_TRAITS_completed_2018_11_d05.csv")
```

## New variables
Subsetting the original dataframe into sets to explore. We restricted our data collection to the order Passeriformes, as passerine birds represent over 82% of the entries in the Atlantic Bird Traits dataset (60,066 entries of a total of 72,483). We restricted our data to individuals that were recorded as adults during the original data collection (n = 43,149). Finally, we used records of individuals that were collected/recorded between 1990 and 2018, which yielded a total of 40,183 passerine entries in our dataset. I will only keep records of species that have at least 50 entries in the dataset. This will reduce the dataset from 435 passerine species to 143 species.
```{r newdataframes,cache=TRUE}
passer90 <- birds %>%
          filter(Year >= 1990, Age == "Adult", Order == "Passeriformes") %>%
          group_by(Binomial) %>%
          filter(n() >= 50)
```

Checking the range of years per species in the *passer90* dataframe:
```{r year_range, cache=TRUE}
by_binomial <- group_by(passer90, Binomial)
year_range <- summarise(passer90,
                        min = min(Year),
                        max = max(Year),
                        range = max-min
                        )

g_year_range <- ggplot(data = year_range, aes(x = range)) + 
                  geom_bar() +
                  theme_linedraw()

g_year_range
```


## Exploratory plots
Plotting some relationships:
```{r plots, cache=TRUE}
# g_wingl <- ggplot(data = passer90, aes(x = Year, y = Wing_length_right.mm., group = Binomial)) +
#            geom_point(aes(alpha = 0.09)) +
#            stat_smooth(method = "lm", se = FALSE, colour = "grey50", size = 0.5, alpha = 0.1) +
#            facet_wrap(~Family) + 
#            theme_linedraw()

# g_wingl

g0 <- ggplot(data = birdsCc, aes(x = Year, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g0

g_lat <- ggplot(data = birdsCc, aes(x = Latitude_decimal_degrees*-1, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g_lat


g0_Tr <- ggplot(data = birdsTf, aes(x = Year, y = Wing_length.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g0_Tr

g_lat_Tr <- ggplot(data = birdsTf, aes(x = Latitude_decimal_degrees*-1, y = Wing_length.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g_lat_Tr

g0_Hu <- ggplot(data = birdsHu, aes(x = Year, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g0_Hu

g_lat_Hu <- ggplot(data = birdsHu, aes(x = Latitude_decimal_degrees*-1, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g_lat_Hu

g0_Tm <- ggplot(data = birdsTm, aes(x = Year, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g0_Tm

g_lat_Tm <- ggplot(data = birdsTm, aes(x = Latitude_decimal_degrees*-1, y = Wing_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g_lat_Tm


g1 <- ggplot(data = birdsCc, aes(x = Year, y = Tarsus_length_right.mm., group = Sex, colour = Sex)) +
            geom_point() +
            stat_smooth(method = "lm", se = FALSE)

g1


g2 <- ggplot(data = passer90, aes(x = Year, y = Annual_mean_temperature)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g2

g3 <- ggplot(data = passer90, aes(x = Year, y = Annual_rainfall)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g3
```

## Exploratory models
A few quick models to check the trends. It seems like we have overall evidence that there is a negative relationship between *Year* and *Wing length* for all species.
```{r models, cache=TRUE}
m0 <- lmer(Wing_length_right.mm. ~ 1 + scale(Year) + (1 + scale(Year) | Binomial),
           data = passer90)

summary(m0)

#brm0 <- brm(Wing_length_right.mm. ~ 1 + scale(Year) + (1 + scale(Year) || Binomial),
#            data = passer90,
#            family = gaussian,
#            prior = c(prior(normal(0, 1), class = Intercept),
#                      prior(normal(0, 1), class = b),
#                      prior(cauchy(0, 1), class = sd)),
#            iter = 4000, 
#            warmup = 1000, 
#            chains = 4, 
#            cores = 4)

#summary(brm0)
#brms::launch_shinystan(brm0)
```

# References
