---
title: "Shrinking body size in Atlantic Forest birds as a response to climate change"
author: 
- Eduardo S.A. Santos
- João C.T. Menezes
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
bibliography: atlantic_bird_bibliography.bib  
---

# Introductory paragraph
Guidelines^[The paper should then start with an introductory paragraph (not abstract) of no more than 200 words, summarizing the background, rationale, main results (introduced by ‘Here we show’ or some equivalent phrase) and implications of the study. This paragraph should be fully referenced. Check other guidelines to **Nature Climate Change** here: https://www.nature.com/nclimate/about/content]

Two of the main responses of organisms to climate change are phenological changes and geographical range shifts [@parmesan_globally_2003]. A third response of organisms to global warming is increasingly being observed, the shrinkage of bodies [@teplitsky_climate_2014; @sheridan_shrinking_2011; @gardner_declining_2011]. Body shrinkage due to climate change has been hypothesized to be a consequence of ecological or evolutionary processes [@ozgul_dynamics_2009]. For instance, reductions in body size have been attributed to genetic microevolutionary responses to warming; smaller individuals dissipate heat better than larger individuals due to the larger ratio between surface:volume of their bodies [e.g., Bergmann's rule, @teplitsky_climate_2014]. The ecological hypothesis posits that climate change influences trophic interactions, limiting an organism's juveniles access to resources [@teplitsky_bergmanns_2008; @rode_reduced_2010].

Based on a large scale analysis of Neotropical passerines that inhabit the Brazilian Atlantic Forest, we show here that annual mean temperature has been increasing, while precipitation^[We need to choose what precipitation variable to use. It would be good to choose one that relates to the availability of food (insects or fruits).] has been decreasing over a period of **XX** years in the areas inhabited by these passerines.

# Data manipulation, plots, models
## Load libraries
```{r libraries, echo=FALSE,cache=TRUE}
library(tidyverse)
library(ggthemes) 
library(brms)
library(lme4)
library(ape)
library(geiger)
library(MCMCglmm)
```

## Load data
We used the Atlantic Bird Traits [@rodrigues_atlantic_2019] dataset as a source of information upon which to address our hypothesis.
```{r data,echo=FALSE, cache=TRUE}
birds <- read_csv("ATLANTIC_BIRD_TRAITS_completed_2018_11_d05.csv")
```

## New variables
Subsetting the original dataframe into sets to explore.

We restricted our data collection to the order Passeriformes, as passerine birds represent over 82% of the entries in the Atlantic Bird Traits dataset (60,066 entries of a total of 72,483). We restricted our data to individuals that were recorded as adults during the original data collection (n = 43,149). Finally, we used records of individuals that were collected/recorded between 1990 and 2018, which yielded a total of 40,183 passerine entries in our dataset. We will only keep records of species that have at least 50 entries in the dataset. This will reduce the dataset from 435 passerine species to 143 species.
```{r newdataframes,cache=TRUE}
passer90 <- birds %>%
          filter(Year >= 1990, Age == "Adult", Sex != "Unknown", Order == "Passeriformes") %>%
          mutate(
            ln_wing_length = log(Wing_length_right.mm.),
            scaled_lat = scale(Latitude_decimal_degrees*-1),
            scaled_yr = scale(Year),
            Binomial = gsub(" ", "_", Binomial),
            Binomial = gsub("Ceratopipra_rubrocapilla", "Pipra_rubrocapilla", Binomial),
            Binomial = gsub("Myiothlypis_leucoblephara", "Basileuterus_leucoblepharus", Binomial),
            Binomial = gsub("Myrmoderus_squamosus", "Myrmeciza_squamosa", Binomial),
            Binomial = gsub("Pyriglena_pernambucensis", "Pyriglena_leuconota", Binomial),
            Binomial = gsub("Rhopias_gularis", "Myrmotherula_gularis", Binomial),
            Binomial = gsub("Tangara_sayaca", "Thraupis_sayaca", Binomial),
            spp = Binomial
          ) %>%
          group_by(Binomial) %>%
          filter(n() >= 50) %>%
            mutate(
                  min = min(Year),
                  max = max(Year),
                  range = max-min
                        ) %>%
            filter(range >= 10)
```


Checking the range of years per species in the *passer90* dataframe:
```{r year_range, cache=TRUE}
by_binomial <- group_by(passer90, Binomial)
year_range <- summarise(by_binomial,
                  min = min(Year),
                  max = max(Year),
                  range = max-min
                        )

year_range <- as.data.frame(year_range)
row.names(year_range) <- year_range$Binomial # having rows named with the species name is required for any PCM

g_year_range <- ggplot(data = year_range, aes(x = range)) + 
                  geom_bar() +
                  theme_fivethirtyeight()

g_year_range

summary(year_range$range)

g_year_min <- ggplot(data = year_range, aes(x = min)) + 
                  geom_bar() +
                  theme_fivethirtyeight()
g_year_min

g_year_max <- ggplot(data = year_range, aes(x = max)) + 
                  geom_bar() +
                  theme_fivethirtyeight()
g_year_max
```

For each species in the dataset, the range of years of data available ranges from 2 to 28. It seems reasonable to use only those species for which we have at least 10 years of data. I haven't looked into what constitutes a good longitudinal dataset, but less than 10 years just seems too little in order to establish a link between temperature and body size.

After filtering out the species with less than 10 years of data, we ended up with a total of 134 passerine species.

## Exploratory plots
Plotting some relationships:
```{r plots, cache=TRUE}
g_wingl <- ggplot(data = passer90, aes(x = scaled_yr, y = scaled_wing_length, group = Binomial)) +
            geom_point(aes(alpha = 0.09)) +
            stat_smooth(method = "lm", se = FALSE, colour = "grey50", size = 0.5, alpha = 0.1) +
            facet_wrap(~Family, scales = "free_y") + 
            theme_fivethirtyeight()
g_wingl

g2 <- ggplot(data = passer90, aes(x = Year, y = Annual_mean_temperature)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g2

g3 <- ggplot(data = passer90, aes(x = Year, y = Annual_rainfall)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

g3
```

## Exploratory models
A few quick models to check the trends. It seems like we have overall evidence that there is a negative relationship between *Year* and *Wing length* for all species.
```{r models, cache=TRUE}
m0 <- lmer(Wing_length_right.mm. ~ scaled_yr + Sex + scaled_lat + (1 + scaled_yr | Binomial),
           data = passer90,
           REML = FALSE,
           control = lmerControl(optimizer ="Nelder_Mead")
           )

summary(m0)

m1 <- lmer(Body_mass.g. ~ scaled_yr + Sex + scaled_lat + (1 + scaled_yr | Binomial),
           data = passer90,
           REML = FALSE,
           control = lmerControl(optimizer ="Nelder_Mead")
           )

summary(m1)

```

Loading phylogeny to include in the multi-level model.
For all analyses, we used species-level Passeriformes trees taken from a sample of 1,000 topologies based on the Hackett backbone from Jetz et al.'s [-@jetz2012global] [BirdTree.org](http://birdtree.org). We will later prune these 5,966-species trees to fit the species in each analyses for which we have succesfully collected complete data.
```{r phylo, cache=TRUE}
trees <- read.nexus("Hackett_trees.nex")

namecheck_passer <- name.check(trees[[1]], year_range) # checking which species are in the phylogeny but not in our data frame

# namecheck_subset <- namecheck_passer$tree_not_data[grep("Thraupis", namecheck_passer$tree_not_data)] # Code to check if the species is available in the phylogeny under a different (synonim) genus name.

trees_passer <- lapply(trees, drop.tip, namecheck_passer$tree_not_data) # pruning these species from all the 1,000 trees

class(trees_passer) <- "multiPhylo"

nrow(year_range) # number of species for which we have complete data
length(trees_passer[[1]]$tip.label) # number of tips in the phylogenic trees

# The phylo object contains information on the relationship between species. Using this information, we can construct a covariance matrix of species.
inv.phylo <- inverseA(trees_passer[[1]], nodes = "TIPS", scale = TRUE)
A <- solve(inv.phylo$Ainv)
rownames(A) <- rownames(inv.phylo$Ainv)

```


Here is a multi-level Bayesian model of the response variable *Wing length* including the phylogeny of the species in question:
```{r model_phylo, cache=TRUE}
start_time <- Sys.time()
brm0 <- brm(Wing_length_right.mm. ~ 1 + Sex + scaled_yr + scaled_lat + (1 + scaled_yr || spp) + (1 | Binomial),
            cov_ranef = list(Binomial = A),
            data = passer90,
            family = gaussian,
            prior = c(prior(normal(71, 15), class = Intercept),
                      prior(normal(0, 10), class = b),
                      prior(cauchy(0, 1), class = sd),
                      prior(cauchy(0, 1), class = sigma)),
            iter = 3000, 
            warmup = 1000, 
            chains = 2, 
            cores = 2, 
            control = list(max_treedepth = 12))

end_time <- Sys.time();end_time - start_time

summary(brm0)
stanplot(brm0)
#brms::launch_shinystan(brm0)

plot(marginal_effects(brm0), points = TRUE)
pp_check(brm0)

hyp <- "sd_Binomial__Intercept^2 / (sd_Binomial__Intercept^2 + sd_spp__Intercept^2 + sigma^2) = 0"
(hyp <- hypothesis(brm0, hyp, class = NULL))

```

Here is a multi-level Bayesian model of the response variable *Body mass* including the phylogeny of the species in question. I want to try using the *get_prior()* function the next time I run this model^[https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/missing-data-and-other-opportunities.html].
```{r model_phylo_mass, cache=TRUE}
start_time <- Sys.time()
brm1 <- brm(Body_mass.g. ~ 1 + Sex + scaled_yr + scaled_lat + (1 + scaled_yr || spp) + (1 | Binomial),
            cov_ranef = list(Binomial = A),
            data = passer90,
            family = gaussian,
            prior = c(prior(normal(71, 15), class = Intercept),
                      prior(normal(0, 10), class = b),
                      prior(cauchy(0, 1), class = sd),
                      prior(cauchy(0, 1), class = sigma)),
            iter = 2000, 
            warmup = 1000, 
            chains = 3, 
            cores = 3, 
            control = list(max_treedepth = 12))

end_time <- Sys.time();end_time - start_time

summary(brm1)
stanplot(brm1)
#brms::launch_shinystan(brm0)

plot(marginal_effects(brm1), points = TRUE)
pp_check(brm1)

hyp <- "sd_Binomial__Intercept^2 / (sd_Binomial__Intercept^2 + sd_spp__Intercept^2 + sigma^2) = 0"
(hyp <- hypothesis(brm1, hyp, class = NULL))

```

# References
