---
title: "Weak and species-specific morphological change in Atlantic Forest birds across two decades of warming"
author:
  - name: Eduardo S.A. Santos
    orcid: 0000-0002-0434-3655
    corresponding: true
    email: esantos2@ualberta.ca
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Validation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: cossee
  - name: Ayumi Mizuno
    orcid: 0000-0003-0822-5637
    corresponding: false
    roles:
      - Investigation
      - Methodology
      - Validation
      - Writing – review & editing
    affiliations:
      - ref: cossee
  - name: Santiago Ortega
    orcid: 0000-0002-3518-276X
    corresponding: false
    roles:
      - Investigation
      - Methodology
      - Validation
      - Writing – review & editing
    affiliations:
      - ref: cossee
  - name: Gustavo Burin
    orcid: 0000-0001-7851-0666
    corresponding: false
    roles:
      - Investigation
      - Methodology
      - Validation
      - Visualization
      - Writing – review & editing
    affiliations:
      - ref: gu
affiliations:
  - id: cossee
    name: "Collaboration for Open Science and Synthesis in Ecology and Evolution, Department of Biological Sciences, University of Alberta, Edmonton, AB, T6G 2E9, Canada"
  - id: gu
    name: "Department of Biological and Environmental Sciences, University of Gothenburg, Sweden"
subtitle: "Running title: Morphological change in Atlantic Forest birds"
keywords:
  - body size
  - climate change
  - phenotypic variation
  - macroecology
  - Atlantic Forest
  - phylogenetic comparative analysis
abstract: |
  Birds are widely expected to shrink as the climate warms, but few tropical studies span decades. In the central Amazon, mean body mass fell in all 77 species studied and a third of species lengthened their wings. We analysed 24 years (1995–2018) of morphology from 73 passerine species (12,571 live-capture records; ATLANTIC BIRD TRAITS) across the Brazilian Atlantic Forest, a region that warmed by about 0.25°C per decade. The records came from 47 research programmes whose sites changed over time, so we modelled contributor and municipality explicitly and checked by simulation that these models recover trends planted in the real sampling structure. Unadjusted models suggested a steep wing-length decline and increases in bill width and trait variability. Once sampling turnover was modelled, the wing decline was less than half as steep (−1.0% per decade, 95% credible interval spanning zero), and the bill-width and variability changes disappeared. Body mass declined by 0.9% per decade, close to the mean Amazonian rate of 1.0%, but its interval spanned zero (−2.8 to +1.0%). Responses varied strongly among species, and most species showed no change distinguishable from zero. Wings did not lengthen, and among individuals measured for both traits, mass tended to be retained relative to wing length. Neither trait was associated with interannual temperature anomalies. Atlantic Forest passerines thus show neither the consistent mass loss across species nor the wing elongation reported from the Amazon, although their average mass change is compatible with the Amazonian rate. Compiled trait databases can extend long-term monitoring to under-studied tropical regions, but only when sampling structure is modelled: ignoring it here would have more than doubled the estimated wing decline.
summary-statement: |
  Atlantic Forest birds lost mass at about the Amazonian average rate but without its consistency across species, and modelling sampling turnover more than halved the apparent wing decline.
plain-language-summary: |
  As global temperatures rise, many animals are expected to become smaller or change shape to cope with heat. In the intact Amazon rainforest, every bird species studied became lighter over four decades, and many grew longer wings. We asked whether birds in another tropical forest, the Brazilian Atlantic Forest, changed in the same way, using 24 years of field measurements (1995–2018) of 73 perching-bird species. On average, Atlantic Forest birds became lighter at about the same rate as Amazonian birds, but the estimate was uncertain and species went in both directions rather than all declining together. Wings did not grow longer, and bills did not change. The data came from many research groups whose field sites changed over the years, and simpler analyses exaggerated how much the birds had changed: allowing for who collected the data, and where, more than halved the apparent shortening of wings. Computer simulations confirmed that this correction recovers real trends rather than hiding them.
date: last-modified
bibliography: references.bib
citation:
  container-title: bioRxiv
number-sections: true
---

```{r load-estimates, include=FALSE}
desc        <- readRDS("../Analysis/output/descriptive_summary.rds")
cw_phylo    <- readRDS("../Analysis/output/controlled_wing_phylo_results.rds")
mass        <- readRDS("../Analysis/output/mass_results.rds")
multi_phylo <- readRDS("../Analysis/output/multitrait_phylo_results.rds")
biv_phylo   <- readRDS("../Analysis/output/bivariate_phylo_results.rds")
var_phylo   <- readRDS("../Analysis/output/variance_phylo_results.rds")
diet_phylo  <- readRDS("../Analysis/output/diet_interaction_phylo.rds")
clim        <- readRDS("../Analysis/output/climate_trends.rds")
es          <- readRDS("../Analysis/output/effect_scale.rds")
# format a count with a thousands separator for inline use
.n <- function(x) format(x, big.mark = ",", trim = TRUE)
# format a signed estimate / CI bound to 2 dp with a true minus sign
.s <- function(x) sub("-", "−", sprintf("%.2f", x))
# format a signed value to 1 dp with a true minus sign
.s1 <- function(x) sub("-", "−", sprintf("%.1f", x))
# format a signed estimate to 4 dp with a true minus sign
.s4 <- function(x) sub("-", "−", sprintf("%.4f", x))
# format a p-value for inline use
.rp <- function(p) if (p < 0.001) "< 0.001" else paste("=", sprintf("%.3f", p))

# pull row from cw_phylo$before_after with optional term filtering
.cw <- function(mod, term = NULL) {
  r <- cw_phylo$before_after[cw_phylo$before_after$model == mod, ]
  if (!is.null(term)) r <- r[r$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", lo = "NA", hi = "NA", dec = "NA", dec_lo = "NA", dec_hi = "NA", n = "NA"))
  list(
    b      = .s(r$estimate[1]),
    lo     = .s(r$ci_lo[1]),
    hi     = .s(r$ci_hi[1]),
    dec    = .s1(r$mm_per_decade[1]),
    dec_lo = .s1(r$mm_per_decade_lo[1]),
    dec_hi = .s1(r$mm_per_decade_hi[1]),
    n      = .n(r$N[1])
  )
}

cw_m0           <- .cw("M0")
cw_m1a          <- .cw("M1a_src")
cw_m1           <- .cw("M1")
cw_m2           <- .cw("M2")
cw_m3           <- .cw("M3")
# (random-effect SDs of the fully adjusted wing model are taken from the Stan tier below)
cw_m3_cc        <- .cw("M3_cc")
cw_m5src_within <- .cw("M5src", "yr_within_src")
cw_m5src_mean   <- .cw("M5src", "yr_src_mean")
cw_m7           <- .cw("M7")
cw_noanom       <- .cw("M3_noanom")
cw_long         <- .cw("M3_longrun")

# effect scale
if (!is.null(es)) {
  .esw <- es$wing$using_scaling_sd
  wing_pct_dec    <- .esw["est", "pct_per_decade"]
  wing_pct_rec    <- .esw["est", "pct_over_record"]
  wing_pct_rec_lo <- .esw["lo",  "pct_over_record"]
  wing_pct_rec_hi <- .esw["hi",  "pct_over_record"]
} else {
  wing_pct_dec <- -2.58; wing_pct_rec <- -5.9; wing_pct_rec_lo <- -9.0; wing_pct_rec_hi <- -2.8
}

# multi-trait comparisons from multi_phylo
.mt_phylo <- function(trait_name, mod, term = "scaled_yr") {
  df <- multi_phylo$comparison_lme4
  r <- df[df$trait == trait_name & df$model == mod & df$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", b2 = "NA", lo = "NA", hi = "NA", lo2 = "NA", hi2 = "NA", pct = "NA", pct_lo = "NA", pct_hi = "NA", n = "NA"))
  list(
    b      = .s4(r$phylo_estimate[1]),
    b2     = .s(r$phylo_estimate[1]),
    lo     = .s4(r$phylo_lower[1]),
    hi     = .s4(r$phylo_upper[1]),
    lo2    = .s(r$phylo_lower[1]),
    hi2    = .s(r$phylo_upper[1]),
    pct    = .s1(r$phylo_pct_per_decade[1]),
    pct_lo = .s1(r$phylo_pct_lower[1]),
    pct_hi = .s1(r$phylo_pct_upper[1]),
    n      = .n(r$phylo_n[1])
  )
}

multi_lme4 <- readRDS("../Analysis/output/multitrait_results.rds")
.mt_lme4 <- function(trait_name, mod, term = "scaled_yr") {
  df <- multi_lme4$table
  r <- df[df$trait == trait_name & df$model == mod & df$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", b2 = "NA", lo = "NA", hi = "NA", lo2 = "NA", hi2 = "NA", pct = "NA", pct_lo = "NA", pct_hi = "NA", n = "NA"))
  list(
    b      = .s4(r$estimate[1]),
    b2     = .s(r$estimate[1]),
    lo     = .s4(r$lower[1]),
    hi     = .s4(r$upper[1]),
    lo2    = .s(r$lower[1]),
    hi2    = .s(r$upper[1]),
    pct    = .s1(r$pct_per_decade[1]),
    pct_lo = .s1(r$pct_per_decade_lower[1]),
    pct_hi = .s1(r$pct_per_decade_upper[1]),
    n      = .n(r$n[1])
  )
}

mass_m0     <- .mt_phylo("ln_body_mass", "M0_baseline")
mass_m1a    <- .mt_lme4("ln_body_mass", "M1a_src")
mass_m1     <- .mt_phylo("ln_body_mass", "M1_src_site")
mass_m3     <- .mt_phylo("ln_body_mass", "M3_ring")
mass_within <- .mt_phylo("ln_body_mass", "MWsrc_mundlak_contributor", "yr_within_src")

bill_m0     <- .mt_phylo("Bill_width.mm.", "M0_baseline")
bill_m1a    <- .mt_lme4("Bill_width.mm.", "M1a_src")
bill_m1     <- .mt_phylo("Bill_width.mm.", "M1_src_site")
bill_m3     <- .mt_phylo("Bill_width.mm.", "M3_ring")
bill_within <- .mt_phylo("Bill_width.mm.", "MWsrc_mundlak_contributor", "yr_within_src")

# variance results
var_s0 <- var_phylo$comparison[var_phylo$comparison$tier == "S0" & var_phylo$comparison$term == "scaled_yr", ]
var_s3 <- var_phylo$comparison[var_phylo$comparison$tier == "S3" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_yr <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_tm <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_tmean", ]

# diet interaction
diet_base <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_ctrl <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A_src_site" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_p90  <- diet_phylo$quantile_slopes_all_specs[diet_phylo$quantile_slopes_all_specs$model == "A_src_site" & diet_phylo$quantile_slopes_all_specs$quantile == "p90", ]

# climate trends
clim_tm <- clim$pooled[clim$pooled$variable == "tmean" & clim$pooled$weighting == "unweighted + year RE", ]
.clim_row <- function(v, dp) { r <- clim$pooled[clim$pooled$variable == v & clim$pooled$weighting == "unweighted + year RE", ]
  f <- function(x) sub("-", "−", sprintf(paste0("%+.", dp, "f"), x))
  list(b = f(r$slope_per_decade), lo = f(r$lower), hi = f(r$upper), est = r$slope_per_decade) }
clim_t <- .clim_row("tmean", 3); clim_p <- .clim_row("prec_annual", 1); clim_s <- .clim_row("spei12_dec", 3)
clim_t_dec2 <- sprintf("%.2f", clim_t$est)                    # rounded rate for the Discussion
clim_t_tot  <- sprintf("%+.2f", clim_t$est * (2018 - 1995) / 10) # linear warming over 1995-2018

# ---- Bayesian (Stan) tier: the source of every model estimate in the text ----
# bayes_summary.rds from bayes_aggregate.R (posterior pooled over 20 trees);
# bayes_derived.rds from bayes_derived.R (species slopes, diet percentiles).
bayes <- readRDS("../Analysis/output/bayes/bayes_summary.rds")
bder  <- readRDS("../Analysis/output/bayes/bayes_derived.rds")
DEC   <- bder$dec                      # SD-of-year units per decade
.p2   <- function(x) sprintf("%.2f", x)
# posterior probability with its relation sign: "= 0.83", or "> 0.99" / "< 0.01"
# when it rounds to 1 or 0 (a posterior probability is never exactly 1)
.pp   <- function(x) if (x >= 0.995) "> 0.99" else if (x < 0.005) "< 0.01" else paste("=", .p2(x))
# signed formatters with a true minus sign and an explicit plus
.ss1 <- function(x) sub("-", "−", sprintf("%+.1f", x)); .ss <- function(x) sub("-", "−", sprintf("%+.2f", x))
.ss4 <- function(x) sub("-", "−", sprintf("%+.4f", x))
# one coefficient of one fit; scale = "mm" (per-decade on the response scale),
# "pct" (log response: % per decade), "sigma" (log residual SD: % per decade),
# "raw" (per unit of the covariate, e.g. per degree C of anomaly)
.bz <- function(fit, par, scale = c("mm", "mm2", "pct", "sigma", "raw")) {
  scale <- match.arg(scale)
  r <- bayes$fixed[bayes$fixed$fit == fit & bayes$fixed$par == par, ]
  na <- list(b = "NA", lo = "NA", hi = "NA", P = "NA", Ppos = "NA", n = "NA")
  if (nrow(r) != 1) return(na)
  f <- switch(scale, mm = function(x) .ss1(x * DEC), mm2 = function(x) .ss(x * DEC), pct = function(x) .ss1(100 * (exp(x * DEC) - 1)),
              sigma = function(x) .ss1(100 * (exp(x * DEC) - 1)), raw = function(x) .ss4(x))
  n <- bayes$meta$n[bayes$meta$fit == fit]
  list(b = f(r$estimate), lo = f(r$lower), hi = f(r$upper), P = .pp(r$p_neg), Ppos = .pp(1 - r$p_neg),
       n = .n(n), est = r$estimate)
}
# sampler diagnostics and agreement with REML, over the fits quoted in the text
local({
  d <- bayes$diag; cm <- bayes$comparison[!grepl("__pcor$", bayes$comparison$fit) &
    grepl("scaled_yr|dT|yr_within|yr_src|yr_mean|:", bayes$comparison$par), ]
  dg <- bder$diag_focal; ds <- bder$diag_sd_spp_int
  bz <<- list(n_fits = nrow(bayes$meta), n_trees = max(bayes$fixed$n_trees),
              foc_rhat = sprintf("%.3f", dg[["rhat_max"]]), foc_ess = .n(round(dg[["ess_bulk_pooled_min"]])),
              foc_ess_tail = .n(round(dg[["ess_tail_pooled_min"]])),
              sd_rhat = sprintf("%.3f", ds[["rhat_max"]]), sd_ess = .n(round(ds[["ess_bulk_min"]])),
              rhat_max = sprintf("%.3f", max(d$rhat_max)),
              ess_min = .n(round(min(d$ess_bulk_min))),
              shift_max = sprintf("%.2f", max(abs(cm$diff_in_sd))),
              sd_ratio_max = sprintf("%.2f", max(cm$sd_ratio)))
})
.fid <- c(
  w0 = "atlantic_parallel_controlled__001__conc.wing.length__M0",
  w3 = "atlantic_parallel_controlled__011__conc.wing.length__M3",
  w3cc = "atlantic_parallel_controlled__023__conc.wing.length__M3_cc",
  w5 = "atlantic_parallel_controlled__015__conc.wing.length__M5src",
  w5cc = "atlantic_parallel_controlled__027__conc.wing.length__M5src_cc",
  w3pcor = "atlantic_parallel_controlled__011__conc.wing.length__M3__pcor",
  m0 = "atlantic_multitrait__001__y", m3 = "atlantic_multitrait__003__y", mw = "atlantic_multitrait__004__y",
  bl0 = "atlantic_multitrait__005__y", bl1 = "atlantic_multitrait__006__y",
  bl3 = "atlantic_multitrait__007__y", blw = "atlantic_multitrait__008__y",
  s1 = "atlantic_variance_sigma__002__conc.wing.length__S1", s3 = "atlantic_variance_sigma__004__conc.wing.length__S3",
  s7 = "atlantic_variance_sigma__009__conc.wing.length__S7",
  dA0 = "atlantic_diet_interaction__002__conc.wing.length", dA1 = "atlantic_diet_interaction__006__conc.wing.length",
  dB1 = "atlantic_diet_interaction__012__conc.wing.length",
  th_w = "referee_reruns__001__wing__site", th_m = "referee_reruns__003__lnmass__site",
  th_i = "referee_reruns__005__iso__site", th_iy = "referee_reruns__006__iso__site",
  th_id = "referee_reruns__013__iso__site", th_il = "referee_reruns__014__iso__site",
  # year-only isometry models on all shared records (no anomaly required;
  # bayes_export_isometry_all.R); the thermal models above use the anomaly subset
  iso_p = "referee_reruns__015__iso__all", iso_f = "referee_reruns__016__iso__all",
  iso_fpcor = "referee_reruns__016__iso__all__pcor",
  lnw = "referee_reruns__017__lnwing__all", lnm = "referee_reruns__018__lnmass__all")
# wing: community year slopes (mm/decade) and % per decade via the effect-scale factor
b_w0 <- .bz(.fid["w0"], "b_scaled_yr"); b_w3 <- .bz(.fid["w3"], "b_scaled_yr")
b_w3cc <- .bz(.fid["w3cc"], "b_scaled_yr")
b_w5in <- .bz(.fid["w5"], "b_yr_within_src"); b_w5btw <- .bz(.fid["w5"], "b_yr_src_mean")
b_w5ccin <- .bz(.fid["w5cc"], "b_yr_within_src")
b_w3pcor <- .bz(.fid["w3pcor"], "b_scaled_yr")
.wpct <- if (!is.null(es)) wing_pct_dec / cw_phylo$before_after$mm_per_decade[cw_phylo$before_after$model == "M0"][1] else NA
b_w0_pct <- .ss1(b_w0$est * DEC * .wpct); b_w3_pct <- .ss1(b_w3$est * DEC * .wpct)
b_w_atten <- sprintf("%.0f", 100 * (1 - b_w3$est / b_w0$est))
# species slopes under the fully adjusted wing model
sc <- as.list(round(bder$species_counts)); sc$median <- .ss(bder$species_counts[["median_mm_decade"]])
# body mass (% per decade)
b_m0 <- .bz(.fid["m0"], "b_scaled_yr", "pct"); b_m3 <- .bz(.fid["m3"], "b_scaled_yr", "pct")
b_mw <- .bz(.fid["mw"], "b_yr_within_src", "pct")
# bill width (mm per decade)
b_bl0 <- .bz(.fid["bl0"], "b_scaled_yr", "mm2"); b_bl1 <- .bz(.fid["bl1"], "b_scaled_yr", "mm2")
b_bl3 <- .bz(.fid["bl3"], "b_scaled_yr", "mm2"); b_blw <- .bz(.fid["blw"], "b_yr_within_src", "mm2")
# residual SD (% per decade; temperature: % per SD of temperature)
b_s1 <- .bz(.fid["s1"], "sigma_scaled_yr", "sigma"); b_s3 <- .bz(.fid["s3"], "sigma_scaled_yr", "sigma")
b_s7t <- local({ r <- bayes$fixed[bayes$fixed$fit == .fid[["s7"]] & bayes$fixed$par == "sigma_scaled_tmean", ]
  if (nrow(r) != 1) list(b = "NA", lo = "NA", hi = "NA", P = "NA") else
    list(b = .ss1(100 * (exp(r$estimate) - 1)), lo = .ss1(100 * (exp(r$lower) - 1)),
         hi = .ss1(100 * (exp(r$upper) - 1)), P = .pp(r$p_neg)) })
# diet: year x invertebrate share (mm per SD year per SD diet), category contrast, percentile slopes
.bzraw2 <- function(fit, par) { x <- .bz(fit, par, "raw"); if (x$b == "NA") return(x)
  r <- bayes$fixed[bayes$fixed$fit == fit & bayes$fixed$par == par, ]
  modifyList(x, list(b = .ss(r$estimate), lo = .ss(r$lower), hi = .ss(r$upper))) }
b_dA0 <- .bzraw2(.fid[["dA0"]], "b_scaled_yr:diet_inv_std"); b_dA1 <- .bzraw2(.fid[["dA1"]], "b_scaled_yr:diet_inv_std")
b_dB1 <- .bzraw2(.fid[["dB1"]], "b_scaled_yr:diet_cat2Invertebrate")
.dq <- function(k) { q <- bder$diet_quantile_slopes; r <- q[q$quantile == k, ]
  list(b = .ss(r$estimate), lo = .ss(r$lower), hi = .ss(r$upper), P = .pp(r$p_neg), inv = r$diet_inv) }
b_p10 <- .dq("p10"); b_p90 <- .dq("p90")
# thermal anomaly (per degree C) and isometry (% per decade)
b_thw <- .bzraw2(.fid[["th_w"]], "b_dT"); b_thm <- .bz(.fid["th_m"], "b_dT", "raw")
b_thi <- .bz(.fid["th_i"], "b_dT", "raw"); b_thiy <- .bz(.fid["th_iy"], "b_dT", "raw")
b_thiy_yr <- .bz(.fid["th_iy"], "b_scaled_yr", "pct")
b_thid <- .bz(.fid["th_id"], "b_dT_detr", "raw"); b_thil <- .bz(.fid["th_il"], "b_dT", "raw")
b_isop <- .bz(.fid["iso_p"], "b_scaled_yr", "pct"); b_isof <- .bz(.fid["iso_f"], "b_scaled_yr", "pct")
b_isofpcor <- .bz(.fid["iso_fpcor"], "b_scaled_yr", "pct")
# the same fully adjusted contrast on the anomaly subset (Methods)
b_isof_clim <- .bz("referee_reruns__010__iso__site", "b_scaled_yr", "pct")
b_lnw <- .bz(.fid["lnw"], "b_scaled_yr", "pct"); b_lnm <- .bz(.fid["lnm"], "b_scaled_yr", "pct")
b_iso_exp <- if (b_lnw$b == "NA") "NA" else .ss1(100 * ((exp(b_lnw$est * DEC))^3 - 1))
# the contrast implied by the two separate fits (log mass slope minus 3 x log wing slope)
b_iso_sep <- if (b_lnw$b == "NA" || b_lnm$b == "NA") "NA" else .ss1(100 * (exp((b_lnm$est - 3 * b_lnw$est) * DEC) - 1))
n_iso_spp <- sum(bayes$species_slope_dev$fit == .fid[["iso_f"]])
# species posterior means (fixed + deviation) of the shared-record wing and mass
# slopes: how many fall in the longer-wing / lower-mass quadrant (Fig. isometry B)
n_iso_decoupled <- local({
  sp <- function(fit) { b0 <- bayes$fixed$estimate[bayes$fixed$fit == fit & bayes$fixed$par == "b_scaled_yr"]
    d <- bayes$species_slope_dev[bayes$species_slope_dev$fit == fit, ]; setNames(b0 + d$estimate, d$par) }
  w <- sp(.fid[["lnw"]]); m <- sp(.fid[["lnm"]])[names(w)]
  if (!length(w)) NA else sum(w > 0 & m < 0, na.rm = TRUE) })
# random-effect SDs (mm) of the fully adjusted wing model, posterior means
.vc <- function(par) .s(bayes$varcomp$estimate[bayes$varcomp$fit == .fid[["w3"]] & bayes$varcomp$par == par])
cw_sd_src <- .vc("sd_src"); cw_sd_site <- .vc("sd_site")

# posterior summary of a variance-component parameter, e.g. h2_slope
.vcz <- function(fit, par) { r <- bayes$varcomp[bayes$varcomp$fit == fit & bayes$varcomp$par == par, ]
  if (nrow(r) != 1) return(list(b = "NA", lo = "NA", hi = "NA"))
  m2 <- function(x) sub("-", "−", sprintf("%.2f", x))
  list(b = m2(r$estimate), lo = m2(r$lower), hi = m2(r$upper)) }
h2s_w <- .vcz(.fid[["w3pcor"]], "h2_slope"); h2s_i <- .vcz(.fid[["iso_fpcor"]], "h2_slope")
cor_w <- .vcz(.fid[["w3pcor"]], "cor_phylo_int_slope")
# REML likelihood-ratio test for a phylogenetic slope component (atlantic_phylo_slopes.R, wing fully adjusted)
.lrt <- readRDS("../Analysis/output/phylo_slopes/phylo_slopes_results.rds")$wing_M3$lrt
lrt_w <- list(sig = sum(.lrt$p_boundary < 0.05, na.rm = TRUE), n = sum(is.finite(.lrt$p_boundary)),
              p_med = sprintf("%.2f", median(.lrt$p_boundary, na.rm = TRUE)))
# lagged anomalies, REML across 50 trees (referee_anomaly_lag.R)
lag <- readRDS("../Analysis/output/referee_reruns/anomaly_lag.rds")
.lg <- function(tbl, response, term, dp = 4) { r <- lag[[tbl]]; r <- r[r$response == response & r$term == term, ]
  stopifnot(nrow(r) == 1); fm <- function(x) sub("-", "−", sprintf(paste0("%+.", dp, "f"), x))
  list(b = fm(r$estimate), lo = fm(r$lower), hi = fm(r$upper), p = .rp(r$p)) }
lg1_i <- .lg("L1", "isometry contrast lnM-3lnL", "dT_lag1"); lg12_i <- .lg("L12", "isometry contrast lnM-3lnL", "dT_lag12")
lg1_w <- .lg("L1", "wing length (mm)", "dT_lag1", 2); lg12_w <- .lg("L12", "wing length (mm)", "dT_lag12", 2)
lg1_m <- .lg("L1", "log body mass", "dT_lag1"); lg12_m <- .lg("L12", "log body mass", "dT_lag12")

rr          <- readRDS("../Analysis/output/referee_reruns/referee_reruns.rds")

# accessors for the referee re-run tables (Analysis/scripts/referee_reruns.R).
# Every number quoted from these blocks is pulled from the saved object, never
# typed into the text.
.rr_row <- function(tbl, ...) {
  f <- list(...); keep <- rep(TRUE, nrow(tbl))
  for (nm in names(f)) keep <- keep & tbl[[nm]] == f[[nm]]
  r <- tbl[keep, , drop = FALSE]
  stopifnot(nrow(r) == 1L)
  r
}
# thermal-anomaly effects (E1-E3)
rr_therm <- function(model, response, term = "dT", tbl = NULL) {
  tb <- if (is.null(tbl)) rr$E3 else tbl
  r <- .rr_row(tb, model = model, response = response, term = term)
  list(z = .s(r$z), p = .rp(r$p), b = .s4(r$estimate), se = sprintf("%.4f", r$se),
       lo = .s4(r$lower), hi = .s4(r$upper), trees = r$trees_used)
}
# isometry under the M3 adjustment set (E4)
rr_iso <- function(model) {
  r <- .rr_row(rr$E4, model = model, response = "isometry contrast lnM-3lnL")
  list(pct = .s1(r$pct_decade), lo = .s1(r$pct_decade_lo), hi = .s1(r$pct_decade_hi),
       z = .s(r$z), p = .rp(r$p))
}
# thermal-anomaly rows quoted in the Results
rr_iso_therm     <- rr_therm("E3_M3_adjusted_phylo",  "isometry contrast lnM-3lnL")
rr_wing_therm    <- rr_therm("E3_M3_adjusted_phylo",  "wing length (mm)")
rr_mass_therm    <- rr_therm("E3_M3_adjusted_phylo",  "log body mass")
rr_iso_therm_e1  <- rr_therm("E1a_anomaly_plus_year", "isometry contrast lnM-3lnL", tbl = rr$E1)
rr_iso_therm_e1b <- rr_therm("E1b_detrended_anomaly", "isometry contrast lnM-3lnL",
                             term = "dT_detr", tbl = rr$E1)
rr_iso_therm_e2  <- rr_therm("E2a_locyear_random_intercept", "isometry contrast lnM-3lnL",
                             tbl = rr$E2)
rr_iso_yr        <- rr_therm("E1a_anomaly_plus_year", "isometry contrast lnM-3lnL",
                             term = "scaled_yr", tbl = rr$E1)
# isometry contrast under the full M3 adjustment set (E4)
rr_iso_m3        <- rr_iso("E4b_iso_M3")
# sample sizes for the thermal models, taken from the fitted tables / data
# the shared-record sample of the Stan anomaly fits (the REML E1/E2 tables use a
# slightly larger sample and are not quoted), and of the year-only isometry fits
rr_n_therm  <- bayes$meta$n[bayes$meta$fit == .fid[["th_i"]]]
n_iso       <- bayes$meta$n[bayes$meta$fit == .fid[["iso_f"]]]
# Counted from passer90.rda (ATLANTIC BIRD TRAITS only). The same count comes
# from passer90_climate.rds, but that file carries WorldClim-derived columns
# whose terms forbid redistribution, and no climate column is needed here.
rr_n_shared <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  nrow(subset(e$passer90, !is.na(conc.wing.length) & !is.na(ln_body_mass)))
})
# temporal coverage within contributors for the wing sample (context for the
# within/between-contributor decomposition): contributors sampled over at least
# `span_min` calendar years, their records, and the share of capture-year variance
# lying within contributors
src_cov <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  d <- subset(e$passer90, !is.na(conc.wing.length))
  span <- tapply(d$Year, d$Main_researcher, function(y) max(y) - min(y) + 1)
  nrec <- tapply(d$Year, d$Main_researcher, length)
  yw <- d$Year - ave(d$Year, d$Main_researcher)
  span_min <- 10
  list(span_min = span_min, n_src = length(span), n_long = sum(span >= span_min),
       rec_long = sum(nrec[span >= span_min]),
       pct_within = sprintf("%.0f", 100 * var(yw) / var(d$Year)))
})
# the same temporal coverage within municipalities (context for the
# within/between-municipality decomposition, M5)
site_cov <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  d <- subset(e$passer90, !is.na(conc.wing.length))
  span <- tapply(d$Year, d$Municipality, function(y) max(y) - min(y) + 1)
  nrec <- tapply(d$Year, d$Municipality, length)
  yw <- d$Year - ave(d$Year, d$Municipality)
  list(n_site = length(span), n_long = sum(span >= src_cov$span_min),
       rec_long = sum(nrec[span >= src_cov$span_min]),
       pct_within = sprintf("%.0f", 100 * var(yw) / var(d$Year)))
})
# contributors and municipalities in the full sample and in the wing sample
prov <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  d <- e$passer90; w <- subset(d, !is.na(conc.wing.length))
  list(n_src = length(unique(d$Main_researcher)), n_site = length(unique(d$Municipality)),
       n_src_w = length(unique(w$Main_researcher)), n_site_w = length(unique(w$Municipality)))
})
# Methods counts pulled from passer90.rda (sampling hierarchy, provenance audit)
meth <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e); d <- e$passer90
  w  <- subset(d, !is.na(conc.wing.length)); sh <- subset(w, !is.na(ln_body_mass)); m <- subset(d, !is.na(ln_body_mass))
  one <- function(x) { t <- tapply(x$Main_researcher, x$Municipality, function(s) length(unique(s)))
    list(single = sum(t == 1), n = length(t), pct = sprintf("%.1f", 100 * mean(t == 1))) }
  pc <- function(x) sprintf("%.0f", 100 * x)
  r2 <- function(f) summary(lm(Year ~ factor(f), data = w))$r.squared
  ea <- subset(w, Year <= 2006); la <- subset(w, Year >= 2013)
  s_ea <- unique(ea$Main_researcher); s_la <- unique(la$Main_researcher)
  list(n_shared = nrow(sh), one_sh = one(sh), one_m = one(m),
       site_multi = pc(mean(tapply(w$Main_researcher, w$Municipality, function(s) length(unique(s))) > 1)),
       src_multi = pc(mean(tapply(w$Municipality, w$Main_researcher, function(s) length(unique(s))) > 1)),
       r2_site = pc(r2(w$Municipality)), r2_src = pc(r2(w$Main_researcher)),
       lat_ea = sub("-", "−", sprintf("%.1f", mean(ea$Latitude_decimal_degrees))),
       lat_la = sub("-", "−", sprintf("%.1f", mean(la$Latitude_decimal_degrees))),
       alt_ea = .n(median(ea$Altitude, na.rm = TRUE)), alt_la = .n(median(la$Altitude, na.rm = TRUE)),
       src_ea = length(s_ea), src_la = length(s_la), src_both = length(intersect(s_ea, s_la)))
})
# REML checks quoted alongside the Stan estimates (glmmTMB, pooled over 50 trees)
cw_m5site_within <- .cw("M5", "yr_within_site")
cw_m6_unk <- .cw("M6")                     # unknown-sex wing records, full M3 controls
# residual-SD year effect with species intercepts in the residual-SD model (REML S6)
var_s6 <- local({ cm <- var_phylo$comparison
  r <- cm[!is.na(cm$tier) & cm$tier == "S6" & cm$component == "sigma" & cm$par == "scaled_yr", ]
  f <- function(x) sub("-", "−", sprintf("%+.1f", x))
  list(b = f(r$pct_decade_phylo), lo = f(r$pct_decade_lo_phylo), hi = f(r$pct_decade_hi_phylo)) })
cw_ss <- cw_phylo$sample_sizes
n_anom_blocks <- nrow(cw_ss$anomalous_blocks); n_anom_rec <- cw_ss$anomalous_records
molt_unrec_pct <- sprintf("%.0f", 100 * cw_ss$molt_unrecorded_share)
# contributors that never recorded moult in the wing sample
molt_src_none <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  d <- subset(e$passer90, !is.na(conc.wing.length))
  sum(tapply(is.na(d$Molt), d$Main_researcher, all))
})
lrt_w$n_fail <- nrow(.lrt) - lrt_w$n
# the focal parameter with the largest per-tree R-hat, and the smallest per-tree ESS
rhat_top <- local({
  d <- bder$diag_terms; d <- d[d$variable != "sd_spp_int", ]
  top <- d[which.max(d$rhat), ]
  list(fit = top$fit, par = top$variable, n_gt_101 = sum(d$rhat > 1.01), n_terms = nrow(d),
       ess_bulk_min = .n(round(min(d$ess_bulk))), ess_tail_min = .n(round(min(d$ess_tail))))
})
```

```{r review-checks, include=FALSE}
# ---- checks added for the Oikos submission (internal review, 2026-09-30) ----
# Amazonian benchmark from Jirinec et al. (2021) Table S6 (jirinec_benchmark.R)
jb  <- readRDS("../Analysis/output/jirinec_benchmark.rds")
.jb <- function(trait) jb$benchmark[jb$benchmark$set == "passerines" & jb$benchmark$trait == trait, ]
amz_m <- .jb("Body mass"); amz_w <- .jb("Wing length")
amz_m_dec <- .ss1(amz_m$mean_pct_dec); amz_w_dec <- .ss1(amz_w$mean_pct_dec)
amz_m_min <- .ss1(amz_m$min_pct_dec); amz_m_max <- .ss1(amz_m$max_pct_dec)
amz_m_degc <- sprintf("%s to %s", .ss1(min(amz_m$per_degC_lo, amz_m$per_degC_hi)), .ss1(max(amz_m$per_degC_lo, amz_m$per_degC_hi)))
amz_warm <- sprintf("%.2f–%.2f", min(jb$warming_bdffp_per_decade), max(jb$warming_bdffp_per_decade))
.af <- function(trait) { r <- jb$ours[jb$ours$trait == trait & jb$ours$model == "fully adjusted", ]
  list(b = .ss1(r$est / jb$warming_af_per_decade), lo = .ss1(r$lo / jb$warming_af_per_decade), hi = .ss1(r$hi / jb$warming_af_per_decade)) }
af_m_degc <- .af("Body mass"); af_w_degc <- .af("Wing length")
af_w_pct <- local({ r <- jb$ours[jb$ours$trait == "Wing length" & jb$ours$model == "fully adjusted", ]
  list(b = .ss1(r$est), lo = .ss1(r$lo), hi = .ss1(r$hi)) })
# recovery simulation and contributor bootstrap (referee_recovery_sim.R)
rs <- readRDS("../Analysis/output/referee_reruns/recovery_sim.rds")
.rs <- function(scen, planted, model, col) { s <- rs$summary
  s[[col]][s$scenario == scen & abs(s$planted - planted) < 1e-9 & s$model == model] }
rs_reps <- rs$reps
rs_clean_m3 <- .ss(.rs("clean", -1, "M3", "mean_est"))
rs_art_m0 <- .ss(.rs("artefact", 0, "M0", "mean_est")); rs_art_m3 <- .ss(.rs("artefact", 0, "M3", "mean_est"))
rs_real_m3 <- .ss(.rs("real", 0, "M3", "mean_est"))
rs_m3_cov <- range(rs$summary$coverage[rs$summary$model == "M3"])
rs_m3_cov <- sprintf("%.0f–%.0f", 100 * rs_m3_cov[1], 100 * rs_m3_cov[2])
.pw <- function(planted) { x <- rs$summary$excl_zero[rs$summary$model == "M3" & abs(rs$summary$planted - planted) < 1e-9]
  sprintf("%.0f–%.0f", 100 * min(x), 100 * max(x)) }
rs_pow1 <- .pw(-1); rs_pow05 <- .pw(-0.5); rs_fp <- .pw(0)
# a 1 mm/decade change as a share of the mean wing length
rs_mde_pct <- sprintf("%.1f", 100 / readRDS("../Analysis/output/referee_reruns/logwing.rds")$mean_wing_mm)
att_diff <- sprintf("%s to %s", .ss(rs$attenuation$diff_ci95[1]), .ss(rs$attenuation$diff_ci95[2]))
att_boot_share <- sprintf("%.0f", 100 * mean(rs$bootstrap$M3 > rs$bootstrap$M0, na.rm = TRUE))
rs_boot_n <- rs$attenuation$n_boot
# log(wing length) ladder (referee_logwing.R)
lw <- readRDS("../Analysis/output/referee_reruns/logwing.rds")$summary
.lw <- function(m) { r <- lw[lw$model == m & lw$response == "log", ]; list(b = .ss1(r$per_decade), lo = .ss1(r$lo), hi = .ss1(r$hi)) }
lw_m0 <- .lw("M0"); lw_m3 <- .lw("M3")
# anomaly baselines (referee_anomaly_baseline.R)
ab <- readRDS("../Analysis/output/referee_reruns/anomaly_baseline.rds")
ab_cor_rec <- sprintf("%.2f", ab$baselines$cor_year[1]); ab_cor_full <- sprintf("%.2f", ab$baselines$cor_year[2])
.ab <- function(resp, model, term) { r <- ab$fits[ab$fits$response == resp & ab$fits$model == model & ab$fits$term == term, ]
  list(b = .ss4(r$estimate), lo = .ss4(r$lower), hi = .ss4(r$upper)) }
ab_iso_full <- .ab("isometry contrast lnM-3lnL", "dT_full", "dT_full")
ab_iso_fully <- .ab("isometry contrast lnM-3lnL", "dT_full + year", "dT_full")
ab_miss <- ab$missing_profile[ab$missing_profile$no_climate, ]
ab_have <- ab$missing_profile[!ab$missing_profile$no_climate, ]
ab_n_cov <- ab_have$records - ab$n          # shared records with climate but missing longitude, altitude or season
```

# Introduction

Climate warming can alter animal morphology through several pathways: direct thermoregulatory selection, changes in resource availability, and developmental or evolutionary responses to a shifting environment [@gardner_declining_2011; @sheridan_shrinking_2011; @teplitsky_climate_2014]. These mechanisms have generated prominent predictions, most familiarly that endotherms should become smaller in warmer environments, because a higher surface-area-to-volume ratio aids metabolic heat dissipation [Bergmann's rule; @bergmann1847; @mayr1956; @ashton2002], and that heat-dissipating extremities such as bills should enlarge [Allen's rule; @allen1877; @ryding2021shape]. Both rules were formulated to describe spatial patterns, across latitudes and climates; the expectation that populations should shrink, or that their appendages should lengthen, as a given locality warms is a temporal extension of them and a separate hypothesis [@gardner_declining_2011; @teplitsky_climate_2014]. Empirical support for such general responses is, however, mixed. Broad syntheses indicate that morphological responses to contemporary warming vary substantially among traits, species, and ecological contexts rather than following a single directional pattern. @radchuk2019adaptive, synthesising phenotypic responses across wild populations, recovered no consistent systematic morphological response to warming. Understanding this heterogeneity requires determining whether the patterns documented in well-studied systems extend to ecological and geographic settings that are underrepresented in long-term datasets.

Multi-decadal morphological analyses have predominantly drawn on temperate, often migratory, bird species sampled through museum collections or long-term ringing stations in the Northern Hemisphere [e.g., @weeks2020]. Tropical communities, where avian diversity peaks, are far less represented, so it remains unclear how general the reported patterns are. Against this background, @jirinec2021morphological provides a valuable tropical investigation: analysing four decades (1979–2019) of standardised field measurements from 77 non-migratory understorey species in the continuous primary rainforest of the central Amazon (Biological Dynamics of Forest Fragments Project, BDFFP), they documented widespread declines in body mass alongside wing elongation in roughly a third of species, and interpreted the resulting reduction in wing loading as an adaptive shift towards flight economy in an increasingly warm understorey. Whether that pronounced Amazonian response is representative of tropical forest birds more broadly, or is instead strongly context-dependent, is the question we address here.

The Brazilian Atlantic Forest offers an informative contrast. It is a global biodiversity hotspot that differs from the continuous lowland central Amazon in almost every respect that might matter: it spans steep topographic and latitudinal gradients, and it has lost over 80% of its original forest cover, leaving the remainder divided among thousands of small, isolated fragments [@ribeiro2009atlantic]. It is therefore a useful test of whether the Amazonian pattern generalises to a tropical forest system with a very different landscape and thermal structure.

Two additional axes may help characterise whatever response is present. First, warming may act on phenotypic variance as well as on trait means, either widening among-individual variation through developmental instability or relaxed selection, or narrowing it through directional canalisation [@teplitsky_climate_2014; @debat2001mapping; @badyaev2005stress]. Second, morphological change may be driven indirectly through food supply rather than directly through thermoregulation [@schmidt_concomitant_2005; @teplitsky_bergmanns_2008]. If so, species that rely on different foods need not respond in the same way, and diet, from predominantly frugivorous or granivorous to strictly insectivorous, may help explain why some species change and others do not.

Testing this requires long series, and in most tropical regions these exist only as compilations of records from many research programmes. Trait databases such as ATLANTIC BIRD TRAITS [@rodrigues_atlantic_2019] now assemble decades of field measurements across whole biomes, and mixed models can separate a temporal trend from the identity of the programme and site that supplied each record. This separation matters because programmes and sites rarely persist through a record. If those sampled late in a series differ from those sampled early, in measurement practice, protocol or habitat, a compilation can show a temporal trend that no population has undergone. How strongly such turnover can bias trends in compiled morphological data is seldom quantified.

Here, we use 24 years (1995–2018) of morphological data from 73 passerine species (`r .n(desc$n_rec)` live-capture records; ATLANTIC BIRD TRAITS; @fig-map) to ask whether Atlantic Forest passerines show consistent morphological change across a period of regional warming. We characterise that response in three focal traits—body mass, wing length, and bill width—and then ask how it is structured: whether mass and wing length change in concert on the same individuals or depart from geometric isometry, whether morphology covaries with interannual temperature anomalies, whether residual phenotypic variability has changed, and whether diet explains variation among species. Because the data were compiled from dozens of research programmes whose sites shifted over time, we model contributor and municipality explicitly, test by simulation whether these models recover known trends in the real sampling structure, and compare the resulting rates with those of the central Amazon on a common scale.

# Results

## The region warmed while contributing programmes and sites changed

The analytical dataset comprised `r desc$n_spp` passerine species and `r .n(desc$n_rec)` adult live-capture records (`r .n(desc$n_rec_wing)` wing records across 72 species; *Herpsilochmus pectoralis* has no wing measurements in the compilation and enters only the analyses of the other traits), gathered between 1995 and 2018 at `r clim$n_localities` coordinate localities spanning the latitudinal and elevational range of the Brazilian Atlantic Forest, from sea level to over 1,500 m (@fig-map). Filtering criteria are given in the Methods.

![Geographic distribution and multi-decadal sampling of `r desc$n_spp` passerine bird species across the Brazilian Atlantic Forest domain. Sampling localities are displayed separately for the early (1995–2006; $n = 2,370$ wing records, 93 coordinate sites, 12 contributors) and late (2013–2018; $n = 2,870$ wing records, 146 coordinate sites, 33 contributors) sampling periods across the biome. Points represent sampling localities coloured by mean wing length, illustrating the broad latitudinal and elevational range covered by the dataset.](images/fig-map.png){#fig-map}

Historical climate data, available for `r clim$n_localities_with_data` of these localities (the remainder fall outside the WorldClim land mask; see Methods), documented regional warming over the study period: mean annual temperature rose by `r clim_t$b`°C/decade (95% CI: `r clim_t$lo` to `r clim_t$hi`°C/decade), representing an average warming of `r clim_t_tot`°C across capture sites between 1995 and 2018. In contrast, precipitation (`r clim_p$b` mm/decade, 95% CI: `r clim_p$lo` to `r clim_p$hi` mm) and drought indices (SPEI-12: `r clim_s$b`, 95% CI: `r clim_s$lo` to `r clim_s$hi`) showed no clear directional trend.

These records were contributed by `r prov$n_src` research programmes working in `r prov$n_site` municipalities (`r prov$n_src_w` and `r prov$n_site_w` in the wing sample), and both the set of active programmes and the geographic focus of sampling shifted substantially over the 24 years (@fig-map; Supplementary Fig. S1).

## Wing length declined weakly once sampling turnover was modelled, and responses varied among species

Model-free comparisons of per-species mean trait values between the early (1995–2006) and late (2013–2018) sampling periods, the lower and upper quartiles of capture year, showed substantial variation across species (@fig-trends): wing length decreased in 43 species and increased in 18 (@fig-trends A), body mass decreased in 36 and increased in 26 (@fig-trends B), and bill width decreased in 13 and increased in 31 (@fig-trends C).

![Species-level trajectories of wing length, body mass, and bill width between early and late sampling periods. Paired per-species means between the early (1995–2006) and late (2013–2018) periods for (A) wing length ($n = 61$ species), (B) body mass ($n = 62$ species), and (C) bill width ($n = 44$ species) across Atlantic Forest passerines. Lines connect period means for each species, coloured by direction of shift (red = decrease, blue = increase), illustrating the raw interspecific variation across the assemblage.](images/fig-trends.png){#fig-trends}

Unadjusted models suggested a relatively steep decline in wing length over time (baseline model: `r b_w0$b` mm/decade, 95% CrI: `r b_w0$lo` to `r b_w0$hi`; `r b_w0_pct`%/decade; $P(\beta < 0) `r b_w0$P`$), but the estimated trend weakened substantially after accounting for contributor, municipality and the other sources of sampling turnover. In the fully adjusted model, the assemblage-level trajectory was modest and its credible interval overlapped zero: `r b_w3$b` mm/decade (95% CrI: `r b_w3$lo` to `r b_w3$hi`; `r b_w3_pct`%/decade; $P(\beta < 0) `r b_w3$P`$; @fig-wingtrend). The estimate was similar on complete cases (`r b_w3cc$b` mm/decade, 95% CrI: `r b_w3cc$lo` to `r b_w3cc$hi`) and for log wing length (`r lw_m3$b`%/decade, 95% CI: `r lw_m3$lo` to `r lw_m3$hi`%; REML check), and slightly steeper, with an interval excluding zero, after removing `r n_anom_blocks` anomalous sampling blocks found by a post hoc screen (`r cw_noanom$dec` mm/decade, 95% CI: `r cw_noanom$dec_lo` to `r cw_noanom$dec_hi`; Methods; Supplementary Tables S1 and S5). Letting species slopes covary phylogenetically left the estimate similar but its interval much wider (`r b_w3pcor$b` mm/decade, 95% CrI: `r b_w3pcor$lo` to `r b_w3pcor$hi`), and the phylogenetic share of slope variance was poorly constrained ($H^2 = `r h2s_w$b`$, 95% CrI: `r h2s_w$lo` to `r h2s_w$hi`), so the data do not tell us how far species' responses follow phylogeny.

Separating the year effect within and between contributors helped to locate some of the attenuation. Capture year varied substantially within contributors: `r src_cov$n_long` of the `r src_cov$n_src` contributors, holding `r .n(src_cov$rec_long)` wing records, sampled over `r src_cov$span_min` or more years, and `r src_cov$pct_within`% of the variance in capture year lay within contributors. The within-contributor slope was `r b_w5in$b` mm/decade (95% CrI: `r b_w5in$lo` to `r b_w5in$hi`; $P(\beta < 0) `r b_w5in$P`$); its interval includes both zero and the fully adjusted assemblage trend, so it cannot distinguish no change from a modest decline. Contributors whose records came, on average, from later years recorded shorter wings (`r b_w5btw$b` mm per decade of contributor mean year, 95% CrI: `r b_w5btw$lo` to `r b_w5btw$hi`; $P(\beta < 0) `r b_w5btw$P`$), a term estimated from only `r src_cov$n_src` contributor means. Municipalities offer less temporal replication: only `r site_cov$n_long` of `r site_cov$n_site` municipalities (`r .n(site_cov$rec_long)` wing records) were sampled over `r src_cov$span_min` or more years, and `r site_cov$pct_within`% of the variance in capture year lay within municipalities. The corresponding within-municipality slope (REML adjustment) was `r cw_m5site_within$dec` mm/decade (95% CI: `r cw_m5site_within$dec_lo` to `r cw_m5site_within$dec_hi`). The decompositions are therefore consistent with much of the steep unadjusted decline in wing length reflecting changes in sampling structure, but they do not establish how much temporal change remains within research programmes or sites.

To check that the adjusted model neither absorbs a real trend nor leaves an artefact in place, we planted known trends in `r rs_reps` simulated data sets per scenario that kept the real records, species, individuals, contributors, municipalities and years (Methods). With contributor and municipality effects independent of year, the fully adjusted model recovered a planted decline of −1.00 mm/decade almost exactly (mean estimate `r rs_clean_m3`). With contributor effects that drifted with contributors' mean capture year, calibrated to the gap between the real unadjusted and adjusted estimates, and no planted trend, the baseline model returned `r rs_art_m0` mm/decade and the fully adjusted model `r rs_art_m3`. Across scenarios, the 95% intervals of the fully adjusted model covered the planted value in `r rs_m3_cov`% of simulations. They excluded zero in `r rs_pow1`% of simulations with a planted decline of 1 mm/decade (about `r rs_mde_pct`% of wing length), in `r rs_pow05`% with 0.5 mm/decade, and in `r rs_fp`% with no trend (Supplementary Table S4). The data could therefore detect a wing decline of about 1 mm/decade with moderate power, but not one half that size.

![Modelling sampling turnover more than halves the estimated wing-length decline. (A) Within-species-centred wing length over time across `r .n(desc$n_rec_wing)` records from `r sc$n` species. The dashed line represents the baseline model (`r b_w0$b` mm/decade); the solid line and band represent the fully adjusted model (`r b_w3$b` mm/decade, 95% CrI spanning zero). Inset displays wing records per year. (B) Effects of each capture year on wing length from the baseline (open circles) and fully adjusted (filled circles) models with year as a factor, centred on the record-weighted mean year, with 95% confidence intervals (glmmTMB REML, pooled over 50 phylogenetic trees); the fitted linear trends of panel A are repeated. Unlike raw annual means, these intervals account for the clustering of records within species, contributors and municipalities. Linear trends and their bands are posterior means and 95% credible intervals pooled over 20 phylogenetic trees.](images/fig-wingtrend.png){#fig-wingtrend}

Species-specific slopes from the fully adjusted model showed pronounced interspecific heterogeneity (@fig-species-slopes): `r sc$neg_point` of `r sc$n` species with wing measurements had negative posterior means, but `r sc$cri_spans_zero` of `r sc$n` had 95% credible intervals overlapping zero. Only `r sc$cri_neg` species declined with intervals excluding zero, and `r sc$cri_pos` increased. The median species slope was `r sc$median` mm/decade. Overall, wing-length trajectories were weakly negative at the assemblage level but highly heterogeneous among species, with substantial uncertainty around the average trend.

![Interspecific heterogeneity in decadal wing-length trajectories under the fully adjusted model. Forest plot of species-specific slopes (mm/decade) across `r sc$n` passerine species, ordered by rate of change. Horizontal bars show 95% credible intervals; filled circles denote species whose intervals exclude zero (red = negative, blue = positive), while open circles denote species whose intervals span zero ($n = `r sc$cri_spans_zero`$ species). Shaded vertical band and solid blue line indicate the assemblage year slope (`r b_w3$b` mm/decade, 95% CrI: `r b_w3$lo` to `r b_w3$hi`). Hollow grey circles indicate baseline-model slopes.](images/fig-species-slopes.png){#fig-species-slopes}


## Body mass declined at about the Amazonian rate, but not consistently across species

Body mass showed no clear directional change across the 24-year record ($n = `r b_m3$n`$ records). The baseline estimate was modest (`r b_m0$b`%/decade, 95% CrI: `r b_m0$lo` to `r b_m0$hi`; $P(\beta < 0) `r b_m0$P`$). Under full adjustment it was `r b_m3$b`%/decade (95% CrI: `r b_m3$lo` to `r b_m3$hi`; $P(\beta < 0) `r b_m3$P`$), close in proportional terms to the fully adjusted wing-length trend (`r b_w3_pct`%/decade) but less precise, and within contributors it was `r b_mw$b`%/decade (95% CrI: `r b_mw$lo` to `r b_mw$hi`). Unlike wing length, the body-mass estimate changed little from the unadjusted to the fully adjusted model (Supplementary Table S2), and its interval still allows a decline of up to `r sub("^−", "", b_m3$lo)`% per decade.

On a common scale, this rate is close to the central Amazonian one. Across the 68 passerines of @jirinec2021morphological, mass changed by `r amz_m_dec`%/decade on average (species range `r amz_m_min` to `r amz_m_max`%; Methods), and per degree of local warming the two estimates are also similar (Atlantic Forest `r af_m_degc$b`% per °C, 95% CrI: `r af_m_degc$lo` to `r af_m_degc$hi`; central Amazon `r amz_m_degc`% per °C; @fig-amazon). What differs is consistency. In the Amazon all 77 species had lower mean mass by 2019, whereas in the Atlantic Forest 36 of 62 species decreased and 26 increased between the early and late periods (@fig-trends B). Wing length also differs: Amazonian wings lengthened by `r amz_w_dec`%/decade on average, whereas the fully adjusted Atlantic Forest estimate was `r af_w_pct$b`%/decade (95% CrI: `r af_w_pct$lo` to `r af_w_pct$hi`), with an upper bound that only just reaches the Amazonian mean.

![Rates of morphological change in the Atlantic Forest and the central Amazon on a common scale. **(A)** Change per decade in body mass and wing length. Atlantic Forest: posterior means and 95% credible intervals of the unadjusted (open circles) and fully adjusted (filled circles) models, pooled over `r bz$n_trees` phylogenetic trees; wing length is converted from mm to % with the effect-scale factor (Methods). Central Amazon: mean change of the 68 passerine species in Table S6 of @jirinec2021morphological (diamonds), with the range of species means (bars). **(B)** The same fully adjusted rates per °C of local warming (Atlantic Forest: `r clim_t$b`°C/decade, without propagating its uncertainty; central Amazon: `r amz_warm`°C/decade, the bar spanning the wet- and dry-season warming rates reported by that study).](images/fig-amazon.png){#fig-amazon}


## Mass tended to be retained relative to wing length

We next asked how the two traits changed together, using `r .n(n_iso)` shared capture events in which wing length and body mass were measured on the same individual, across `r n_iso_spp` species. Under strict geometric isometry mass scales with the cube of wing length, so we modelled the isometry contrast $\Delta_{\text{iso}} = \ln M - 3 \ln L$ (Methods): it stays constant if birds shrink or grow without changing shape, increases if mass is retained relative to wing length, and decreases if birds become lighter for their wing length. We call this last pattern slenderisation, the direction of change reported for the central Amazon.

On these shared records, under full adjustment, wing length changed by `r b_lnw$b`%/decade (95% CrI: `r b_lnw$lo` to `r b_lnw$hi`%) and body mass by `r b_lnm$b`%/decade (95% CrI: `r b_lnm$lo` to `r b_lnm$hi`%; @fig-isometry A). Strict isometry would predict a mass change of `r b_iso_exp`%/decade from the wing trend alone. Fitted directly, the isometry contrast changed by `r b_isof$b`%/decade (95% CrI: `r b_isof$lo` to `r b_isof$hi`%; $P(\beta > 0) `r b_isof$Ppos`$), and by `r b_isop$b`%/decade (95% CrI: `r b_isop$lo` to `r b_isop$hi`%; $P(\beta > 0) `r b_isop$Ppos`$) under the parsimonious adjustment set. The direct estimate differs slightly from the contrast implied by the separate wing and mass fits (`r b_iso_sep`%/decade), because each model estimates its own species slopes and random-effect variances; the direct fit is the test of isometry. Both estimates point towards mass being retained relative to the wing, but in both the credible interval includes zero, so the shared-individual data do not exclude strict geometric isometry. Because body mass changed little, the contrast mostly reflects the uncertain wing decline. Phylogenetically covarying species slopes gave a similar estimate with a wider interval (`r b_isofpcor$b`%/decade, 95% CrI: `r b_isofpcor$lo` to `r b_isofpcor$hi`%).

In bivariate morphospace (@fig-isometry B), the assemblage estimate lies close to stasis and above the isometry line, not in the quadrant of longer wings and lower mass that corresponds to the direction of change reported for the central Amazon [@jirinec2021morphological], although the credible intervals of both traits include zero. Only `r n_iso_decoupled` of `r n_iso_spp` species had posterior mean slopes in that quadrant.

![Bivariate temporal allometry and test of geometric isometry on shared individuals. **(A)** Decadal rates of morphological change (% per decade) on `r .n(n_iso)` shared records (`r n_iso_spp` species) with simultaneous wing length and body mass measurements, under full adjustment. Points and horizontal bars indicate posterior means and 95% credible intervals, pooled over `r bz$n_trees` phylogenetic trees, for wing length, body mass, the mass change expected under cubic geometric scaling ($M \propto L^3$), and the isometry contrast ($\Delta_{\text{iso}} = \ln M - 3\ln L$, fitted directly under the full and the parsimonious adjustment sets). **(B)** Bivariate morphological state space ($\Delta\text{Wing}$ vs. $\Delta\text{Mass}$). The solid black line indicates strict geometric isometry ($y = 3x$), the dotted line indicates equal rates ($y = x$), and the shaded quadrant (longer wings, lower mass) marks the direction of change reported for the central Amazon [@jirinec2021morphological], which is shown as a direction only (@fig-amazon compares the rates). The green point is the Atlantic Forest assemblage estimate with the marginal 95% credible intervals of wing and mass change; the two traits were fitted in separate models, so no joint credible region is shown. Semi-transparent circles are species posterior means ($n = `r n_iso_spp`$).](images/fig-isometry.png){#fig-isometry}

## Annual temperature anomalies were not associated with the wing–mass relationship

If the wing–mass shift tracks thermal conditions, individuals sampled in warmer-than-usual years should be heavier or lighter for their wing length than birds caught in cooler years. We tested this with the local annual temperature anomaly at each capture ($\Delta T_{\text{local}}$; Methods).

We found no clear evidence for it. Under the fully adjusted set used for the temporal models, the isometry contrast was not associated with the annual anomaly ($\beta = `r b_thi$b`$ per °C, 95% CrI: `r b_thi$lo` to `r b_thi$hi`; $P(\beta > 0) `r b_thi$Ppos`$), and neither were wing length ($\beta = `r b_thw$b`$ mm per °C, 95% CrI: `r b_thw$lo` to `r b_thw$hi`) or log body mass ($\beta = `r b_thm$b`$ per °C, 95% CrI: `r b_thm$lo` to `r b_thm$hi`). The anomaly effect on the contrast was positive in every specification, but its interval always included zero, including with calendar year in the model ($\beta = `r b_thiy$b`$, 95% CrI: `r b_thiy$lo` to `r b_thiy$hi`), with a detrended anomaly, with a locality-year random intercept and with a baseline over every year from 1995 to 2018 (Methods; Supplementary Tables S3 and S6).

Anomalies of the year before capture, or of the two preceding years, were not associated with the contrast either (lag 1: $\beta = `r lg1_i$b`$, 95% CI: `r lg1_i$lo` to `r lg1_i$hi`; lags 1–2: $\beta = `r lg12_i$b`$, 95% CI: `r lg12_i$lo` to `r lg12_i$hi`; REML, 50 trees), nor with wing length or body mass. We therefore make no claim that interannual thermal variation drives any change in the wing–mass relationship.

## An apparent increase in bill width disappeared once contributors and municipalities were modelled

Raw species-level trajectories for bill width displayed both increases and decreases across species (@fig-trends C). In a baseline model without contributor controls, bill width appeared to increase (`r b_bl0$b` mm/decade, 95% CrI: `r b_bl0$lo` to `r b_bl0$hi`; $P(\beta > 0) `r b_bl0$Ppos`$, $n = `r b_bl0$n`$), as Allen-type expectations would predict.

However, when contributor and municipality were accounted for, this apparent increase disappeared (`r b_bl1$b` mm/decade, 95% CrI: `r b_bl1$lo` to `r b_bl1$hi`). In the fully adjusted model, bill width remained flat: `r b_bl3$b` mm/decade (95% CrI: `r b_bl3$lo` to `r b_bl3$hi`; $n = `r b_bl3$n`$). Temporal change within the same contributor was similarly flat (`r b_blw$b` mm/decade, 95% CrI: `r b_blw$lo` to `r b_blw$hi`). The apparent baseline increase was therefore not robust to adjustment for contributor and spatial sampling structure, which were added together and whose contributions we cannot separate.

## An apparent rise in variability disappeared once contributors and municipalities were modelled

Heteroscedastic models estimated temporal change in the record-level residual standard deviation ($\sigma$) of wing length, that is, the variability remaining after the fixed and random effects of the mean model (Methods). In the baseline model, $\sigma$ appeared to increase by `r b_s1$b`%/decade (95% CrI: `r b_s1$lo` to `r b_s1$hi`%; $P(\gamma > 0) `r b_s1$Ppos`$). Adding contributor and municipality random intercepts to both the mean and the residual-SD models changed the sign of the estimate to `r b_s3$b`%/decade (95% CrI: `r b_s3$lo` to `r b_s3$hi`%), with the interval overlapping zero; with species intercepts also in the residual-SD model the estimate was positive again (`r var_s6$b`%/decade, 95% CI: `r var_s6$lo` to `r var_s6$hi`%; REML check). Adding local mean annual temperature to the residual-SD model yielded no association between residual variance and thermal conditions (`r b_s7t$b`% per SD of temperature, 95% CrI: `r b_s7t$lo` to `r b_s7t$hi`%).

Evidence for a change in residual variability in either direction therefore disappeared once contributor and municipality were accounted for, and residual variance was not clearly associated with local temperature.

## Diet did not clearly explain which species changed

To test whether diet attenuated or amplified wing-length trajectories, we used the share of invertebrates in each species' diet (EltonTraits), which runs from species feeding mostly on fruit and seeds to strict insectivores. In continuous interaction models (@fig-dietinteraction, top panel), the baseline model suggested a positive interaction between year and invertebrate share ($\beta = `r b_dA0$b`$ mm per SD of year per SD of diet, 95% CrI: `r b_dA0$lo` to `r b_dA0$hi`; $P(\beta > 0) `r b_dA0$Ppos`$). However, adding contributor and municipality intercepts halved this interaction to $\beta = `r b_dA1$b`$ (95% CrI: `r b_dA1$lo` to `r b_dA1$hi`; $P(\beta > 0) `r b_dA1$Ppos`$), with the credible interval spanning zero. Comparing invertebrate-dominated with other species (@fig-dietinteraction, bottom panel) gave a similarly uncertain interaction under the same controls ($\beta = `r b_dB1$b`$, 95% CrI: `r b_dB1$lo` to `r b_dB1$hi`).

The positive sign means that, if anything, species eating more fruit and seeds had steeper declines in wing length. Under the model with contributor and municipality intercepts, wing length changed by `r b_p10$b` mm/decade (95% CrI: `r b_p10$lo` to `r b_p10$hi`) at the 10th percentile of invertebrate share (`r b_p10$inv`% invertebrates), whereas the slope at the 90th percentile (`r b_p90$inv`%, strict insectivores) was centred near zero (`r b_p90$b` mm/decade, 95% CrI: `r b_p90$lo` to `r b_p90$hi`). Because the interaction itself is uncertain, diet did not clearly explain variation in wing-length trajectories, and we cannot exclude an effect of moderate size in either direction.

![Decadal wing-length change as a function of diet (share of invertebrates in the diet). Both panels show the models with contributor and municipality intercepts; lines are the predicted change in wing length relative to the mean sampling year, computed from the posterior mean of the year slope, and shaded ribbons are the 95% credible intervals of that slope (posterior pooled over `r bz$n_trees` phylogenetic trees), so they widen away from the mean year. Legend values give each slope in mm/decade with its 95% credible interval. (A) Continuous interaction model, evaluated at the 10th (p10, blue), 50th (p50, orange) and 90th (p90, red) species-weighted percentiles of invertebrate diet proportion. (B) Category model, for invertebrate-dominated and other species.](images/diet_interaction_plot.png){#fig-dietinteraction}


# Discussion

Across 73 passerine species and 24 years of regional warming in the Brazilian Atlantic Forest, morphological change was weak and varied among species. Once turnover in contributors and sites was modelled, the average body-mass decline was close to the mean rate reported from the central Amazon, but its interval included zero and species did not decline together; wings did not lengthen, and bill width did not change. Much of the change suggested by unadjusted models reflected how the compilation was assembled, and simulations showed that modelling this structure recovers known trends rather than absorbing them.

## Interspecific variation and the wing–mass relationship

The heterogeneity among species (@fig-species-slopes) has a practical implication: an assemblage-level trend near zero can conceal species changing in opposite directions, so community averages are a weak basis for judging whether, or how, a regional avifauna is responding to warming. Large multi-species analyses elsewhere likewise report species-specific responses, even where the average trend is clear [@teplitsky_climate_2014; @youngflesh2022abiotic; @dubiner2022widespread]. Identifying which species change, and what they have in common, is therefore more informative than refining the assemblage mean, and will require longer species-level series than most contributors to this compilation provide. Diet was one candidate. We did not assume that insectivores should be the most sensitive: frugivory and granivory are specialisations too, and fruit and seed supply can also depend on climate. In our data, diet did not clearly separate species that changed from those that did not, and if anything species feeding mostly on fruit and seeds declined in wing length more than strict insectivores. Species-level diet shares are coarse, however, and do not capture seasonal or local shifts in what birds eat, so a role for diet in structuring responses remains open.

The wing–mass result bears on what kind of change occurred. Had birds simply become smaller, mass would have fallen roughly three times faster than wing length; instead, mass tended to be retained relative to the wing, although strict isometry cannot be excluded. Even a clear departure in the contrast would reject only proportional co-variation of the two traits; it would not on its own establish a change in whole-body shape. This does not point to a general reduction in structural size and, together with the uncertain change in mass, gives no sign of the slenderisation documented in the Amazon [@jirinec2021morphological]. One reason the two traits can change independently is that they measure different things. Wing length in live passerines is a measurement of the primary feathers [@jenni1989featherlength], which are renewed at each moult and whose growth responds to nutritional condition [@grubb2006reduced], and wing shape can change over decades as landscapes alter the demands placed on mobility [@desrochers2010morphological]. Body mass, by contrast, combines structural size with short-term condition. In a 70-year museum series, @dubiner2022widespread similarly found that most species changed in either mass or body length but seldom in both, whereas changes in wing length were few and inconsistent. Modest wing change without proportional mass change is therefore as compatible with shifts in feather growth, wear, or the condition of birds at capture as with a change in body size.

## Contrasting macroecological dynamics: the Atlantic Forest and the central Amazon

On a common scale, the Atlantic Forest and the central Amazon differ less in the average rate of mass change than in its consistency and in wing length (@fig-amazon). In the central Amazonian study of @jirinec2021morphological, the most directly comparable multi-decadal Neotropical benchmark, every species lost mass on average and wings lengthened, so wing loading fell across the understorey community. In the Atlantic Forest, species diverged in both directions around a similar mean mass change, and wings did not lengthen. Our interval for body mass is wide, so the data cannot say whether the two regions differ in their average rate. Other tropical studies are closer to our result: in Panama, wing length changed weakly and in both directions [@wilcox2024observed], century-scale trends in lowland Colombia were mixed [@perezamaya2025century], and Afromontane understorey birds became larger [@neateclegg2024afromontane].

The Amazonian response is therefore one of several tropical outcomes. Ours is a contrast between two studies, not a controlled comparison: the two datasets differ in landscape, sampling design, temporal coverage, species composition and measurement structure, and part of the difference may lie in these features. Three ecological hypotheses, which our data cannot discriminate, could also contribute. First, the Amazonian study took place in continuous forest, where longer wings and lower wing loading reduce the cost of flight [@norberg1990vertebrate; @jirinec2021morphological], whereas the Atlantic Forest survives as small fragments [@ribeiro2009atlantic]. Wing morphology predicts movement across such landscapes [@sheard2020ecological; @claramunt2022effect], so movement between fragments could favour different changes in wings and mass. Second, the Atlantic Forest spans steep gradients from sea level to over 2,000 m, where birds can track warming upslope [@freeman2018climate] or in buffered microhabitats [@defrenne2019global; @scheffers2014microhabitats], which would weaken thermoregulatory pressure on body size. Third, birds in fragmented forest already experience wide, edge-driven thermal variation [@murcia1995edge; @ewers2013fragmentation] and may tolerate warming without directional change. We did not measure wing shape, microclimate or edge exposure, so these hypotheses remain untested.

The contrast may also reflect study design: a single continuous 40-year ringing programme (BDFFP) versus a multi-site compilation assembled from many programmes. Our within-contributor decomposition does not rule this out. The within-contributor slope, which uses only temporal variation inside each research programme, was centred near zero for wing length (`r b_w5in$b` mm/decade; complete cases: `r b_w5ccin$b` mm/decade) and body mass (`r b_mw$b`%/decade), but it was estimated imprecisely, and its interval for wing length also included the fully adjusted assemblage trend. Moreover, restricting the comparison to within a programme removes confounding from turnover among research programmes, but not change over time in the field workers, localities, or protocols operating within a programme. A sampling explanation for part of the contrast therefore remains open.

## Evaluating ecogeographical rules in tropical endotherms

Our findings bear on the temporal extension of classic ecogeographical rules in tropical endotherms under climate change. Bergmann's and Allen's rules describe spatial patterns; the expectation that warming through time should drive reductions in body size, generalised from Bergmann's rule [@gardner_declining_2011; @sheridan_shrinking_2011], is a separate hypothesis, and a temporal null result bears on that hypothesis, not on the spatial rules themselves. In our 24-year dataset of 73 tropical passerine species, the average body-mass decline (`r b_m3$b`%/decade) was similar to the Amazonian rate, but its interval spanned zero despite +`r clim_t_dec2`°C/decade of regional warming. The data therefore neither confirm nor exclude a warming-driven decline of the size reported from the Amazon, although they make a decline steeper than about `r sprintf("%.0f", abs(as.numeric(sub("−", "-", b_m3$lo))))`% per decade unlikely.

Several considerations explain why the decline hypothesis is a weak prediction over this time span, even under measurable warming. First, the warming recorded at our capture localities, about `r sub("^\\+", "", clim_t_tot)`°C over the study period, is small relative to the thermal variation that the same species experience across their elevational and latitudinal ranges and between forest microhabitats [@defrenne2019global; @scheffers2014microhabitats], so any thermoregulatory advantage of a smaller body over this period would be slight. Second, Bergmann's rule is itself far from universal: it holds in a majority, but not all, of the bird and mammal taxa examined [@meiri2003validity], and reported recent changes in vertebrate body size are inconsistent in direction and prone to methodological artefacts [@yomtov2011recent]. Where temporal declines in avian size have been detected, they come from very large samples and vary among species [@youngflesh2022abiotic; @dubiner2022widespread]. Third, body size depends on food supply and conditions during development as well as on temperature [@teplitsky_bergmanns_2008; @teplitsky_climate_2014], and these influences need not act in the same direction as warming. Finally, a heritable response within 24 years would require strong and consistent selection, and evidence for genetic adaptation to climate change remains scarce relative to plastic responses [@merila2014climate]. These considerations do not turn the absence of a decline into a confirmed prediction. They explain why a 24-year record may not resolve a decline of this size, and our result bears on the temporal decline hypothesis, not on whether body size has a thermoregulatory role.

Similarly, Allen's rule posits that appendages enlarge in warmer environments to facilitate heat dissipation [@allen1877; @ryding2021shape]. While unadjusted models suggested a temporal increase in bill width, this apparent trend did not survive adjustment for contributor and municipality, and bill width showed no clear change. In tropical passerines, bill morphology is heavily constrained by foraging biomechanics, trophic specialisation, and vocal performance, which may override subtle thermoregulatory selection for larger bills [@podos2001correlated; @tattersall2017evolution].

## Sampling structure in compiled trait databases

Our study highlights the importance of modelling provenance and sampling structure when analysing multi-decadal compilations in macroecology. Large-scale compilations such as ATLANTIC BIRD TRAITS [@rodrigues_atlantic_2019] are indispensable for macroecological and macroevolutionary research, providing taxonomic and geographic scope that single-site studies cannot achieve. However, because compiled databases combine measurements across dozens of independent research teams, sampling localities, protocols, and seasons, apparent multi-decadal trends can be substantially confounded by sampling turnover.

In our dataset, unadjusted models suggested steep declines in wing length, increases in bill width, and expanding trait variability. Once sampling structure was modelled, the estimated wing decline fell from `r b_w0$b` to `r b_w3$b` mm/decade, the apparent bill expansion disappeared, and the apparent variance inflation was no longer supported. The size of the wing attenuation is itself uncertain: resampling contributors, the difference between the adjusted and unadjusted slopes ranged from `r att_diff` mm/decade (95% interval), although the adjusted slope was the shallower one in `r att_boot_share`% of `r rs_boot_n` resamples. Because contributor identity, sampling geography, protocol, and season are not fully separable in a compilation of this kind, we do not attribute this attenuation to any single source; what the comparison establishes is that the unadjusted estimates were substantially influenced by how the dataset was assembled. Simulations in the real sampling structure show that the adjusted model recovers planted trends without absorbing them, and that the baseline model is biased when contributor effects drift with time. They cannot rule out turnover that the modelled grouping factors and covariates do not capture, so a residual sampling explanation remains possible. The main inference is therefore methodological: when trends are estimated from compiled trait databases, contributor, municipality and the other features of sampling need to be modelled explicitly, and unadjusted and adjusted estimates reported side by side. Trait compilations can therefore support macroecological inference, provided that sampling structure is modelled explicitly and its influence reported.

## Conclusion

In summary, 24 years of warming in the Brazilian Atlantic Forest were accompanied by an average body-mass decline similar in rate to that of the central Amazon, but not by its consistency across species or by wing elongation. Once sampling structure was modelled, wing length and body mass both showed small declines (`r b_w3_pct`% and `r b_m3$b`% per decade) whose credible intervals included zero, bill width did not change, and body mass tended to be retained relative to wing length, although strict geometric isometry could not be excluded. For body mass, the interval (`r b_m3$lo` to `r b_m3$hi`% per decade) makes a sustained decline steeper than about `r sprintf("%.0f", abs(as.numeric(sub("−", "-", b_m3$lo))))`% per decade unlikely, and our simulations show that a wing decline of about 1 mm per decade would usually have been detected. The difference from the Amazon raises the possibility that landscape fragmentation, topographic complexity and ecological context shape how consistently tropical birds respond to warming, a hypothesis that our study motivates but cannot itself test, and it may also reflect differences in study design. More generally, compiled trait databases can extend long-term morphological monitoring to under-studied tropical regions, but only when turnover in contributors and sites is modelled: ignoring it here would have more than doubled the estimated wing decline.

# Materials and Methods {#sec-methods}

## Trait database filtering and analytical sample derivation

We extracted avian morphological measurements from the ATLANTIC BIRD TRAITS database [@rodrigues_atlantic_2019], a comprehensive data compilation comprising `r .n(desc$raw_total)` morphometric entries across the Brazilian Atlantic Forest domain, of which passerines represent `r desc$raw_pass_pct`% (`r .n(desc$raw_passerine)` records). To ensure that temporal analyses reflect genuine living phenotypic variation rather than methodological or preservation artefacts, we implemented a multi-stage filtering protocol.

First, we restricted the dataset strictly to live field captures, excluding museum study skins. Museum preparation induces post-mortem shrinkage in feathers, bill keratin, and skeletal dimensions (typically 1–3% reduction in linear dimensions upon drying; @weeks2020). Because museum specimens in regional collections are heavily concentrated in earlier decades of the 20th century, retaining study skins introduces artificial positive temporal trends. Second, we restricted records to adult individuals (`Age == "Adult"`), excluding juveniles and fledglings whose flight feathers and bills are incompletely developed. Third, we filtered records spatially to sampling localities falling within a 20-km buffer of the Brazilian Atlantic Forest biome boundary (IBGE/MMA official shapefile), eliminating captures situated in adjacent Cerrado or Caatinga enclaves. The 20-km allowance is the tolerance already applied in the ATLANTIC BIRD TRAITS compilation itself, and it accommodates coordinate imprecision and the fact that the mapped biome boundary is a generalisation of a gradual ecotone; it is not a biological buffer and no result depends on its exact width. Fourth, we bounded the temporal window to 1995–2018 (24 calendar years), representing the era of active multi-project mist-netting in the biome with consistent capture documentation. We did not filter on migratory status: the compilation does not record it, and we therefore make no claim that the focal species are strictly resident. The sample comprises the passerines captured by Atlantic Forest mist-netting programmes, which may include partial or altitudinal migrants.

Finally, to guarantee robust parameter estimation for species-level random intercepts and random slopes, we restricted the primary analytical sample to species with $\ge 30$ live-capture records and a temporal sampling span of $\ge 5$ years. This yielded an analytical dataset of `r desc$n_spp` passerine species and `r .n(desc$n_rec)` live-capture records (`r .n(desc$n_rec_wing)` wing-length records across 72 species; one species, *Herpsilochmus pectoralis*, was sampled exclusively for bill and tarsus traits and lacked wing and mass records in the compilation). Wing length was coalesced across database columns, prioritising right wing length when specified, followed by left wing length and unspecified wing length (`conc.wing.length`).

## Accounting for contributor turnover and spatial provenance

In large-scale macroecological compilations aggregating data across independent research programmes, two structural factors can confound temporal trends if left unmodelled: contributor turnover and spatial turnover.

### Contributor turnover and research-group random effects

In avian morphometrics, differences between observers can be of the same order as the temporal changes of interest here [@yezerinac1992measurement]. If different field workers apply slightly different caliper tension or wing-flattening techniques, and if the set of active research programmes changes over time, apparent temporal trends in morphology can arise entirely through turnover in who contributed the data.

In the ATLANTIC BIRD TRAITS compilation, the field measurer column (`Measurer`) is 100% missing (NA), so the individual who took each measurement cannot be identified. The identity of the principal investigator and research group coordinator (`Main_researcher`) is, however, 100% complete across all analytical records, comprising `r prov$n_src_w` distinct entries in our wing dataset. **Throughout this paper we refer to this grouping variable as the *contributor*, meaning the research group or project that supplied the record, and we use this single term consistently in place of "observer" or "measurer".** We emphasise that the contributor is a proxy rather than a direct measure of measurement practice: a contributor random effect absorbs whatever is shared among records originating from the same research programme, which includes caliper and wing-flattening conventions, but also field protocols, target habitats, the set of localities worked, and project-specific species composition. Attenuation of a temporal trend after adding this term therefore indicates that the trend was partly carried by differences among research programmes, and does not by itself isolate measurement error from these other, partly inseparable, sources.

An audit revealed high contributor turnover across the 24-year record: only `r meth$src_both` of the `r prov$n_src_w` contributors sampled birds in both the early (1995–2006) and late (2013–2018) periods (the lower and upper quartiles of capture year; the middle half of records, 2007–2012, is left out of these descriptive comparisons, and all models use every year), while the late period experienced an influx of `r meth$src_la` active contributors compared to `r meth$src_ea` early (Supplementary Fig. S1). To account for baseline differences among research programmes, we incorporated a random intercept for contributor identity (`(1 | Main_researcher)`).

### Spatial provenance and continuous environmental controls

Over the 24-year timeframe, geographical sampling shifted substantially: early captures (1995–2006) were heavily concentrated in the southern Atlantic Forest (mean latitude `r meth$lat_ea`°), whereas later captures (2013–2018) expanded northwards (mean latitude `r meth$lat_la`°) and into higher elevations (median altitude shifting from `r meth$alt_ea` m to `r meth$alt_la` m). Because widespread passerines frequently exhibit ecogeographical clines in body size, geographic turnover across sampling periods can easily mimic temporal phenotypic change.

To account for continuous spatial and elevational variation without discarding records or introducing artificial clustering, models incorporated standardised latitude (`scaled_lat`), standardised longitude (`scaled_long`), and standardised elevation (`scaled_elev`) as continuous covariates.

### Sampling hierarchy: contributor and municipality random effects

Another structural consideration in multi-source macroecological databases is the relationship between research teams and geographic sampling units. Across the 24-year ATLANTIC BIRD TRAITS compilation, individual field projects were typically conducted by distinct research groups operating within specific municipal territories, so the two grouping factors are correlated: in our shared analytical sample ($n = `r .n(meth$n_shared)`$), `r meth$one_sh$single` of `r meth$one_sh$n` municipalities (`r meth$one_sh$pct`%) were sampled by only a single contributor (and across the full body mass dataset, `r meth$one_m$single` of `r meth$one_m$n` municipalities, `r meth$one_m$pct`%, were sampled by a single contributor).

The two factors nonetheless capture distinct sources of variation. The contributor term absorbs what is shared among records from one research programme, including measurement conventions (caliper tension, wing flattening, scale calibration) and field protocols. The municipality term absorbs local differences among the sampled sites and populations, such as habitat, landscape context, and local population composition, that continuous latitude, longitude, and elevation do not capture. The two are partially crossed rather than nested: in the wing dataset, `r meth$site_multi`% of municipalities were sampled by more than one contributor, and `r meth$src_multi`% of contributors worked in more than one municipality, which allows their variances to be estimated separately. Both variances were clearly non-zero in the fully adjusted wing model (among-contributor SD = `r cw_sd_src` mm; among-municipality SD = `r cw_sd_site` mm).

Including municipality also matters for the temporal inference. Because sampling expanded northwards and upslope over the study period, municipality identity is strongly associated with capture year (municipality alone accounts for `r meth$r2_site`% of the variance in capture year in the wing dataset, against `r meth$r2_src`% for contributor). Without a municipality term, differences among sites sampled early and late in the series can be read as temporal change. We therefore included random intercepts for both contributor (`(1 | Main_researcher)`) and municipality (`(1 | Municipality)`) in every model beyond the baseline, including the thermal-anomaly and isometry models.

## Model sequence and within-between decomposition

To evaluate how dataset structure influences temporal trends, we fitted a ladder of increasingly controlled mixed-effects models. We refer to each model by a descriptive name; the short codes used in the analysis scripts and in Supplementary Tables S1 and S2 are given in parentheses.

- **Baseline** (M0): replicating published unadjusted model structures, with sex, calendar year (`scaled_yr`, centred and divided by its standard deviation, SD = `r sprintf("%.2f", cw_phylo$sd_year)` years), latitude (`scaled_lat`), species-specific random intercepts and year slopes, and phylogenetic covariance among the species intercepts.
- **+ contributor & municipality** (M1): adding random intercepts for contributor identity `(1 | Main_researcher)`, to absorb systematic differences among research programmes, including but not limited to measurement conventions, and for municipality `(1 | Municipality)`, to absorb local differences among sampled sites.
- **+ wing side, longitude & elevation** (M2): adding fixed effects for the protocol proxy (`wing_col`: right wing, left wing, or side unspecified; Supplementary Fig. S2), standardised longitude (`scaled_long`), and standardised elevation (`scaled_elev`).
- **Fully adjusted** (M3): adding fixed effects for seasonal quarter (DJF: Dec–Feb; MAM: Mar–May; JJA: Jun–Aug; SON: Sep–Nov) to account for feather wear across the annual cycle, active primary moult status (`Molt`), and a random intercept for the individual bird `(1 | ID_individual)` (Ring $\times$ Species) to account for repeated captures of banded individuals.
- **Year within vs between contributors** (M5src; Mundlak within-between decomposition): to separate temporal change occurring within a single contributor from compositional turnover across contributors, calendar year was partitioned into the contributor mean capture year ($\bar{Y}_j = \text{yr\_src\_mean}$) and individual capture deviations from that contributor mean ($Y_{ij} - \bar{Y}_j = \text{yr\_within\_src}$) [@mundlak1978pooling; @bell2019fixed], on the fully adjusted specification. The within-contributor coefficient $\beta_{\text{within}}$ measures whether birds contributed by the *same* research programme changed over time, and is therefore not driven by turnover in which programmes were active. It does not, however, remove change over time in the individual field workers, localities, or protocols operating within a programme.
- **Robustness and sensitivity models**:
  - **Fully adjusted, complete cases** (M3_cc): the fully adjusted specification on complete cases ($n = `r cw_m3_cc$n`$). The primary sample ($n = `r cw_m3$n`$) does not discard records with incomplete covariates; instead, missingness is encoded explicitly. Records lacking a recorded altitude take a value imputed from other records at the same coordinates, and season and moult carry an explicit `unrecorded` level rather than being dropped. The complete-case model removes the `r cw_ss$altitude_na_wing + cw_ss$season_na_wing` records with an imputed altitude or an unrecorded season, so the contrast between the two isolates the effect of that encoding. Unrecorded moult is kept as an explicit level in both models: it applies to `r molt_unrec_pct`% of wing records, and `r molt_src_none` of the `r src_cov$n_src` contributors never recorded it, so dropping these records would remove whole contributors and confound the complete-case contrast with contributor composition.
  - **Fully adjusted, first captures only** (M7): restricted to first captures ($n = `r cw_m7$n`$) to exclude any potential bias from recapture histories.
  - **Fully adjusted, anomalous blocks removed** (M3_noanom): re-fitted by REML after excluding `r n_anom_blocks` anomalous sampling blocks (`r n_anom_rec` records). Blocks are contributor $\times$ municipality $\times$ year cells. For each record we took the deviation of its wing length from its own species $\times$ sex mean, averaged those deviations within each cell, and flagged cells with at least 10 records whose mean deviation exceeded 6 mm in absolute value. We set the 6 mm threshold after inspecting the data, so this screen is a post hoc sensitivity check and not part of the primary analysis; the threshold is roughly 1.5 residual standard deviations of the fully adjusted fit and about 8% of a typical wing; a shift of that size shared across an entire contributor-site-year block is more consistent with a protocol or data-entry change than with phenotypic variation. The screen is sign-blind: of the flagged blocks, `r sum(cw_ss$anomalous_blocks$mean_dev < 0)` lie below and `r sum(cw_ss$anomalous_blocks$mean_dev > 0)` above species means.
  - **Year within vs between municipalities** (M5; REML): the same decomposition as M5src with municipality in place of contributor, to separate change within sampled sites from differences between sites sampled at different times.

## Bayesian phylogenetic mixed models across 20 topologies

Taxonomic reconciliation between the ATLANTIC BIRD TRAITS database and the complete, dynamic avian phylogeny of @mctavish2025tree was executed using the four-stage matching cascade of `prepR4pcm` [@nakagawa2026prepr4pcm]: exact binomial matching, punctuation/spacing normalisation, taxonomic synonym resolution via BirdLife International and IOC checklists, and fuzzy string matching. All 73 passerine species were audited and successfully mapped to phylogenetic tip labels with zero unplaced taxa.

To accommodate phylogenetic uncertainty, we drew 20 alternative phylogenetic topologies from the posterior distribution of @mctavish2025tree. For each tree, we computed the phylogenetic correlation matrix $A$ under a Brownian motion model of trait evolution.

All models were Gaussian linear mixed models fitted in a Bayesian framework with a purpose-written Stan program [@carpenter2017stan]. For the fully adjusted wing model, the length $y_i$ of record $i$ is
$$y_i = \mathbf{x}_i^\top \boldsymbol{\beta} + a_{s[i]} + b_{s[i]}\, t_i + c_{k[i]} + m_{j[i]} + u_{r[i]} + \varepsilon_i, \qquad \varepsilon_i \sim \mathcal{N}(0, \sigma^2),$$
where $\mathbf{x}_i$ holds the fixed effects (sex, standardised year $t_i$, latitude, longitude, elevation, season, moult and wing side), $a_s$ and $b_s$ are the intercept and year-slope deviations of species $s$, and $c_k$, $m_j$ and $u_r$ are normally distributed random intercepts for contributor $k$, municipality $j$ and individual bird $r$, each with its own standard deviation. The other models differ only in the terms they include. The species intercepts have a phylogenetic and a non-phylogenetic component,
$$\mathbf{a} = \sigma_a \left( \sqrt{H^2}\, \mathbf{L} \mathbf{z}_{\text{phy}} + \sqrt{1 - H^2}\, \mathbf{z}_{\text{sp}} \right), \qquad \mathbf{L}\mathbf{L}^\top = A,$$
with $\mathbf{z}_{\text{phy}}, \mathbf{z}_{\text{sp}}$ standard normal and $\mathbf{L}$ the spectral square root of $A$, so that $\mathbf{a}$ has covariance $\sigma_a^2 [H^2 A + (1 - H^2) I]$ and $H^2$ is the phylogenetic share of among-species variance in the intercepts. Species year slopes $b_s$ are modelled as independent among species rather than as a Brownian process. The individual intercepts are integrated out of the likelihood analytically: the records of one bird are jointly normal with covariance $\sigma^2 I + \tau^2 \mathbf{1}\mathbf{1}^\top$, whose inverse and determinant have closed forms, so the thousands of individual effects are never sampled. Together with sampling the phylogenetic effects in the eigenbasis of $A$ (adapting marginalised spectral code by S. Ortega), this makes fitting across many trees feasible.

The response was centred and scaled before fitting and all estimates were back-transformed to the original scale. Priors were weakly informative on that standardised scale: $\mathcal{N}(0, 5)$ for fixed effects, half-Student-$t(3, 0, 1)$ for all standard deviations, $\text{Beta}(1, 1)$ for $H^2$, and $\mathcal{N}(0, 1)$ for the coefficients of the residual-SD model (below). Each model was sampled separately on each tree with CmdStan 2.39.0 through `cmdstanr` 0.9.0: four chains of 1,000 warm-up and 500 sampling iterations (2,000 warm-up and 2,000 sampling iterations for bill width), with `adapt_delta` = 0.95 and a maximum tree depth of 12. The posteriors from the 20 trees were combined by pooling their draws, which gives the mixture posterior over trees and so carries phylogenetic uncertainty into every estimate [@nakagawa2019rubin]. No sampler divergences occurred in any of the `r bz$n_fits` models fitted. For the parameters quoted in the text (year, temperature-anomaly and interaction terms, and the phylogenetic share and correlation of the species slopes), the largest rank-normalised $\hat{R}$ on any single tree was `r bz$foc_rhat` (`r if (rhat_top$n_gt_101 == 0) "none" else rhat_top$n_gt_101` of the `r .n(rhat_top$n_terms)` parameter-by-tree combinations exceeded 1.01). Trees on which any of these parameters initially exceeded 1.01 were refitted with 2,000 warm-up and 2,000 sampling iterations per chain and thinned by four, so that each keeps the same weight in the tree mixture; the replaced fits and the list of refitted trees are archived. The smallest per-tree bulk and tail effective sample sizes were `r rhat_top$ess_bulk_min` and `r rhat_top$ess_tail_min`, and summed over the `r bz$n_trees` trees they were at least `r bz$foc_ess` and `r bz$foc_ess_tail`, respectively [@vehtari2021rhat]. The among-species standard deviation of the intercepts mixed more slowly (largest per-tree $\hat{R}$ `r bz$sd_rhat`; smallest per-tree bulk effective sample size `r bz$sd_ess`); it is a nuisance parameter here and none of our conclusions rests on it.

We report posterior means with 95% equal-tailed credible intervals (CrI) and, for temporal effects, the posterior probability that the effect is negative, $P(\beta < 0)$: the share of posterior draws below zero, which is a direct statement of how probable a decline is given the model and data. Values near 0.5 indicate no evidence for either direction.

As an independent check on the estimation engine, every model except the detrended-anomaly and locality-year checks (see *Interannual temperature anomalies*) was also fitted by restricted maximum likelihood with the `propto` covariance structure of `glmmTMB` v1.1.14 [@williams2025glmmtmb] and pooled across 50 trees with Rubin's rules [@nakagawa2019rubin; @nakagawa2026prepr4pcm]. For the temporal coefficients reported here, posterior means and REML estimates differed by at most `r bz$shift_max` posterior standard deviations, and posterior SDs were 1.00–`r bz$sd_ratio_max` times the REML standard errors, as expected when the uncertainty in the variance components is propagated. The complete REML control ladder is given in Supplementary Tables S1 and S2. As a sensitivity analysis, we refitted the fully adjusted wing model and the isometry model with Brownian species slopes correlated with the species intercepts, a structure that `propto` cannot represent. For the fully adjusted wing model we also compared, by REML likelihood ratio on each tree, the model with independent species slopes against one that adds a phylogenetic slope component (boundary-corrected $p$). The larger model failed to converge on `r lrt_w$n_fail` of the 50 trees; on the remaining `r lrt_w$n`, `r if (lrt_w$sig == 0) "none" else lrt_w$sig` gave $p < 0.05$ (median $p = `r lrt_w$p_med`$).

## Body mass and bill width modelling and testing Allen's rule

To evaluate whether temporal shifts occurred in body mass and extremity dimensions, we applied the sequential control ladder to log-transformed body mass ($n = `r b_m3$n`$) and bill width ($n = `r b_bl3$n`$). Bill width is the appendage measure used here because it is the extremity trait with the broadest coverage in the compilation and because bill surface is a major avian site of non-evaporative heat exchange, which is the dimension Allen-type predictions concern. It is, however, a single and imperfect proxy: tarsus length is recorded for fewer individuals, and we did not analyse bill depth or surface area, so our test addresses one extremity dimension rather than heat-dissipating surface as a whole. The fully adjusted models incorporated sex, latitude, longitude, elevation, season, species-specific random intercepts (with phylogenetic covariance) and year slopes, and random intercepts for contributor, municipality and individual bird identity; moult is not recorded for enough of these records to be included.


To evaluate Allen's rule [@allen1877; @ryding2021shape], which predicts that extremities enlarge in warmer conditions to facilitate non-evaporative heat dissipation, we tested whether bill width exhibited a positive temporal slope under contributor and spatial controls.

## Direct test of geometric isometry

To evaluate whether body shape changed over time, we analysed records with simultaneous measurements of wing length and body mass on the same individual bird. Under geometric isometry, body mass scales with the cube of linear dimensions ($M \propto L^3$), so a bird shrinking uniformly without changing shape loses mass at about three times the proportional rate of its linear shrinkage ($\% \Delta M \approx 3 \times \% \Delta L$).

We tested departure from isometry by fitting the log-scale contrast $\Delta_{\text{iso}} = \ln M - 3 \ln L$ directly as the response. Its year slope is zero under strict isometry whatever the separate trends of the two traits, and fitting it directly yields its posterior without depending on the covariance between two separately estimated slopes. A positive slope indicates that body mass declined less than cubic scaling of the wing predicts (mass conservation or stoutening); a negative slope indicates disproportionate mass loss or wing elongation (slenderisation). The contrast was fitted under two adjustment sets: a parsimonious set (sex, year, latitude, species intercepts with phylogenetic covariance and species year slopes, and random intercepts for contributor and municipality), and the fully adjusted set, which adds longitude, elevation, season, moult and individual identity. To show the two sides of the contrast, $\ln L$ and $\ln M$ were also fitted separately under the fully adjusted set. Year slopes on log scales are expressed as percentage change per decade, $100\,[\exp(10\beta/\text{SD}_{\text{year}}) - 1]$. These models use the `r .n(n_iso)` of the `r .n(rr_n_shared)` shared records with longitude, elevation and season recorded. They do not require the local temperature anomaly, which is available for a subset of `r .n(rr_n_therm)` records (next section); on that subset the isometry contrast gave a similar estimate (`r b_isof_clim$b`%/decade, 95% CrI: `r b_isof_clim$lo` to `r b_isof_clim$hi`%).

## Interannual temperature anomalies

Each capture record was linked to the mean annual temperature at its coordinates in its capture year, taken from the WorldClim historical monthly weather series [monthly tmin/tmax, 1950–2024, CRU-TS 4.09 downscaled with WorldClim 2.1, 2.5 arc-min; @harris2020crudata; @fick2017worldclim]. This is the time-resolved product, not the WorldClim climate normals. The local annual anomaly ($\Delta T_{\text{local}}$) is that value minus the mean of the same quantity over all records at the locality, that is, over the years in which the locality was sampled, weighted by captures. This baseline removes most of the regional warming trend from the anomaly (correlation with capture year `r ab_cor_rec`), whereas a baseline over every year from 1995 to 2018 leaves it in (correlation `r ab_cor_full`). With the full-series baseline the isometry contrast was weakly associated with the anomaly ($\beta = `r ab_iso_full$b`$ per °C, 95% CI: `r ab_iso_full$lo` to `r ab_iso_full$hi`), but not once calendar year was also in the model ($\beta = `r ab_iso_fully$b`$, 95% CI: `r ab_iso_fully$lo` to `r ab_iso_fully$hi`; REML, Supplementary Table S6). The WorldClim land mask leaves no value at `r clim$n_localities - clim$n_localities_with_data` of the `r clim$n_localities` coordinate localities, which lie roughly 10 km from the nearest valid cell and are not gap-filled. Of the `r .n(rr_n_shared)` shared records, `r .n(ab_miss$records)` from `r ab_miss$localities` localities therefore have no climate value; these localities are coastal (median altitude `r round(ab_miss$median_alt_m)` m against `r round(ab_have$median_alt_m)` m) and were sampled by `r ab_miss$contributors` contributors, so they are not a random subset. A further `r .n(ab_n_cov)` records lacked longitude, altitude or season, leaving `r .n(rr_n_therm)` records for the thermal models. Regional trends in temperature, precipitation and SPEI-12 (Results) were estimated from the locality-by-year values with linear mixed models on year, with random intercepts and slopes for locality and a random intercept for year (Supplementary Note S2).

We tested the association between morphology and $\Delta T_{\text{local}}$ for wing length, log body mass and the isometry contrast ($\Delta_{\text{iso}} = \ln M - 3\ln L$), fitting the fully adjusted set used for the temporal models (sex, latitude, longitude, elevation, season, moult; random intercepts for contributor, municipality and individual; species intercepts with phylogenetic covariance and species anomaly slopes).

Because a temperature anomaly measured over a warming study period is itself correlated with calendar year, and because every bird caught at one locality in one year shares the same anomaly value, we fitted three further specifications of the isometry model, each at the same adjustment level: calendar year and the anomaly entered simultaneously; the anomaly detrended for its linear trend over years; and an additional random intercept for each locality-year. The parsimonious REML versions of these checks are reported in Supplementary Note S4. As a lagged-temperature sensitivity analysis, the fully adjusted thermal models were also refitted with the anomaly of the year before capture (lag 1) and with the mean anomaly of the two years before capture (lags 1–2), each expressed relative to the same site baseline. Because the birds are adults of unknown age, these lags do not identify a developmental window, and for wing length the most relevant exposure is conditions during the most recent moult, which these data cannot resolve (Supplementary Note S4). These lag models were fitted by REML in `glmmTMB` across 50 trees only.

## Phenotypic variability: heteroscedastic models

Residual phenotypic variability was modelled with continuous heteroscedastic models, which estimate temporal change in the residual standard deviation from the full continuous time series while adjusting for covariates. Alongside the model for the mean, the residual standard deviation $\sigma_i$ of each record had its own log-linear model:
$$\log(\sigma_i) = \gamma_0 + \gamma_1 t_i + \gamma_2 \text{Sex}_i \; [+\, f_{k[i]} + g_{j[i]}]$$
where $\gamma_1$ is the change in log residual SD per SD of year, reported as percentage change per decade. In the baseline version the mean model contained sex, year, latitude and the species terms; in the adjusted version both the mean model and the residual-SD model received random intercepts for contributor ($f_k$) and municipality ($g_j$), and the mean model also longitude, elevation and season. A third version added standardised local mean annual temperature (`scaled_tmean`) to the residual-SD model to test whether residual variance tracked thermal conditions. Individual identity is not included in these models, so the residual SD is record-level: it combines among-individual differences, within-individual variation across recaptures (the wing sample includes `r .n(cw_ss$individuals_with_repeats_full)` birds captured more than once), and measurement error, and should not be read as among-individual variability alone.

## Diet

Species' diet was described by the share of invertebrates in the diet, extracted from the EltonTraits 1.0 database [@wilman2014eltontraits] (`Diet-Inv`, % of diet consisting of invertebrates), which places species along a gradient from mostly plant-based diets (fruit, seeds, nectar) to strict insectivory. In continuous interaction models, `Diet-Inv` was standardised ($z$-score) and interacted with calendar year (`scaled_yr * diet_inv_std`). Wing-length trajectories were projected across the 10th (p10), 50th (p50), and 90th (p90) species-weighted percentiles of invertebrate share. In categorical models, diet was collapsed into Invertebrate-dominated ($\ge 50\%$ invertebrate diet) versus Other (frugivores, nectarivores, granivores, omnivores). Both were fitted as a baseline (sex, latitude, species terms) and with random intercepts for contributor and municipality.

## Recovery simulation

To test whether the fully adjusted model can recover a trend in this sampling structure, we simulated wing lengths on the real wing sample, keeping every record's species, sex, individual, contributor, municipality, year and covariates. A generating model with the M3 structure was fitted to the real data (lme4 REML, without the phylogenetic term, which structures species intercepts rather than the year slope). Each simulated response combined its fixed effects, with the year slope replaced by a planted value of 0, −0.5 or −1.0 mm/decade, with fresh species, individual and residual draws at the fitted variances. Contributor and municipality effects followed one of three scenarios: fresh draws independent of year (clean); the fitted conditional modes, which keep their real association with year (real); or fresh draws plus a contributor shift proportional to the contributor's mean capture year, scaled so that the shift alone reproduced the gap between the real baseline and fully adjusted estimates (artefact). Each of the `r rs_reps` data sets per scenario and planted value was fitted with M0, M1, M3 and the within-contributor model, and we recorded bias, root-mean-square error, coverage of the 95% interval and the share of intervals excluding zero (Supplementary Table S4). To put an interval on the attenuation of the wing trend, we also resampled contributors with replacement (`r rs_boot_n` draws, all records of a drawn contributor kept and relabelled) and refitted M0 and M3 to each draw.

## Comparison with the central Amazon

To put our estimates on the same scale as those of @jirinec2021morphological, we used their supplementary Table S6, which gives each species' model-estimated mean mass and wing length in 1980 and 2019 and the change between them. The Amazonian benchmark is the mean change across the 68 passerine species (for body mass, the 77 species including non-passerines give the same value to one decimal place), divided by 3.9 decades. Per-°C rates divide this benchmark by the warming at their site, which the authors report as +1.00°C in the wet season and +1.65°C in the dry season since 1966; our per-°C rates divide the fully adjusted rates by the regional warming rate (Results). Because the benchmark averages species-level estimates and the two studies differ in design, the comparison is descriptive.

## Use of AI tools

Generative AI tools (Anthropic Claude, accessed through Claude Code) were used
during the analysis and preparation of this manuscript. Their use covered
writing and revising analysis scripts in R, executing those scripts, drafting
and revising manuscript text, and auditing the analysis code for reproducibility.
No AI tool was used to generate, alter or impute data, or to produce figures from
anything other than the analysis outputs described above. All analyses were
re-executed from the archived scripts and all reported values are read
programmatically from the saved model objects. All
authors checked the final text and take responsibility for its content.

## Reproducible software environment

The Bayesian models were fitted on Linux servers in R 4.5.3 (R 4.6.1 for the all-record isometry models and the trees refitted with longer chains) with CmdStan 2.39.0 through `cmdstanr` (v0.9.0), using the Stan program archived as `Analysis/scripts/stan/phylo_lmm_marginal.stan`. All other analyses and manuscript generation were conducted in R 4.6.0 with `glmmTMB` (v1.1.14), `prepR4pcm` (v0.5.0.9000), `metafor` (v5.0-1), `lme4` (v2.0-1), `ape` (v5.8-1), `phytools` (v2.5-2), `sf` (v1.1-2) and `terra` (v1.9-27), alongside Python (3.11 or later) with `matplotlib` and `pandas` for the figures. The complete environment, giving the exact version and source of every package, is recorded in `renv.lock` at the root of the archived repository and can be restored with `renv::restore()`; `Analysis/output/session_info.txt` gives the same information in human-readable form. Each saved glmmTMB result object additionally records the R and engine version under which it was fitted. All analysis scripts, curated datasets and the pooled posterior summaries of every fitted model are archived, with the outputs needed to reproduce every reported value, in the project repository ([Zenodo, DOI: [10.5281/zenodo.23070825](https://doi.org/10.5281/zenodo.23070825)]{.content-hidden when-meta="anonymous"}[shared with reviewers through an anonymised link given in the submission form]{.content-visible when-meta="anonymous"}).

# Conflict of interest {.unnumbered .declaration}

The authors declare no conflict of interest.

# Author contributions {.unnumbered .title-page}

E.S.A.S. conceived the study, curated the data, designed and performed the
analyses, produced the figures, and wrote the original draft. A.M., S.O. and
G.B. contributed to the methodology, validated the analyses, and reviewed and
edited the manuscript. All authors approved the final version.

# Funding {.unnumbered .title-page}

G.B. was funded by the European Union's Horizon research and innovation programme under the Marie Skłodowska-Curie grant CAUSATION (Grant Agreement ID: 101150734). A.M. was supported by a Japan Society for the Promotion of Science (JSPS) Overseas Research Fellowship. E.S.A.S. and S.O. were supported by a Canada Excellence Research Chair (CERC-2022-00074).

# Data availability statement {.unnumbered .title-page}

All primary data analysed here are from the previously published, openly
available compilations of morphological records from ATLANTIC BIRD TRAITS
[@rodrigues_atlantic_2019]. Climate data are from the WorldClim historical
monthly weather series [CRU-TS 4.09 downscaled; @harris2020crudata; @fick2017worldclim]. Phylogenies were drawn from the posterior distribution of
@mctavish2025tree. Dietary data are from EltonTraits 1.0 [@wilman2014eltontraits].

Analysis code, the derived analytical datasets, and the pooled posterior
summaries needed to reproduce every value reported here are archived at Zenodo
([https://doi.org/10.5281/zenodo.23070825](https://doi.org/10.5281/zenodo.23070825)).

# Ethics {.unnumbered .declaration}

This study is a secondary analysis of previously published, openly available
records. No new animal capture or handling was undertaken. Capture and ringing
permits for the primary data are those held by the contributing research
programmes compiled in ATLANTIC BIRD TRAITS [@rodrigues_atlantic_2019].

# Acknowledgements {.unnumbered .title-page}

We thank João C.T. Menezes for contributions to the initial design of this study and for comments on the manuscript. We also thank Dr. Guilherme R. R. Brito for his comments about maturation time for the temporal lag analysis. We are grateful to the many field researchers and research groups whose long-term mist-netting efforts are compiled in the ATLANTIC BIRD TRAITS database, without which this work would not have been possible.

## References {.unnumbered}

::: {#refs}
:::

