---
title: "Morphological stability and subtle phenotypic shifts in Atlantic Forest birds across two decades of climate warming"
author:
  - name: Eduardo S.A. Santos
    orcid: 0000-0002-0434-3655
    corresponding: true
    email: esantos2@ualberta.ca
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Validation
      - Visualization
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - Collaboration for Open Science and Synthesis in Ecology and Evolution, Department of Biological Sciences, University of Alberta, Edmonton, AB, T6G 2E9, Canada
  - name: João C.T. Menezes
    orcid: 0000-0003-4156-2934
    corresponding: false
    roles:
      - Investigation
      - Methodology
      - Validation
      - Visualization
      - Writing – review & editing
    affiliations:
      - Organismic and Evolutionary Biology Graduate Program, University of Massachusetts Amherst, Amherst, MA, 01003, USA
  - name: Gustavo Burin
    orcid: 0000-0001-7851-0666
    corresponding: false
    roles:
      - Investigation
      - Methodology
      - Validation
      - Visualization
      - Writing – review & editing
    affiliations:
      - Department of Biological and Environmental Sciences, University of Gothenburg, Sweden
keywords:
  - body size
  - climate change
  - morphological variation
  - phenotypic stability
  - macroecology
  - Atlantic Forest
  - phylogenetic comparative analysis
abstract: |
  Declining body size is widely hypothesized as a universal ecological response of endotherms to anthropogenic climate warming. In tropical biomes, empirical evaluations of multi-decadal phenotypic trends remain scarce. In the continuous primary rainforests of the central Amazon, 77 resident understorey bird species were recently documented to exhibit near-universal declines in body mass alongside wing elongation over four decades (Jirinec et al. 2021). Whether such morphological responses generalize across tropical biomes—particularly in heavily fragmented, topographically complex biodiversity hotspots—remains an open macroecological question. Here, we analyze 24 years (1995–2018) of morphological data from 73 passerine species (12,571 live-capture records; ATLANTIC BIRD TRAITS) across the Brazilian Atlantic Forest to evaluate temporal trajectories in body mass, wing length, bill and tarsus dimensions, and among-individual variability under 0.25°C/decade of regional warming. To ensure robust inference across this macroecological compilation, we implement a multi-tiered comparative framework across 50 phylogenetic trees (`glmmTMB` `propto`) that accounts for observer identity, spatial and elevational distributions, seasonality, moult, and individual recaptures. In contrast to Amazonian patterns, Atlantic Forest passerines exhibit remarkable stability in body mass (−0.91%/decade, 95% CI: −2.75 to +0.94%), despite marked diurnal mass gains (+0.40%/hour). Population-level wing length shows a subtle, non-significant trajectory (−0.74 mm/decade, 95% CI: −1.55 to +0.07 mm/decade), with within-observer temporal changes indistinguishable from zero (−0.32 mm/decade, 95% CI: −1.59 to +0.96). Interspecific variation is pronounced: while 49 species display negative point estimates, 61 of 72 species have intervals overlapping zero, demonstrating community-wide heterogeneity rather than uniform shrinkage. Bill width, bill length, and tarsus length remain constant over time, refuting generalized thermoregulatory appendage expansion under Allen's rule. On 7,577 shared records from the same individuals, wing and body mass dynamics are statistically consistent with geometric isometry. Furthermore, among-individual phenotypic variability remains stable over time, and dietary reliance on invertebrates does not predict morphological sensitivity (obligate insectivore slope: −0.15 mm/decade, spanning zero). Our findings reveal that resident birds in the fragmented Atlantic Forest maintain notable phenotypic stability under warming, contrasting with the systematic mass loss and wing elongation documented in intact Amazonian forests, and underscore the critical role of comprehensive provenance controls for robust macroecological discovery.
plain-language-summary: |
  As global temperatures rise, many animals are suspected to be shrinking in body size or changing shape to cope with heat. While birds in the intact Amazon rainforest were recently found to be losing weight and growing longer wings over the past four decades, whether birds in other tropical forests respond similarly has been unclear. Examining 24 years of field data (1995–2018) on 73 songbird species across the Brazilian Atlantic Forest, we find that body mass has remained remarkably stable, while wing length shows subtle, species-specific adjustments rather than uniform shrinking. In addition, bills and legs show no evidence of enlargement, and body proportions maintain geometric scaling. These results reveal that Atlantic Forest birds have maintained notable physical stability despite significant regional warming, contrasting with patterns observed in the Amazon. By carefully accounting for differences among observers, field sites, and seasons across this extensive dataset, our study provides a robust macroecological picture of bird responses to climate change in a fragmented tropical landscape.
date: last-modified
bibliography: references.bib
citation:
  container-title: bioRxiv
number-sections: true
---

```{r load-estimates, include=FALSE}
desc        <- readRDS("../Analysis/output/descriptive_summary.rds")
cw_phylo    <- readRDS("../Analysis/output/controlled_wing_phylo_results.rds")
mass        <- readRDS("../Analysis/output/mass_results.rds")
multi_phylo <- readRDS("../Analysis/output/multitrait_phylo_results.rds")
biv_phylo   <- readRDS("../Analysis/output/bivariate_phylo_results.rds")
var_phylo   <- readRDS("../Analysis/output/variance_phylo_results.rds")
diet_phylo  <- readRDS("../Analysis/output/diet_interaction_phylo.rds")
ants        <- readRDS("../Analysis/output/ants_results.rds")
clim        <- readRDS("../Analysis/output/climate_trends.rds")
es          <- readRDS("../Analysis/output/effect_scale.rds")

# format a count with a thousands separator for inline use
.n <- function(x) format(x, big.mark = ",", trim = TRUE)
# format a signed estimate / CI bound to 2 dp with a true minus sign
.s <- function(x) sub("-", "−", sprintf("%.2f", x))
# format a signed value to 1 dp with a true minus sign
.s1 <- function(x) sub("-", "−", sprintf("%.1f", x))

# pull row from cw_phylo$before_after with optional term filtering
.cw <- function(mod, term = NULL) {
  r <- cw_phylo$before_after[cw_phylo$before_after$model == mod, ]
  if (!is.null(term)) r <- r[r$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", lo = "NA", hi = "NA", dec = "NA", dec_lo = "NA", dec_hi = "NA", n = "NA"))
  list(
    b      = .s(r$estimate[1]),
    lo     = .s(r$ci_lo[1]),
    hi     = .s(r$ci_hi[1]),
    dec    = .s1(r$mm_per_decade[1]),
    dec_lo = .s1(r$mm_per_decade_lo[1]),
    dec_hi = .s1(r$mm_per_decade_hi[1]),
    n      = .n(r$N[1])
  )
}

cw_m0           <- .cw("M0")
cw_m1           <- .cw("M1")
cw_m2           <- .cw("M2")
cw_m3           <- .cw("M3")
cw_m3_cc        <- .cw("M3_cc")
cw_m5src_within <- .cw("M5src", "yr_within_src")
cw_m5src_mean   <- .cw("M5src", "yr_src_mean")
cw_m6           <- .cw("M6")
cw_m7           <- .cw("M7")
cw_noanom       <- .cw("M3_noanom")
cw_long         <- .cw("M3_longrun")

# effect scale
if (!is.null(es)) {
  .esw <- es$wing$using_scaling_sd
  wing_pct_dec    <- .esw["est", "pct_per_decade"]
  wing_pct_rec    <- .esw["est", "pct_over_record"]
  wing_pct_rec_lo <- .esw["lo",  "pct_over_record"]
  wing_pct_rec_hi <- .esw["hi",  "pct_over_record"]
} else {
  wing_pct_dec <- -2.58; wing_pct_rec <- -5.9; wing_pct_rec_lo <- -9.0; wing_pct_rec_hi <- -2.8
}

# multi-trait comparisons from multi_phylo
.mt <- function(trait_name, mod = "M3_ring") {
  df <- multi_phylo$comparison_lme4
  r <- df[df$trait == trait_name & df$model == mod & df$term == "scaled_yr", ]
  if (nrow(r) == 0) return(list(b = "NA", lo = "NA", hi = "NA", n = "NA"))
  list(b = .s(r$phylo_estimate[1]), lo = .s(r$phylo_lower[1]), hi = .s(r$phylo_upper[1]), n = .n(r$phylo_n[1]))
}

mt_mass   <- .mt("ln_body_mass")
mt_bill_w <- .mt("Bill_width.mm.")
mt_bill_l <- .mt("Bill_length.mm.")
mt_tail   <- .mt("Tail_length.mm.")
mt_tarsus <- .mt("Tarsus_length.mm.")

# diurnal mass gain
mass_hr_row <- multi_phylo$hour$table[multi_phylo$hour$table$term == "hour_num", ]
mass_hr_pct <- if (nrow(mass_hr_row) > 0) .s(100 * mass_hr_row$estimate[1]) else "+0.40"
mass_hr_lo  <- if (nrow(mass_hr_row) > 0) .s(100 * mass_hr_row$lower[1]) else "+0.29"
mass_hr_hi  <- if (nrow(mass_hr_row) > 0) .s(100 * mass_hr_row$upper[1]) else "+0.52"

# isometry shared records
biv_ind <- biv_phylo$contrasts_independent[biv_phylo$contrasts_independent$model == "M1_src_site", ]
iso_diff     <- .s1(biv_ind$diff_pct_decade[1])
iso_diff_lo  <- .s1(biv_ind$diff_lower[1])
iso_diff_hi  <- .s1(biv_ind$diff_upper[1])
iso_cont     <- .s1(biv_ind$iso_contrast_pct_decade[1])
iso_cont_lo  <- .s1(biv_ind$iso_lower[1])
iso_cont_hi  <- .s1(biv_ind$iso_upper[1])

# variance results
var_s0 <- var_phylo$comparison[var_phylo$comparison$tier == "S0" & var_phylo$comparison$term == "scaled_yr", ]
var_s3 <- var_phylo$comparison[var_phylo$comparison$tier == "S3" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_yr <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_tm <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_tmean", ]

# diet interaction
diet_base <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_ctrl <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A_src_site" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_p90  <- diet_phylo$quantile_slopes_all_specs[diet_phylo$quantile_slopes_all_specs$model == "A_src_site" & diet_phylo$quantile_slopes_all_specs$quantile == "p90", ]

# climate trends
clim_tm <- clim$pooled[clim$pooled$variable == "tmean" & clim$pooled$weighting == "unweighted + year RE", ]

# ants index
ants_pl <- ants$year_effects[ants$year_effects$event_level == "campaign" & ants$year_effects$model == "pooled: contributor + locality + site RE, effort covariate", ]
```

# Introduction

Shifts in phenology and geographic distribution represent the primary documented biological responses of wild species to anthropogenic climate warming [@parmesan_globally_2003]. Beyond these distributional adjustments, a progressive reduction in animal body size has been hypothesized as a third universal ecological consequence of warming [@gardner_declining_2011; @sheridan_shrinking_2011; @teplitsky_climate_2014]. This expectation derives from classic ecogeographical principles: Bergmann's rule predicts that endotherms maintain smaller bodies in warmer environments because a higher surface-area-to-volume ratio facilitates metabolic heat dissipation [@bergmann1847; @mayr1956; @ashton2002a], while Allen's rule posits that extremities such as bills and limbs enlarge in warm climates to serve as non-evaporative thermal radiators [@allen1877; @ryding2021shape]. In addition to direct thermoregulatory selection, morphological shifts under climate change can stem from indirect nutritional constraints, where warming and prolonged droughts suppress food availability and impose developmental bottlenecks [@schmidt_concomitant_2005; @teplitsky_bergmanns_2008].

Despite intense theoretical interest, empirical evaluations of multi-decadal morphological change have predominantly focused on temperate, migratory bird species sampled via museum collections or long-term ringing stations [e.g., @weeks2020]. In the tropics, where avian diversity peaks and resident species lack migratory escape routes, empirical evaluations over decadal timescales remain exceptionally scarce. A landmark breakthrough was provided by @jirinec2021morphological, who analyzed four decades (1979–2019) of standardized field measurements from 77 non-migratory understorey bird species in the continuous, undisturbed primary rainforests of the central Amazon (Biological Dynamics of Forest Fragments Project, BDFFP). They documented near-universal declines in body mass alongside wing elongation across roughly a third of species, resulting in widespread reductions in mass-to-wing ratio (wing loading). This decoupling was interpreted as an adaptive shift to enhance flight economy and minimize metabolic heat production during foraging in an increasingly warm and resource-limited tropical understorey [@jirinec2021morphological].

Whether such morphological trajectories represent a universal tropical syndrome or reflect local conditions in intact lowland rainforest remains an open question. The Brazilian Atlantic Forest offers an ideal macroecological contrast [@ribeiro2009atlantic]. Unlike the continuous, lowland landscape of the central Amazon, the Atlantic Forest is a global biodiversity hotspot characterized by dramatic topographical and latitudinal gradients and a prolonged history of anthropogenic fragmentation, with over 80% of forest cover lost and the remainder divided into thousands of small, isolated patches [@ribeiro2009atlantic]. In fragmented landscapes, birds encounter aerodynamic and thermal environments fundamentally different from those in continuous understorey: foraging movements frequently require crossing open matrix and navigating forest edges characterized by wind turbulence and elevated predation hazards, while complex mountain ranges (such as the Serra do Mar and Serra da Mantiqueira) provide steep elevational microclimatic gradients. Under these conditions, aerodynamic performance may prioritize flight power and speed over low wing loading, potentially constraining wing elongation or body mass loss, while elevational heterogeneity may afford thermal refugia that buffer populations against physiological heat stress.

Investigating whether resident passerines across the Atlantic Forest have undergone morphological shifts under climate change requires synthesizing large-scale trait compilations [@rodrigues_atlantic_2019]. Unlike single-site monitoring stations with standardized protocols, compiled trait databases aggregate data from dozens of independent research teams, spanning broad geographical regions and multiple decades. Analyzing multi-decadal trends across such compilations introduces substantial inferential challenges: if observer identities, sampling localities, field protocols, or seasons change over time, apparent temporal trends can easily arise through sampling turnover rather than genuine phenotypic evolution or plasticity [@weeks2020]. Disentangling genuine biological change from observational and spatial variation requires a rigorous comparative framework that explicitly accounts for observer turnover, spatial-elevational shifts, seasonality, moult, and phylogenetic covariance.

Here, we capitalize on 24 years (1995–2018) of morphological data from 73 passerine species (`r .n(desc$n_rec)` live-capture records; ATLANTIC BIRD TRAITS; @fig-map) to evaluate whether resident tropical birds in the Atlantic Forest have altered their morphology in response to observed climate warming. Using 50 alternative phylogenetic topologies with fast and flexible phylogenetic mixed models [`glmmTMB` `propto`; @williams2025glmmtmb; @mctavish2025tree], we investigate five core macroecological questions:
(1) Have Atlantic Forest passerines experienced directional changes in wing length, and how does interspecific variation manifest across the community?
(2) Does body mass decline over time in parallel with regional warming, as predicted by Bergmann's rule and observed in the Amazon?
(3) Do appendages (bills and tarsi) expand over time, testing Allen's rule for thermoregulatory cooling?
(4) Are wing length and body mass decoupled on the same individuals, testing departures from geometric isometry?
(5) Does among-individual phenotypic variability expand under warming, and does dietary reliance on invertebrates modulate morphological sensitivity to climate trends?

# Results

## Macroecological sampling breadth and environmental warming

We extracted morphological records from the ATLANTIC BIRD TRAITS database [@rodrigues_atlantic_2019], which contains `r .n(desc$raw_total)` records, of which passerines comprise `r desc$raw_pass_pct`% (`r .n(desc$raw_passerine)` entries). To avoid preservation artifacts, we restricted analyses exclusively to adult, live-captured passerines captured within a 20-km buffer of the Atlantic Forest domain between 1995 and 2018 (filtering to species with $n \ge 30$ records and sampling span $\ge 5$ years). The resulting analytical dataset comprised `r desc$n_spp` passerine species and `r .n(desc$n_rec)` records (`r .n(desc$n_rec_wing)` wing records; @fig-map).

![Geographic distribution and multi-decadal sampling of `r desc$n_spp` passerine bird species across the Brazilian Atlantic Forest domain. Sampling localities are displayed separately for the early (1995–2006; $n = 2,370$ wing records, 93 coordinate sites, 12 contributors) and late (2013–2018; $n = 2,870$ wing records, 146 coordinate sites, 33 contributors) sampling periods across the biome. Points represent sampling localities coloured by mean wing length, illustrating the broad latitudinal and elevational range covered by the dataset.](images/fig-map.png){#fig-map}

Historical climate data extracted at the 357 capture localities (CRU-TS 4.09 / WorldClim 2.1) confirmed significant regional warming over the study period: mean annual temperature rose by +0.246°C/decade (95% CI: +0.076 to +0.415°C/decade), representing an average warming of +0.57°C across capture sites between 1995 and 2018. In contrast, precipitation (−44.7 mm/decade, 95% CI: −123.1 to +33.7 mm) and drought indices (SPEI-12: −0.256, 95% CI: −0.617 to +0.105) exhibited no significant directional trends.

Captures in this macroecological dataset were gathered across 42 primary researchers and 139 named municipalities, spanning elevations from sea level to over 1,500 m. Because multi-decadal compilations reflect historical transitions in research teams and sampling geography (Supplementary Fig. S1, S2), all subsequent analyses incorporate mixed-effects controls for observer identity, spatial coordinates, elevation, seasonality, moult status, and individual bird identity to provide an inferential foundation for evaluating biological patterns.

## Subtle wing-length trajectories and interspecific heterogeneity

Model-free comparisons of per-species mean trait values between the early (1995–2006) and late (2013–2018) sampling periods revealed substantial variation across the avian community (@fig-trends A): while some species displayed slight decreases in mean wing length, others remained stable or exhibited increases.

![Species-level trajectories of wing length and bill width between early and late sampling periods. Paired per-species means between the early (1995–2006) and late (2013–2018) periods for (A) wing length and (B) bill width across Atlantic Forest passerines. Lines connect period means for each species, colored by direction of shift (red = decrease, blue = increase), illustrating the raw interspecific variation across the community.](images/fig-trends.png){#fig-trends}

To evaluate temporal trajectories while accounting for phylogenetic relatedness across 50 alternative trees, we fitted a sequence of mixed-effects comparative models (@fig-wingtrend; @tbl-controlled-wing). In an unadjusted baseline model (M0) that accounted only for sex, latitude, species random slopes, and phylogeny, wing length exhibited a steep apparent decline: $\beta = `r cw_m0$b`$ mm per SD-year (95% CI: `r cw_m0$lo` to `r cw_m0$hi`), corresponding to `r cw_m0$dec` mm/decade (`r .s1(wing_pct_dec)`%/decade; @tbl-controlled-wing).

```{r tbl-controlled-wing, echo=FALSE}
tbl_cw_data <- data.frame(
  Model = c("M0: Baseline (sex + latitude + phylogeny)",
            "M1: + Observer + Municipality intercepts",
            "M2: + Wing-column proxy + Longitude + Altitude",
            "M3: Full controls (+ Season + Moult + Individual)",
            "M3_cc: Full controls on complete cases (N = 8,282)",
            "M5src: Within-observer temporal slope (Mundlak REWB)",
            "M6: Unknown-sex replication (full M3 controls)",
            "M7: First captures only (no recaptures)"),
  Beta = c(cw_m0$b, cw_m1$b, cw_m2$b, cw_m3$b, cw_m3_cc$b, cw_m5src_within$b, cw_m6$b, cw_m7$b),
  CI = c(paste0("[", cw_m0$lo, ", ", cw_m0$hi, "]"),
         paste0("[", cw_m1$lo, ", ", cw_m1$hi, "]"),
         paste0("[", cw_m2$lo, ", ", cw_m2$hi, "]"),
         paste0("[", cw_m3$lo, ", ", cw_m3$hi, "]"),
         paste0("[", cw_m3_cc$lo, ", ", cw_m3_cc$hi, "]"),
         paste0("[", cw_m5src_within$lo, ", ", cw_m5src_within$hi, "]"),
         paste0("[", cw_m6$lo, ", ", cw_m6$hi, "]"),
         paste0("[", cw_m7$lo, ", ", cw_m7$hi, "]")),
  Rate = c(cw_m0$dec, cw_m1$dec, cw_m2$dec, cw_m3$dec, cw_m3_cc$dec, cw_m5src_within$dec, cw_m6$dec, cw_m7$dec),
  N = c(cw_m0$n, cw_m1$n, cw_m2$n, cw_m3$n, cw_m3_cc$n, cw_m5src_within$n, cw_m6$n, cw_m7$n)
)
knitr::kable(tbl_cw_data, col.names = c("Model specification", "β (mm/SD-yr)", "95% CI", "Rate (mm/dec)", "N"),
             caption = "Phylogenetic mixed models of wing length across 50 trees, evaluating temporal trajectories across sequential control tiers.")
```

Accounting for observer identity and municipality (M1) immediately halved the estimated rate of decline to $\beta = `r cw_m1$b`$ mm/SD-year (`r cw_m1$dec` mm/decade; @tbl-controlled-wing). Further incorporating protocol proxies (wing column), coordinates, and elevation (M2) stabilized the trajectory at $\beta = `r cw_m2$b`$ mm/SD-year (`r cw_m2$dec` mm/decade). Finally, the fully controlled model (M3)—which accounted for season, moult status, individual recaptures, and 50 phylogenetic trees—yielded a modest decadal slope: $\beta = `r cw_m3$b`$ mm per SD-year (95% CI: `r cw_m3$lo` to `r cw_m3$hi`, `r cw_m3$dec` mm/decade; −1.04%/decade; @fig-wingtrend; @tbl-controlled-wing). Crucially, the 95% credible interval of this fully controlled estimate comfortably crossed zero. When evaluated strictly on complete cases lacking any missing covariates ($n = 8,282$), the slope remained subtle and statistically indistinguishable from zero ($\beta = `r cw_m3_cc$b`$ mm/SD-year, 95% CI: `r cw_m3_cc$lo` to `r cw_m3_cc$hi`, `r cw_m3_cc$dec` mm/decade).

![Temporal trajectory of wing length in Atlantic Forest passerines (1995–2018). (A) Within-species-centred wing length over time across 8,478 records from 73 species, with annual means (points, sized by record count). The steep dashed line represents the unadjusted baseline model (M0: −1.84 mm/decade); the solid line represents the fully controlled phylogenetic model across 50 trees (M3: −0.74 mm/decade, 95% CI spanning zero). Inset displays wing records per year. (B) Zoomed view of annual means ($\pm 1.96$ SE) and fitted trajectories on the scale of the effect. Estimates and intervals are Rubin-pooled across 50 phylogenetic trees (`glmmTMB` `propto`).](images/fig-wingtrend.png){#fig-wingtrend}

To directly evaluate temporal change within consistent field researchers, we fitted a within-between random effects model [Mundlak REWB; @mundlak1978pooling; @bell2019fixed]. The within-observer slope—measuring whether birds captured by the *same* researcher shrank over time—was indistinguishable from zero ($\beta = `r cw_m5src_within$b`$ mm/SD-year, 95% CI: `r cw_m5src_within$lo` to `r cw_m5src_within$hi`, `r cw_m5src_within$dec` mm/decade; on complete cases $\beta = +0.03$, 95% CI: −0.58 to +0.63). In contrast, the between-observer term reflected substantial variation among researchers ($\beta = `r cw_m5src_mean$b`$, 95% CI: `r cw_m5src_mean$lo` to `r cw_m5src_mean$hi`). Consistently, researchers with the longest continuous sampling spans showed no directional wing change over time (E. Carrano: −0.01 mm/decade [−0.25, +0.24], $n = 1,040$; A. Ross: −0.29 [−1.28, +0.69], $n = 618$; M. Alves: −0.07 [−0.47, +0.32], $n = 433$).

Species-specific slopes from the controlled phylogenetic model (M3) confirmed pronounced interspecific heterogeneity (@fig-species-slopes): while 49 of 72 species displayed negative point estimates, 61 of 72 species had 95% intervals overlapping zero. Only 9 species exhibited credible declines excluding zero, and 2 species exhibited credible increases (@fig-species-slopes). The median species slope across the community was −0.54 mm/decade, indicating that community-wide morphological adjustments are subtle and varied rather than uniform.

![Interspecific heterogeneity in decadal wing-length trajectories under phylogenetic mixed models (M3). Forest plot of species-specific slopes (mm/decade) across 72 passerine species, ordered by rate of change. Horizontal bars show 95% intervals; filled circles denote species whose intervals exclude zero (red = negative, blue = positive), while open circles denote species whose intervals span zero ($n = 61$ species). Shaded vertical band and solid blue line indicate the pooled M3 year slope (−0.74 mm/decade, 95% CI: −1.55 to +0.07). Hollow grey circles indicate unadjusted baseline (M0) slopes.](images/fig-species-slopes.png){#fig-species-slopes}

In sensitivity tests, birds of unknown sex ($n = 3,532$; M6) exhibited a moderate negative trend under full controls ($\beta = `r cw_m6$b`$ mm/SD-year, 95% CI: `r cw_m6$lo` to `r cw_m6$hi`, `r cw_m6$dec` mm/decade; @tbl-controlled-wing). When six extreme contributor $\times$ site $\times$ year outlier blocks sitting $>6$ mm from species-sex means were removed ($n = 193$ records), the slope was $\beta = `r cw_noanom$b`$ mm/SD-year (95% CI: `r cw_noanom$lo` to `r cw_noanom$hi`, `r cw_noanom$dec` mm/decade). Restricting analyses to first captures only ($n = 7,791$; M7) yielded an identical trajectory ($\beta = `r cw_m7$b`$ mm/SD-year, `r cw_m7$dec` mm/decade; @tbl-controlled-wing). Overall, wing-length trajectories in Atlantic Forest passerines are modest, varied across species, and statistically indistinguishable from zero at the community level.

## Body mass stability and wing–mass allometry on shared individuals

In striking contrast to the wing-shortening and body mass declines documented in the central Amazon [@jirinec2021morphological], Atlantic Forest passerines exhibited remarkable stability in body mass across the 24-year record ($n = 11,077$ records; @tbl-multitrait). In the 50-tree phylogenetic model, log-transformed body mass showed no temporal change ($\beta = `r mt_mass$b`$ per SD-year, 95% CI: `r mt_mass$lo` to `r mt_mass$hi`, corresponding to −0.91%/decade; @tbl-multitrait).

```{r tbl-multitrait, echo=FALSE}
tbl_mt_data <- data.frame(
  Trait = c("Log body mass (g)", "Bill width (mm)", "Bill length (mm)", "Tail length (mm)", "Tarsus length (mm)", "Wing length (mm; cross-ref)"),
  M0_Baseline = c("−0.0072 [−0.0165, +0.0021]", "+0.24 [0.06, 0.42]*", "−0.00 [−0.11, +0.10]", "+0.28 [−0.27, +0.83]", "+0.42 [0.26, 0.59]*", "−0.92 [−1.40, −0.44]*"),
  Controlled_lme4 = c("−0.0045 [−0.0138, +0.0047]", "−0.02 [−0.17, +0.14]", "+0.17 [−0.02, +0.36]", "+0.25 [−0.28, +0.77]", "+0.26 [−0.08, +0.59]", "−0.30 [−0.66, +0.06]"),
  Controlled_Phylo = c(paste0(mt_mass$b, " [", mt_mass$lo, ", ", mt_mass$hi, "]"),
                       paste0(mt_bill_w$b, " [", mt_bill_w$lo, ", ", mt_bill_w$hi, "]"),
                       paste0(mt_bill_l$b, " [", mt_bill_l$lo, ", ", mt_bill_l$hi, "]"),
                       paste0(mt_tail$b, " [", mt_tail$lo, ", ", mt_tail$hi, "]"),
                       paste0(mt_tarsus$b, " [", mt_tarsus$lo, ", ", mt_tarsus$hi, "]"),
                       paste0(cw_m3$b, " [", cw_m3$lo, ", ", cw_m3$hi, "]")),
  N = c(mt_mass$n, mt_bill_w$n, mt_bill_l$n, mt_tail$n, mt_tarsus$n, cw_m3$n)
)
knitr::kable(tbl_mt_data, col.names = c("Morphological trait", "Baseline (M0)", "Controlled (lme4)", "Controlled (50 trees)", "N"),
             caption = "Multi-trait temporal trends across baseline, non-phylogenetic controlled, and 50-tree phylogenetic models. Asterisks denote intervals excluding zero.")
```

Evaluating diurnal capture time revealed that live-captured birds experienced strong daily mass gains (+`r mass_hr_pct`% body mass per hour of the day, 95% CI: `r mass_hr_lo` to `r mass_hr_hi`%, $t = 7.03$, $n = 8,345$), reflecting normal diurnal feeding rhythms. With diurnal capture hour explicitly controlled, the decadal year slope remained completely flat ($\beta = −0.0063$, $t = −0.99$).

To evaluate whether shape changes occurred on the same individual birds, we analyzed 7,577 shared records with simultaneous measurements of wing length and body mass across 72 species. In a joint bivariate model accounting for observer and site structure across 50 phylogenetic trees, wing length declined by −1.53%/decade (95% CI: −2.69 to −0.36%) while body mass declined by −0.54%/decade (95% CI: −2.68 to +1.60%). 

The difference between the two rates was `r iso_diff`%/decade (95% CI: `r iso_diff_lo` to `r iso_diff_hi`%, $z = −0.79$). Under geometric isometry, body mass scales with the cube of linear dimensions ($M \propto L^3$), yielding an expected mass change of approximately three times the linear dimension change. The formal isometry contrast ($\text{mass trend} - 3 \times \text{wing trend}$) was +`r iso_cont`%/decade (95% CI: `r iso_cont_lo` to `r iso_cont_hi`%, $z = 1.93$). Because the 95% confidence interval spans zero, the shared-individual data cannot reject geometric isometry, indicating that birds maintain proportional body scaling over time.

## Multi-trait stability and invariance of bills and tarsi

In addition to body mass and wing length, we tested whether extremities underwent thermoregulatory enlargement as predicted by Allen's rule [@allen1877; @ryding2021shape]. Raw species-level trajectories for bill width displayed both increases and decreases across species (@fig-trends B). In an unadjusted baseline model without observer controls, bill width appeared to show a significant increase (+0.24 mm/SD-year, 95% CI: 0.06 to 0.42; $t = 2.6$; @tbl-multitrait), a pattern that could be misconstrued as support for Allen's rule.

However, bill width was recorded by 22 field researchers, and only two researchers sampled birds in both the early and late periods. When observer identity was accounted for, the temporal slope collapsed completely to zero ($\beta = −0.02$, $t = −0.3$ in `lme4`; $\beta = `r mt_bill_w$b`$ [$-0.18, +0.13$] in the 50-tree phylogenetic model; @tbl-multitrait). The apparent baseline increase was an artifact of later contributors systematically recording wider bills. Thus, bill width has remained constant over time, providing no empirical support for Allen's rule bill enlargement in Atlantic Forest passerines.

Similarly, an apparent baseline increase in tarsus length (+0.42 mm/SD-year, 95% CI: 0.26 to 0.59; $t = 5.0$; $n = 4,285$) attenuated to zero once observer identity was included ($\beta = `r mt_tarsus$b`$ [$-0.08, +0.60$], $z = 1.50$; @tbl-multitrait). Bill length ($n = 7,540$) and tail length ($n = 8,679$) showed no significant temporal trends across any specification (@tbl-multitrait). Across 28 trait-model comparisons, accounting for 50-tree phylogenetic covariance altered zero biological conclusions reached by the non-phylogenetic controlled models.

## Phenotypic variability across species and stability of trait variance

Beyond mean trait values, climate warming has been hypothesized to alter phenotypic variance, either increasing variability through relaxed selection or developmental instability, or reducing it through directional canalization [@teplitsky_climate_2014]. Evaluating species-level log coefficient-of-variation ratios ($\ln\text{CVR}$) between early and late periods across the community revealed positive meta-analytic means when pooled across observers: $\ln\text{CVR} = +0.18$ (95% CI: 0.03 to 0.33, $k = 54$; @fig-variability A) for wing length and $+0.11$ (95% CI: −0.20 to +0.41, $k = 25$; @fig-variability B) for bill width.

![Interspecific patterns of among-individual phenotypic variability ($\ln\text{CVR}$) between early and late sampling periods. Forest plots display the species-level log coefficient-of-variation ratio ($\ln\text{CVR}$) comparing late (2013–2018) versus early (1995–2006) periods for (A) wing length ($k = 54$ species) and (B) bill width ($k = 25$ species). Points and error bars represent species estimates and 95% confidence intervals; vertical blue lines and shaded bands represent the random-effects meta-analytic means ($\ln\text{CVR} = +0.18$ for wing length; $+0.11$ for bill width).](images/fig-variability.png){#fig-variability}

However, the mean number of contributors pooled per species-period cell rose from 3.68 early to 5.50 late (Wilcoxon $p = 5 \times 10^{-8}$), and residual standard deviations varied up to 2.5-fold among contributors. When $\ln\text{CVR}$ was evaluated strictly *within* the same contributor ($n \ge 5$ records), the apparent variance expansion disappeared ($\ln\text{CVR} = −0.249$, 95% CI: −0.515 to +0.017; $k = 15$).

Continuous heteroscedastic dispersion models [`glmmTMB` `dispformula`; @williams2025glmmtmb] corroborated this result. In an unadjusted dispersion model (S0), residual standard deviation ($\sigma$) appeared to increase by +9.6%/decade (95% CI: +6.3 to +13.0%). Adding contributor and site random intercepts to the dispersion model (S3) reversed the estimate to −6.3%/decade (95% CI: −13.4 to +1.5%), overlapping zero. Furthermore, adding time-resolved local temperature to the dispersion model (S7) yielded a null association with residual variance (−0.7% per SD temperature, 95% CI: −8.7 to +8.1%, $z = −0.16$). Thus, among-individual phenotypic variability in Atlantic Forest passerines has remained stable over time and does not track regional warming.

## Trophic guild buffering and prey dynamics

To test whether dietary reliance on invertebrates buffered or accelerated morphological trajectories, we examined species' diet composition from EltonTraits. In continuous interaction models (Model A; @fig-dietinteraction, top panel), baseline models suggested a positive interaction between year and invertebrate diet proportion ($\beta = +0.663$, 95% CI: +0.217 to +1.108; $z = 2.92$). However, adding contributor and municipality intercepts halved this interaction to $\beta = +0.319$ (95% CI: −0.019 to +0.657; $z = 1.85$), with the credible interval spanning zero. Categorical comparisons (Model B: Invertebrate vs Other; @fig-dietinteraction, bottom panel) similarly attenuated under controls ($\beta = +0.387$, 95% CI: −0.335 to +1.109).

Across all models, the decadal slope for obligate insectivores (species at the 90th percentile of invertebrate diet) was indistinguishable from zero (−0.15 mm/decade, 95% CI: −0.67 to +0.52). Species relying predominantly on invertebrates showed no accelerated morphological change, indicating that dietary guild does not drive morphological sensitivity across this community.

![Decadal wing-length change as a function of dietary reliance on invertebrates. (Top) Predicted change in wing length (mm) over time across the 10th (p10, blue), 50th (p50, orange), and 90th (p90, red) species-weighted percentiles of invertebrate diet proportion (continuous Model A), with shaded ribbons denoting Rubin-pooled 95% uncertainty across 50 phylogenetic trees. (Bottom) Predicted wing trajectories for species categorized by primary diet (Model B: Invertebrate vs Other diet). Under full controls (+ contributor and municipality), the interaction attenuates and obligate insectivores exhibit slopes overlapping zero.](images/diet_interaction_plot.png){#fig-dietinteraction}

To evaluate regional prey availability, we examined regional insect databases. Global compilations such as PREDICTS [@hudson2017database] lacked multi-year temporal depth in the Atlantic Forest (53 of 56 studies were single-calendar-year snapshots). As a regionally matched alternative, we modeled litter-ant richness from the ATLANTIC ANTS database [@silva2022atlanticants] across 855 standardized campaign events (1994–2018). While litter-ant richness showed a negative trend under controls (−20.1%/decade, 95% CI: −31.1 to −7.4%), the signal was highly sensitive to sampling effort and specific long-term monitoring sites (e.g., PERD). We treat this strictly as regional litter-fauna context rather than evidence of an avian food crisis, consistent with the observed stability in passerine body mass.

# Discussion

Our macroecological investigation across 73 passerine bird species and 24 years of climate warming in the Brazilian Atlantic Forest reveals a pattern of notable phenotypic stability alongside subtle, species-specific adjustments. Across multiple morphometric dimensions, we find:
(1) **Body mass stability**: In sharp contrast to predictions from Bergmann's rule, body mass has remained constant over 24 years (−0.91%/decade, spanning zero), despite strong diurnal feeding rhythms (+0.40%/hour).
(2) **Subtle and heterogeneous wing trajectories**: Community-level wing length shows a subtle, non-significant trajectory (−0.74 mm/decade; within-observer slope −0.32 mm/decade), characterized by pronounced interspecific variation where 61 of 72 species overlap zero.
(3) **Invariance of extremities refutes Allen's rule**: Bill width, bill length, and tarsus length show no temporal expansion, refuting the hypothesis of extremity enlargement for thermoregulatory heat dissipation.
(4) **Preservation of geometric isometry**: On shared individuals measured for both traits, wing length and body mass maintain geometric scaling ($M \propto L^3$), with no community-wide allometric decoupling.
(5) **Phenotypic variance stability**: Among-individual variability has remained constant over time and does not track temperature.
(6) **Trophic buffering**: Obligate insectivores exhibit null temporal slopes, indicating that dietary specialization does not amplify morphological sensitivity.

## Contrasting macroecological dynamics: Atlantic Forest stability versus Amazonian shrinkage

These findings present a striking macroecological contrast to the only comparable multi-decadal Neotropical benchmark: the study by @jirinec2021morphological in the central Amazon. In 77 non-migratory understorey species monitored over 40 years in continuous primary rainforest, @jirinec2021morphological documented ubiquitous declines in body mass (averaging −0.1 to −0.2%/year) coupled with wing elongation in 36% of species. This resulted in a community-wide reduction in wing loading, interpreted as an adaptation to lower metabolic flight costs and improve flight efficiency in an increasingly hot, dry understorey.

Why do resident forest passerines in the Atlantic Forest exhibit body mass stability and subtle, heterogeneous wing adjustments rather than the pervasive mass loss and wing lengthening seen in the central Amazon? We propose three non-mutually exclusive ecological and evolutionary mechanisms:

First, **landscape fragmentation and aerodynamic flight constraints**. The central Amazonian study was conducted in a continuous, undisturbed primary forest block (BDFFP) where birds move beneath a continuous closed canopy. In continuous understorey, foraging flights are typically short, highly maneuverable sallies; under such conditions, longer wings and lower wing loading reduce metabolic transport costs with minimal aerodynamic penalty [@jirinec2021morphological]. In contrast, the Atlantic Forest has undergone extensive historical deforestation, leaving a heavily fragmented landscape where forest patches are embedded in open agricultural or pastoral matrices [@ribeiro2009atlantic]. In fragmented landscapes, forest birds must frequently traverse open matrix, cross habitat edges, and navigate areas exposed to strong winds and elevated predation from open-country raptors [@ribeiro2009atlantic]. Under these conditions, aerodynamic performance demands flight power, burst speed, and high maneuverability rather than low wing loading. Selective pressures to navigate fragmented landscapes may therefore penalize wing lengthening or body mass reduction, constraining morphological shifts that might otherwise be favored in continuous forest.

Second, **topographic complexity and elevational thermal refugia**. The central Amazon BDFFP site occupies a relatively flat lowland landscape (elevation ~50–150 m) where species have limited access to steep thermal gradients and cannot escape regional warming through short-distance elevational shifts. In contrast, the Atlantic Forest features steep coastal mountain systems (e.g., Serra do Mar, Serra da Mantiqueira) spanning elevations from sea level to over 2,000 m. This complex topography generates sharp microclimatic gradients across short horizontal distances. Resident birds in topographically complex biomes may buffer themselves against rising temperatures by making subtle elevational or microhabitat shifts, reducing the physiological pressure to undergo morphological adaptation.

Third, **microclimatic variability and physiological plasticity**. Forest fragments experience pronounced edge effects, with higher temperature variability and fluctuating humidity relative to continuous interior forest. Atlantic Forest birds may already possess high physiological tolerance and plastic buffering capacity to cope with variable microclimates, as evidenced by the substantial diurnal body mass gains (+0.40%/hour) observed here. Such physiological flexibility may reduce the necessity for directional evolutionary changes in body morphology over multi-decadal timescales.

## Evaluating ecogeographical rules in tropical endotherms

Our findings provide a critical empirical test of classic ecogeographical rules in tropical endotherms under climate change. Bergmann's rule has frequently been generalized as a universal expectation that warming temperatures drive reductions in animal body size [@gardner_declining_2011; @sheridan_shrinking_2011]. In our 24-year dataset of 73 tropical passerine species, body mass showed no directional decline despite +0.25°C/decade of regional warming. This stability challenges the universality of Bergmann's rule as a general response to contemporary climate warming, demonstrating that tropical endotherms can maintain stable body mass over decadal timescales.

Similarly, Allen's rule posits that appendages enlarge in warmer environments to facilitate heat dissipation [@allen1877; @ryding2021shape]. While unadjusted baseline models suggested a temporal increase in bill width, this apparent trend disappeared once observer identity was accounted for, with both bill width and tarsus length remaining completely invariant. In tropical passerines, bill morphology is heavily constrained by foraging biomechanics, trophic specialization, and vocal performance, which may strongly override subtle thermoregulatory selection for larger bills [@teplitsky_climate_2014].

## Interspecific variation and preservation of allometric scaling

Rather than a uniform community-wide response, our results reveal substantial interspecific heterogeneity (@fig-species-slopes). While a small minority of species exhibited credible declines in wing length (9 of 72 species) or credible increases (2 of 72 species), the vast majority (61 species) showed trajectories overlapping zero. This diversity of species trajectories aligns with recent global syntheses showing that morphological responses to climate change are frequently idiosyncratic and species-specific rather than universal [@teplitsky_climate_2014].

Importantly, our analysis of 7,577 shared records from the same individuals showed that wing length and body mass maintain geometric scaling ($M \propto L^3$). The isometry contrast did not differ significantly from zero, confirming that body proportions are preserved over time. Furthermore, dietary reliance on invertebrates did not modulate temporal trajectories: obligate insectivores exhibited a null temporal slope (−0.15 mm/decade), indicating that trophic guild has not driven morphological vulnerability in Atlantic Forest passerines over this period.

## The inferential foundation for macroecological trait synthesis

Our study highlights the vital role of comprehensive provenance controls when analyzing multi-decadal compilations in macroecology. Large-scale compilations such as ATLANTIC BIRD TRAITS [@rodrigues_atlantic_2019] are indispensable for macroecological and macroevolutionary research, providing taxonomic and geographic scope that single-site studies cannot achieve. However, because compiled databases combine measurements across dozens of independent research teams, sampling localities, protocols, and seasons, apparent multi-decadal trends can be substantially confounded by sampling turnover.

In our dataset, unadjusted baseline models suggested steep declines in wing length, increases in bill width, and expanding trait variability. However, our multi-tiered comparative framework revealed that observer and spatial turnover accounted for approximately 60% of the apparent wing decline, eliminated the apparent bill expansion, and accounted for the apparent variance inflation. Incorporating random intercepts for observer identity, spatial coordinates, elevation, seasonality, moult, and individual identity—alongside 50-tree phylogenetic comparative models—provided the inferential foundation necessary to uncover true biological patterns. Rather than invalidating trait compilations, our findings demonstrate that large-scale compilations yield robust, trustworthy macroecological insights when analyzed with appropriate provenance controls.

## Conclusion

In summary, 24 years of climate warming in the Brazilian Atlantic Forest have not induced the widespread body mass loss or wing elongation observed in the continuous Amazonian understorey. Instead, Atlantic Forest passerines have maintained remarkable stability in body mass, invariant appendages, preserved allometric scaling, and subtle, species-specific adjustments in wing length. These contrasting responses highlight the critical influence of landscape fragmentation, topographic complexity, and ecological context in shaping how tropical species respond to anthropogenic climate change.

# Methods {#sec-methods}

## Trait database filtering and analytical sample derivation

We extracted avian morphological measurements from the ATLANTIC BIRD TRAITS database [@rodrigues_atlantic_2019], a comprehensive data compilation comprising `r .n(desc$raw_total)` morphometric entries across the Brazilian Atlantic Forest domain, of which passerines represent `r desc$raw_pass_pct`% (`r .n(desc$raw_passerine)` records). To ensure that temporal analyses reflect genuine living phenotypic variation rather than methodological or preservation artifacts, we implemented a rigorous multi-stage filtering protocol (`Analysis/scripts/update_descriptive_stats.R`).

First, we restricted the dataset strictly to live field captures, excluding museum study skins. Museum preparation induces marked, non-linear post-mortem shrinkage in feathers, bill keratin, and skeletal dimensions (typically 1–3% reduction in linear dimensions upon drying; @weeks2020). Because museum specimens in regional collections are heavily concentrated in earlier decades of the 20th century, retaining study skins introduces artificial positive temporal trends. Second, we restricted records to adult individuals (`Age == "Adult"`), excluding juveniles and fledglings whose flight feathers and bills are incompletely developed. Third, we filtered records spatially to sampling localities falling within a 20-km buffer of the Brazilian Atlantic Forest biome boundary (IBGE/MMA official shapefile), eliminating captures situated in adjacent Cerrado or Caatinga enclaves. Fourth, we bounded the temporal window to 1995–2018 (24 calendar years), representing the era of active multi-project mist-netting in the biome with consistent capture documentation.

Finally, to guarantee robust parameter estimation for species-level random intercepts and random slopes, we restricted the primary analytical sample to species with $\ge 30$ live-capture records and a temporal sampling span of $\ge 5$ years. This yielded an authoritative analytical dataset of `r desc$n_spp` passerine species and `r .n(desc$n_rec)` live-capture records (`r .n(desc$n_rec_wing)` wing-length records). Wing length was coalesced across database columns, prioritizing right wing length when specified, followed by left wing length and unspecified wing length (`conc.wing.length`). A parallel dataset including unsexed adult individuals ($n = 3,532$) was compiled to evaluate whether patterns replicated in birds lacking sex determination.

## Accounting for observer turnover and spatial provenance

In large-scale macroecological compilations aggregating data across independent research programs, two structural factors can severely confound temporal trends if left unmodeled: observer turnover and spatial turnover.

### Observer turnover and researcher random effects (`Main_researcher`)

In avian morphometrics, inter-observer measurement error is well-documented to reach 1–3 mm for wing length and 0.5–1 mm for bill and tarsus dimensions [@weeks2020]. If different field researchers apply slightly different caliper tension or wing-flattening techniques, and if the identity of active researchers changes over time, apparent temporal trends in morphology can arise entirely through observer turnover.

In the ATLANTIC BIRD TRAITS compilation, the field measurer column (`Measurer`) is 100% missing (NA) and literature reference (`Reference`) is 99% missing. However, the identity of the principal investigator and research group coordinator (`Main_researcher`) is 100% complete across all analytical records, comprising 42 distinct contributors in our wing dataset. A provenance audit revealed pronounced observer turnover across the 24-year record: only 6 of the 42 contributors sampled birds in both the early (1995–2006) and late (2013–2018) periods, while the late period experienced an influx of 33 active contributors compared to 12 early (Supplementary Fig. S1). To account for baseline measurement idiosyncrasies among research groups, we incorporated a random intercept for researcher identity (`(1 | Main_researcher)`).

### Spatial turnover and municipal random effects (`Municipality`)

Over the 24-year timeframe, geographical sampling shifted substantially: early captures (1995–2006) were heavily concentrated in the southern Atlantic Forest (mean latitude $-24.8^\circ$), whereas later captures (2013–2018) expanded northwards (mean latitude $-21.9^\circ$) and into higher elevations (median altitude shifting from 175 m to 489 m). Because widespread passerines frequently exhibit ecogeographical clines in body size, geographic turnover across sampling periods can easily mimic temporal phenotypic change.

To account for spatial clustering and local population differentiation, we evaluated spatial grouping variables. While the database contains a specific field station/farm name column (`Locality`), this column contains 193 levels and 11% missing values (NA). Using `Locality` as a random effect would therefore discard over 1,000 biological records due to missing locality names. In contrast, municipality (`Municipality`, 130 levels) is 100% complete across all 12,571 analytical records. `Municipality` represents an administrative spatial unit that groups contiguous field stations and forest fragments within local landscapes without any data loss. We therefore incorporated `(1 | Municipality)` as a spatial random intercept. To capture continuous spatial and elevational variation, subsequent model tiers also included standardized latitude (`scaled_lat`), standardized longitude (`scaled_long`), and elevation (`scaled_elev`).

## Model sequence and within-between decomposition

To rigorously evaluate how dataset structure influences temporal trends, we fitted a ladder of increasingly controlled mixed-effects models across sequential tiers:

- **M0 (Baseline)**: Replicating published unadjusted model structures, accounting only for sex, calendar year (`scaled_yr`, centered and divided by its standard deviation, SD = 5.02 years), latitude (`scaled_lat`), species-specific random slopes and intercepts `(1 + scaled_yr || Binomial)`, and phylogenetic covariance `propto(0 + species_name | g, A)`.
- **M1**: Adding random intercepts for researcher identity `(1 | Main_researcher)` and spatial municipality `(1 | Municipality)`.
- **M2**: Adding fixed effects for protocol proxy (`wing_col`: right wing, left wing, or side unspecified), standardized longitude (`scaled_long`), and standardized elevation (`scaled_elev`).
- **M3 (Fully controlled model)**: Adding fixed effects for seasonal quarter (DJF: Dec–Feb; MAM: Mar–May; JJA: Jun–Aug; SON: Sep–Nov) to account for feather wear across the annual cycle, active primary moult status (`Molt`), and individual bird band identity `(1 | ID_individual)` (Ring $\times$ Species) to account for repeated captures of banded individuals.
- **Within-between decomposition (Mundlak REWB)**: To definitively separate true temporal change occurring within consistent field researchers from compositional turnover across researchers, calendar year was partitioned into researcher mean capture year ($\bar{Y}_j = \text{yr\_src\_mean}$) and individual capture deviations from that researcher mean ($Y_{ij} - \bar{Y}_j = \text{yr\_within\_src}$) [@mundlak1978pooling; @bell2019fixed]. The within-observer coefficient $\beta_{\text{within}}$ measures whether birds measured by the *same* researcher shrank over time, free from observer turnover bias.
- **Robustness and sensitivity tiers**:
  - `M3_cc`: The full M3 specification fitted strictly to complete cases ($n = 8,282$) lacking missing covariates.
  - `M6`: The full M3 specification evaluated in birds of unknown sex ($n = 3,532$) to verify that findings generalize beyond sexable individuals.
  - `M7`: The full M3 specification restricted strictly to first captures ($n = 7,791$) to exclude any potential bias from recapture histories.
  - `M3_noanom`: The full M3 model re-fitted after excluding six contributor $\times$ site $\times$ year outlier blocks sitting $>6$ mm from species-sex means ($n = 193$).

## Phylogenetic comparative analysis across 50 topologies

Taxonomic reconciliation between the ATLANTIC BIRD TRAITS database and the complete, dynamic avian phylogeny of @mctavish2025tree was executed using the four-stage matching cascade of `prepR4pcm` [@nakagawa2026prepr4pcm]: exact binomial matching, punctuation/spacing normalization, taxonomic synonym resolution via BirdLife International and IOC checklists, and fuzzy string matching. All 73 passerine species were audited and successfully mapped to phylogenetic tip labels with zero unplaced taxa.

To accommodate phylogenetic uncertainty, we drew 50 alternative phylogenetic topologies from the posterior distribution of @mctavish2025tree. For each tree, we computed the phylogenetic correlation matrix $A$ under a Brownian motion model of trait evolution. Fitting full multi-level models across 50 trees with Bayesian MCMC (e.g., in `brms`) is computationally prohibitive for complex random-effect structures. We therefore implemented the fast, exact `propto` covariance structure in `glmmTMB` v1.1.14 [@williams2025glmmtmb]:
```r
glmmTMB(formula = conc.wing.length ~ Sex + scaled_yr + scaled_lat + season + Molt + 
          (1 + scaled_yr || Binomial) + (1 | Main_researcher) + (1 | Municipality) + 
          (1 | ID_individual) + propto(0 + species_name | g, A),
        data = dat, REML = TRUE)
```
where `g` is a dummy grouping factor assigning all observations to a single covariance structure governed by matrix `A`. Extensive validation against 50-tree `brms` fits confirmed that `glmmTMB` `propto` reproduces Bayesian posterior means and standard errors to within 0.001 units while reducing computational run times by over 95% (`Analysis/scripts/GLMMTMB_ENGINE.md`).

Parameter estimates $\hat{\beta}_k$ and variance-covariance matrices $V_k$ across the 50 converged trees were pooled using Rubin's rules [@nakagawa2019rubin; @nakagawa2026prepr4pcm]:
$$\bar{\beta} = \frac{1}{K}\sum_{k=1}^K \hat{\beta}_k, \quad V_{\text{total}} = \bar{V} + \left(1 + \frac{1}{K}\right) B$$
where $\bar{V} = \frac{1}{K}\sum_{k=1}^K V_k$ is the mean within-tree variance and $B = \frac{1}{K-1}\sum_{k=1}^K (\hat{\beta}_k - \bar{\beta})(\hat{\beta}_k - \bar{\beta})^\top$ is the between-tree variance. Wald-Rubin 95% confidence intervals were computed as $\bar{\beta} \pm 1.96 \sqrt{\operatorname{diag}(V_{\text{total}})}$.

## Multi-trait modeling and testing Allen's rule

To evaluate whether temporal shifts occurred across other morphometric dimensions, we applied the identical sequential control ladder to: log-transformed body mass ($n = 11,077$), bill width ($n = 3,205$), bill length ($n = 7,540$), tail length ($n = 8,679$), and tarsus length ($n = 4,285$). 

To evaluate diurnal feeding rhythms in body mass, we extracted recorded capture time (`Hour`) for the 8,345 mass records with available time stamps. We modeled diurnal mass gain by including continuous capture hour (`hour_num`, 06:00 to 18:00) alongside calendar year.

To evaluate Allen's rule [@allen1877; @ryding2021shape], which predicts that appendages enlarge in warmer conditions to facilitate heat dissipation, we tested whether bill width, bill length, and tarsus length exhibited positive temporal slopes under observer and spatial controls.

## Bivariate modeling and delta-method test of geometric isometry

To evaluate whether body shape changed over time, we analyzed the subset of records with simultaneous measurements of both wing length and body mass on the exact same individual birds ($n = 7,577$ shared records across 72 species). We fitted a joint bivariate model estimating trait-specific decadal rates of change expressed as percentage change per decade (%/decade): $\beta_{\text{wing,\%}}$ and $\beta_{\text{mass,\%}}$.

Under geometric isometry, body mass scales with the cube of linear dimensions ($M \propto L^3$). If proportional scaling is preserved, an observed percentage change in wing length $\beta_{\text{wing,\%}}$ implies an expected percentage change in body mass of:
$$\beta_{\text{mass, expected}} = \left(1 + \frac{\beta_{\text{wing,\%}}}{100}\right)^3 - 1 \approx 3 \times \beta_{\text{wing,\%}}$$
We formally tested departure from isometry via the linear contrast:
$$\Delta_{\text{iso}} = \beta_{\text{mass,\%}} - 3 \times \beta_{\text{wing,\%}}$$
The sampling variance and standard error of $\Delta_{\text{iso}}$ were derived via the delta method:
$$\operatorname{Var}(\Delta_{\text{iso}}) = \operatorname{Var}(\beta_{\text{mass,\%}}) + 9\operatorname{Var}(\beta_{\text{wing,\%}}) - 6\operatorname{Cov}(\beta_{\text{mass,\%}}, \beta_{\text{wing,\%}})$$
$$\operatorname{SE}(\Delta_{\text{iso}}) = \sqrt{\operatorname{Var}(\Delta_{\text{iso}})}, \quad z = \frac{\Delta_{\text{iso}}}{\operatorname{SE}(\Delta_{\text{iso}})}$$
If the 95% confidence interval of $\Delta_{\text{iso}}$ spans zero, morphological dynamics cannot reject geometric isometry.

## Phenotypic variability meta-analysis and continuous dispersion modeling

Among-individual phenotypic variability was modeled using two complementary approaches:

1. **Meta-analytic $\ln\text{CVR}$**: For species with $\ge 5$ records in both the early (1995–2006) and late (2013–2018) periods, we computed the log coefficient-of-variation ratio ($\ln\text{CVR}$) with small-sample bias correction [@nakagawa2015metaanalysis]:
   $$\ln\text{CVR} = \ln\left(\frac{\operatorname{CV}_{\text{late}}}{\operatorname{CV}_{\text{early}}}\right) + \frac{1}{2(n_{\text{late}} - 1)} - \frac{1}{2(n_{\text{early}} - 1)}$$
   where sampling variance incorporates the observed correlation between log mean and log standard deviation within each period. Effect sizes were pooled in random-effects meta-analytic models using `metafor` [@viechtbauer2010metafor]. To distinguish genuine biological variance shifts from observer pooling artifacts, $\ln\text{CVR}$ was calculated both pooled across all observers ($k = 54$ for wing; $k = 25$ for bill width) and strictly within the same primary researcher ($k = 15$).
2. **Continuous heteroscedastic dispersion models**: In `glmmTMB`, we specified a dispersion submodel (`dispformula`) allowing residual standard deviation $\sigma_i$ to vary continuously over time and across observers:
   $$\log(\sigma_i) = \alpha_0 + \alpha_1 \text{scaled\_yr}_i + \alpha_2 \text{Sex}_i + u_{\text{researcher}} + u_{\text{municipality}}$$
   where $\alpha_1$ captures the decadal rate of change in residual standard deviation. In model tier S7, standardized local mean annual temperature (`scaled_tmean`) was added to the dispersion formula to test whether residual variance tracked environmental warming.

## Trophic interactions and regional prey context

Species' dietary reliance on invertebrates was extracted from the EltonTraits 1.0 database [@wilman2014eltontraits] (`Diet-Inv`, % of diet consisting of invertebrates). In continuous interaction models (Model A), `Diet-Inv` was standardized ($z$-score) and interacted with calendar year (`scaled_yr * diet_inv_std`). Wing-length trajectories were projected across the 10th (p10), 50th (p50), and 90th (p90) species-weighted percentiles of invertebrate reliance. In categorical models (Model B), diet was collapsed into Invertebrate-dominated ($\ge 50\%$ invertebrate diet) versus Other (frugivores, nectarivores, granivores, omnivores). Both models were fitted across 50 phylogenetic trees with observer and municipality random intercepts.

To evaluate regional prey trends, we modeled litter-ant richness from the ATLANTIC ANTS database [@silva2022atlanticants] across 855 standardized field campaigns (1994–2018). We fitted negative-binomial generalized linear mixed models in `glmmTMB` specifying observed ant species richness per campaign as a function of calendar year, with the log of sampling effort (number of pitfall traps $\times$ operational days) as an offset, and random intercepts for contributor, municipality, and sampling locality.

## Reproducible software environment

Analyses and manuscript generation were conducted using R version 4.4+ with packages `glmmTMB` (v1.1.14), `prepR4pcm` (v0.1.0), `metafor` (v4.6+), `ape` (v5.8+), `sf` (v1.0+), `terra` (v1.8+), and `tidyverse` (v2.0+), alongside Python (version 3.11+) with `matplotlib` (v3.10+) and `pandas` (v2.2+). All analysis scripts, curated datasets, and fitted model objects are openly archived and fully reproducible in the project repository (DOI: 10.5281/zenodo.atlantic_birds).

## References {.unnumbered}

::: {#refs}
:::

