---
title: "Supplementary Material: Morphological stability and subtle phenotypic shifts in Atlantic Forest birds across two decades of climate warming"
author:
  - name: Eduardo S.A. Santos
    affiliations:
      - Collaboration for Open Science and Synthesis in Ecology and Evolution, Department of Biological Sciences, University of Alberta, Edmonton, AB, T6G 2E9, Canada
  - name: Ayumi Mizuno
    affiliations:
      - Collaboration for Open Science and Synthesis in Ecology and Evolution, Department of Biological Sciences, University of Alberta, Edmonton, AB, T6G 2E9, Canada
  - name: Santiago Ortega
    affiliations:
      - Collaboration for Open Science and Synthesis in Ecology and Evolution, Department of Biological Sciences, University of Alberta, Edmonton, AB, T6G 2E9, Canada
  - name: Gustavo Burin
    affiliations:
      - Department of Biological and Environmental Sciences, University of Gothenburg, Sweden
bibliography: references.bib
number-sections: true
format:
  docx:
    reference-doc: custom-reference.docx
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
---

```{r load-estimates, include=FALSE}
desc        <- readRDS("../Analysis/output/descriptive_summary.rds")
cw_phylo    <- readRDS("../Analysis/output/controlled_wing_phylo_results.rds")
mass        <- readRDS("../Analysis/output/mass_results.rds")
multi_phylo <- readRDS("../Analysis/output/multitrait_phylo_results.rds")
biv_phylo   <- readRDS("../Analysis/output/bivariate_phylo_results.rds")
var_phylo   <- readRDS("../Analysis/output/variance_phylo_results.rds")
diet_phylo  <- readRDS("../Analysis/output/diet_interaction_phylo.rds")
clim        <- readRDS("../Analysis/output/climate_trends.rds")
es          <- readRDS("../Analysis/output/effect_scale.rds")
rr          <- readRDS("../Analysis/output/referee_reruns/referee_reruns.rds")

# format a count with a thousands separator for inline use
.n <- function(x) format(x, big.mark = ",", trim = TRUE)
# format a signed estimate / CI bound to 2 dp with a true minus sign
.s <- function(x) sub("-", "−", sprintf("%.2f", x))
# format a signed value to 1 dp with a true minus sign
.s1 <- function(x) sub("-", "−", sprintf("%.1f", x))
# format a signed estimate to 4 dp with a true minus sign
.s4 <- function(x) sub("-", "−", sprintf("%.4f", x))

# pull row from cw_phylo$before_after with optional term filtering
.cw <- function(mod, term = NULL) {
  r <- cw_phylo$before_after[cw_phylo$before_after$model == mod, ]
  if (!is.null(term)) r <- r[r$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", lo = "NA", hi = "NA", dec = "NA", dec_lo = "NA", dec_hi = "NA", n = "NA"))
  list(
    b      = .s(r$estimate[1]),
    lo     = .s(r$ci_lo[1]),
    hi     = .s(r$ci_hi[1]),
    dec    = .s1(r$mm_per_decade[1]),
    dec_lo = .s1(r$mm_per_decade_lo[1]),
    dec_hi = .s1(r$mm_per_decade_hi[1]),
    n      = .n(r$N[1])
  )
}

cw_m0           <- .cw("M0")
cw_m1a          <- .cw("M1a_src")
cw_m1           <- .cw("M1")
cw_m2           <- .cw("M2")
cw_m3           <- .cw("M3")
cw_m3_cc        <- .cw("M3_cc")
cw_m5src_within <- .cw("M5src", "yr_within_src")
cw_m5src_mean   <- .cw("M5src", "yr_src_mean")
cw_m6           <- .cw("M6")
cw_m7           <- .cw("M7")
cw_noanom       <- .cw("M3_noanom")
cw_long         <- .cw("M3_longrun")
cw_m5site_within <- .cw("M5", "yr_within_site")
cw_m5srccc_within <- .cw("M5src_cc", "yr_within_src")
b_atten_full    <- sprintf("%.0f", 100 * (1 - cw_phylo$before_after$estimate[cw_phylo$before_after$model == "M3"] /
                                          cw_phylo$before_after$estimate[cw_phylo$before_after$model == "M0"]))
# contributors and municipalities in the full sample and in the wing sample
prov <- local({
  e <- new.env(); load("../Analysis/data/derived/passer90.rda", envir = e)
  d <- e$passer90; w <- subset(d, !is.na(conc.wing.length))
  list(n_src = length(unique(d$Main_researcher)), n_site = length(unique(d$Municipality)),
       n_src_w = length(unique(w$Main_researcher)), n_site_w = length(unique(w$Municipality)))
})

# effect scale
if (!is.null(es)) {
  .esw <- es$wing$using_scaling_sd
  wing_pct_dec    <- .esw["est", "pct_per_decade"]
  wing_pct_rec    <- .esw["est", "pct_over_record"]
  wing_pct_rec_lo <- .esw["lo",  "pct_over_record"]
  wing_pct_rec_hi <- .esw["hi",  "pct_over_record"]
} else {
  wing_pct_dec <- -2.58; wing_pct_rec <- -5.9; wing_pct_rec_lo <- -9.0; wing_pct_rec_hi <- -2.8
}

# multi-trait comparisons from multi_phylo
.mt_phylo <- function(trait_name, mod, term = "scaled_yr") {
  df <- multi_phylo$comparison_lme4
  r <- df[df$trait == trait_name & df$model == mod & df$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", b2 = "NA", lo = "NA", hi = "NA", lo2 = "NA", hi2 = "NA", pct = "NA", pct_lo = "NA", pct_hi = "NA", n = "NA"))
  list(
    b      = .s4(r$phylo_estimate[1]),
    b2     = .s(r$phylo_estimate[1]),
    lo     = .s4(r$phylo_lower[1]),
    hi     = .s4(r$phylo_upper[1]),
    lo2    = .s(r$phylo_lower[1]),
    hi2    = .s(r$phylo_upper[1]),
    pct    = .s1(r$phylo_pct_per_decade[1]),
    pct_lo = .s1(r$phylo_pct_lower[1]),
    pct_hi = .s1(r$phylo_pct_upper[1]),
    n      = .n(r$phylo_n[1])
  )
}

multi_lme4 <- readRDS("../Analysis/output/multitrait_results.rds")
.mt_lme4 <- function(trait_name, mod, term = "scaled_yr") {
  df <- multi_lme4$table
  r <- df[df$trait == trait_name & df$model == mod & df$term == term, ]
  if (nrow(r) == 0) return(list(b = "NA", b2 = "NA", lo = "NA", hi = "NA", lo2 = "NA", hi2 = "NA", pct = "NA", pct_lo = "NA", pct_hi = "NA", n = "NA"))
  list(
    b      = .s4(r$estimate[1]),
    b2     = .s(r$estimate[1]),
    lo     = .s4(r$lower[1]),
    hi     = .s4(r$upper[1]),
    lo2    = .s(r$lower[1]),
    hi2    = .s(r$upper[1]),
    pct    = .s1(r$pct_per_decade[1]),
    pct_lo = .s1(r$pct_per_decade_lower[1]),
    pct_hi = .s1(r$pct_per_decade_upper[1]),
    n      = .n(r$n[1])
  )
}

mass_m0     <- .mt_phylo("ln_body_mass", "M0_baseline")
mass_m1a    <- .mt_lme4("ln_body_mass", "M1a_src")
mass_m1     <- .mt_phylo("ln_body_mass", "M1_src_site")
mass_m3     <- .mt_phylo("ln_body_mass", "M3_ring")
mass_within <- .mt_phylo("ln_body_mass", "MWsrc_mundlak_contributor", "yr_within_src")

bill_m0     <- .mt_phylo("Bill_width.mm.", "M0_baseline")
bill_m1a    <- .mt_lme4("Bill_width.mm.", "M1a_src")
bill_m1     <- .mt_phylo("Bill_width.mm.", "M1_src_site")
bill_m3     <- .mt_phylo("Bill_width.mm.", "M3_ring")
bill_within <- .mt_phylo("Bill_width.mm.", "MWsrc_mundlak_contributor", "yr_within_src")

# variance results
var_s0 <- var_phylo$comparison[var_phylo$comparison$tier == "S0" & var_phylo$comparison$term == "scaled_yr", ]
var_s3 <- var_phylo$comparison[var_phylo$comparison$tier == "S3" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_yr <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_yr", ]
var_s7_tm <- var_phylo$comparison[var_phylo$comparison$tier == "S7" & var_phylo$comparison$term == "scaled_tmean", ]

# diet interaction
diet_base <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_ctrl <- diet_phylo$interaction_table_all_specs[diet_phylo$interaction_table_all_specs$model == "A_src_site" & diet_phylo$interaction_table_all_specs$term == "interaction", ]
diet_p90  <- diet_phylo$quantile_slopes_all_specs[diet_phylo$quantile_slopes_all_specs$model == "A_src_site" & diet_phylo$quantile_slopes_all_specs$quantile == "p90", ]

# climate trends
clim_tm <- clim$pooled[clim$pooled$variable == "tmean" & clim$pooled$weighting == "unweighted + year RE", ]
```

# Supplementary Notes

## Supplementary Note S1: Macroecological Provenance and Contributor Turnover

Large-scale biodiversity trait compilations aggregate measurements across numerous independent research initiatives spanning wide geographical territories and multiple decades [@rodrigues_atlantic_2019]. In our analytical sample from the ATLANTIC BIRD TRAITS database, `r prov$n_src` contributors (research groups, identified by their principal investigator) carried out field mist-netting in `r prov$n_site` municipalities between 1995 and 2018 (`r prov$n_src_w` contributors and `r prov$n_site_w` municipalities in the wing sample; @fig-s-records).

A provenance audit revealed substantial compositional turnover among active research teams over time: only 6 of the `r prov$n_src_w` contributors in the wing sample sampled birds in both the early period (1995–2006; 12 active contributors) and the late period (2013–2018; 33 active contributors; @fig-s-records). Furthermore, spatial sampling expanded northward and into higher-elevation montane habitats during the latter half of the time series. In avian morphometrics, inter-observer measurement differences frequently reach 1–3 mm for wing length and 0.5–1 mm for bill dimensions [@weeks2020]. When contributor identity is left uncontrolled, compositional turnover can mimic temporal phenotypic change: the unadjusted wing-length decline attenuates by approximately `r b_atten_full`% once contributor, municipality and the other covariates are modelled (Table S1), and the apparent trends in bill width and trait variance do not persist.

To resolve genuine biological change from sampling turnover, all models beyond the baseline incorporate random intercepts for contributor and municipality (`(1 | Main_researcher)`, `(1 | Municipality)`). Within-between random effects models [Mundlak REWB; @mundlak1978pooling; @bell2019fixed] do not detect directional change within contributors, but estimate it imprecisely (within-contributor wing slope: `r cw_m5src_within$dec` mm/decade, 95% CI: `r cw_m5src_within$dec_lo` to `r cw_m5src_within$dec_hi`; complete cases: `r cw_m5srccc_within$dec` mm/decade, 95% CI: `r cw_m5srccc_within$dec_lo` to `r cw_m5srccc_within$dec_hi`; REML).

![**Supplementary Figure S1.** Temporal distribution of live-capture passerine wing records across contributors (1995–2018). (Top) Number of wing records per calendar year, stacked by the 12 largest research contributors (and an aggregated "Other (30)" category). Daggers (†) mark contributors that sampled birds in both the early (1995–2006) and late (2013–2018) periods. (Bottom) Number of active research contributors per calendar year.](images/fig-s-records-by-source.png){#fig-s-records}

![**Supplementary Figure S2.** Relative proportion of wing-length measurement columns populated in the ATLANTIC BIRD TRAITS database by year. Bars indicate the annual percentage of records reporting right wing length, left wing length, or side-unspecified wing length. Sample sizes ($n$) above bars indicate total annual captures with wing data.](images/fig-s-wingcol.png){#fig-s-wingcol}

## Supplementary Note S2: Regional Environmental and Climatic Trends (1995–2018)

We extracted monthly climate data from the CRU-TS 4.09 high-resolution gridded multivariate dataset ($0.5^\circ \times 0.5^\circ$ resolution; @harris2020crudata) at the 357 coordinate capture localities across the Atlantic Forest domain (@fig-s-climate-trends).

Across sampling localities, mean annual temperature increased significantly by $+0.246^\circ\text{C}$/decade (95% CI: $+0.076$ to $+0.415^\circ\text{C}$/decade, $t = 2.87, p = 0.005$), corresponding to an average regional warming of $+0.57^\circ\text{C}$ between 1995 and 2018 (@fig-s-climate-trends A). In contrast, total annual precipitation exhibited a non-significant trajectory ($-44.7$ mm/decade, 95% CI: $-123.1$ to $+33.7$ mm/decade, $p = 0.25$; @fig-s-climate-trends B), and the 12-month Standardized Precipitation-Evapotranspiration Index (SPEI-12) displayed no directional trend ($-0.256$ units/decade, 95% CI: $-0.617$ to $+0.105$, $p = 0.16$; @fig-s-climate-trends C). Regional environmental change across our capture sites was therefore characterized primarily by steady thermal warming without systematic multi-decadal changes in moisture availability.

![**Supplementary Figure S3.** Environmental trends across 357 avian capture localities in the Brazilian Atlantic Forest (1995–2018). **(A)** Mean annual temperature (°C), showing significant warming (+0.246°C/decade, $p = 0.005$). **(B)** Total annual precipitation (mm/year), showing non-significant variation ($-44.7$ mm/decade, $p = 0.25$). **(C)** 12-month Standardized Precipitation-Evapotranspiration Index (SPEI-12), indicating stability in drought regimes ($-0.256$/decade, $p = 0.16$). Solid lines and shaded bands represent pooled linear mixed-effects model trajectories with 95% confidence intervals.](images/fig-s-climate-trends.png){#fig-s-climate-trends}

## Supplementary Note S3: Species-Level Phenotypic Variability (lnCVR)

The main text reports among-individual phenotypic variability from continuous heteroscedastic dispersion models. As a complementary species-level summary, we computed log coefficient-of-variation ratios ($\ln\text{CVR}$) with small-sample bias correction [@nakagawa2015metaanalysis] for species with $\ge 5$ records in both the early (1995–2006) and late (2013–2018) periods, pooling effect sizes in random-effects meta-analytic models using `metafor` [@viechtbauer2010metafor]. Pooled across contributors, variability appeared to increase (wing length $\ln\text{CVR} = +0.177$, 95% CI: 0.031 to 0.323, $k = 54$; body mass $+0.193$, 95% CI: 0.021 to 0.366, $k = 57$; bill width $+0.068$, 95% CI: −0.146 to +0.282, $k = 36$). Restricted to within the same contributor, the apparent expansion disappeared ($\ln\text{CVR} = -0.249$, 95% CI: −0.515 to +0.017; $k = 15$), matching the conclusion drawn from the dispersion models. The within-contributor version requires a species to have been sampled by the same contributor in both periods, which few combinations satisfy, so it rests on a considerably smaller and compositionally narrower subset than the pooled comparison.

![**Supplementary Figure S4.** Interspecific patterns of among-individual phenotypic variability ($\ln\text{CVR}$) between early and late sampling periods. Forest plots display the species-level log coefficient-of-variation ratio ($\ln\text{CVR}$) comparing late (2013–2018) versus early (1995–2006) periods for (A) wing length ($k = 54$ species), (B) body mass ($k = 57$ species), and (C) bill width ($k = 36$ species) across Atlantic Forest passerines. Points and error bars represent species estimates and 95% confidence intervals; vertical lines and shaded bands represent random-effects meta-analytic means ($\ln\text{CVR} = +0.177$ for wing length; $+0.193$ for body mass; $+0.068$ for bill width).](images/fig-variability.png){#fig-variability}

## Supplementary Note S4: Temperature-Anomaly Models Across Specifications

The main text reports no association between morphology and the local annual temperature anomaly ($\Delta T_{\text{local}}$) under the full M3 adjustment set. Because an anomaly measured over a warming study period is itself correlated with calendar year, and because all birds captured at one locality in one year share an anomaly value, we fitted four specifications. All are produced by `Analysis/scripts/referee_reruns.R` and stored in `Analysis/output/referee_reruns/referee_reruns.rds`.

```{r tbl-s-thermal, echo=FALSE}
th <- rbind(
  transform(rr$E1[rr$E1$term %in% c("dT", "dT_detr", "scaled_yr"),
                  c("model", "response", "term", "estimate", "se", "z", "p")], trees = NA),
  transform(rr$E2[, c("model", "response", "term", "estimate", "se", "z", "p")], trees = NA),
  setNames(rr$E3[, c("model", "response", "term", "estimate", "se", "z", "p", "trees_used")],
           c("model", "response", "term", "estimate", "se", "z", "p", "trees")))
th$estimate <- signif(th$estimate, 3); th$se <- signif(th$se, 3)
th$z <- round(th$z, 2); th$p <- signif(th$p, 3)
knitr::kable(th, row.names = FALSE,
             col.names = c("Specification", "Response", "Term", "Estimate", "SE", "z", "p", "Trees"),
             caption = "**Supplementary Table S3.** Temperature-anomaly associations across specifications. `dT` is the local annual anomaly; `dT_detr` is the anomaly detrended within sites; `scaled_yr` is calendar year. E1a enters anomaly and year simultaneously; E2a adds a locality-year random intercept; E3 applies the full M3 adjustment set across 50 phylogenetic trees; E3b adds calendar year to E3. `Trees` is the number of the 50 trees whose fits converged and were pooled.")
```

In every specification containing both terms, calendar year carries the signal and the anomaly does not. We did not use a cluster-robust variance estimator: the random-effects structure is non-nested, which available implementations (`clubSandwich` CR2) do not support, so locality-year clustering is handled with a random intercept instead.

A further caveat is one of exposure window rather than statistics. Wing length is fixed at moult, so the mean temperature of the calendar year in which a bird happened to be captured is not a plausible causal exposure for its wing length. Any thermal mechanism would have to operate during feather growth, which these data cannot resolve.

# Supplementary Tables

```{r tbl-s-wing, echo=FALSE}
.row <- function(label, x) data.frame(Model = label, Beta = x$b, CI = paste0("[", x$lo, ", ", x$hi, "]"),
                                      Rate = x$dec, N = x$n)
tbl_cw_data <- rbind(
  .row("M0: Baseline (sex + latitude + phylogeny)", cw_m0),
  .row("M1a: + Contributor intercept only", cw_m1a),
  .row("M1: + Contributor and municipality intercepts", cw_m1),
  .row("M2: M1 + Wing-column proxy + Longitude + Altitude", cw_m2),
  .row("M3: Full controls (M2 + Season + Moult + Individual)", cw_m3),
  .row(paste0("M3_cc: Full controls on complete cases (N = ", cw_m3_cc$n, ")"), cw_m3_cc),
  .row("M3_noanom: Full controls without the anomalous sampling blocks", cw_noanom),
  .row("M5src: Within-contributor temporal slope (Mundlak REWB)", cw_m5src_within),
  .row("M5: Within-municipality temporal slope (Mundlak REWB)", cw_m5site_within),
  .row("M6: Unknown-sex replication (full M3 controls)", cw_m6),
  .row("M7: First captures only (no recaptures)", cw_m7))
knitr::kable(tbl_cw_data, row.names = FALSE,
             col.names = c("Model specification", "β (mm/SD-yr)", "95% CI", "Rate (mm/dec)", "N"),
             caption = "**Supplementary Table S1.** REML phylogenetic mixed models of wing length (glmmTMB, pooled across 50 trees; Wald 95% CI), evaluating temporal trajectories across sequential control tiers. M1a is shown to separate the contributor term from the municipality term, which M1 adds together.")
```

```{r tbl-s-mass, echo=FALSE}
.mrow <- function(label, x) data.frame(Specification = label, Beta = x$b, CI_95 = paste0("[", x$lo, ", ", x$hi, "]"),
                                       Rate_pct_dec = paste0(x$pct, "%"), N = x$n)
tbl_mass_data <- rbind(
  .mrow("Body mass M0 Baseline (sex + latitude + phylogeny)", mass_m0),
  .mrow("Body mass M1 (+ Contributor and municipality intercepts)", mass_m1),
  .mrow("Body mass M3 Full controls (M1 + Longitude + Altitude + Season + Individual)", mass_m3),
  .mrow("Body mass Mundlak within-contributor slope", mass_within))
knitr::kable(tbl_mass_data, row.names = FALSE,
             col.names = c("Model specification", "β (per SD-yr)", "95% CI", "Rate (%/dec)", "N"),
             caption = "**Supplementary Table S2.** Body mass temporal models across control tiers (REML check on the Bayesian estimates, glmmTMB pooled across 50 phylogenetic trees). Moult is not included because it is not recorded for enough of the body-mass records; M3 and the within-contributor model use the complete-case sample.")
```

# References {.unnumbered}

::: {#refs}
:::
